---
title: "Visualizing subnational democracy using Varieties of Democracy (V-Dem) survey
  data"
author: 'Patrick McQuestion^[Department of Political Science & Kroc Institute, University of Notre Dame. Email: <pmcquest@nd.edu>], Michael Coppedge^[Department of Political Science, University of Notre Dame. Email: <mcoppedg@nd.edu>], Matthew Sisk^[Department of Applied and Computational Mathematics and Statistics, University of Notre Dame. Email: <msisk1@nd.edu>]'
date: "First version: 26 April 2024. This version: `r format(Sys.Date(), '%d %B %Y')`."
output:
  word_document: default
  pdf_document:
    latex_engine: xelatex
    number_sections: yes
abstract: "In this paper, we introduce a procedure for measuring levels of democracy at the subnational-level using Varieties of Democracy (V-Dem) data. This effort in democracy studies is data-driven and descriptive, and is designed to examine the methodological potential for V-Dem to be applicable within countries. The underlying premise of our approach to measurement, which largely coincides with the broader V-Dem project, is that the use of subjective judgements from experts in democracy may provide a valid alternative to measures that rely solely on objective data, such as voter turnout or ruling party strength. Combining both subjective and observational data for criteria experts agree are relevant for understanding electoral and civil liberties dimensions of democracy, our measurement approach is both data-driven and amenable to different scales of analysis. This paper begins by situating our subnational democracy measure in comparison to existing measures, highlighting the importance of including both dimensions of elections and civil liberties for measurement. We then describe the methodological components and design of our measure by presenting the indicators we selected for measuring these dimensions in Colombia between 2000 and 2020. Once collected, the data are reduced and weighted by V-Dem proportional scores from four subnational surveys. The results are presented in graphic form and interpreted briefly. Then, we offer two approaches to validation of our meausure, finding that our technique, once scaled to the national-level, is statistically correlated with other existing measures of democracy. We conclude with a brief reflection on how this measure may be further refined and incorporated into future research of democracy at the subnational level around the world."
editor_options:
  markdown:
    wrap: 72
bibliography: References/BiblioRA24.bib
csl: G:/Shared drives/snvdem/snvdem-col/report/References/ajps.csl
link-citations: yes
linkcolor: black
subtitle: An application using the case of Colombia
header-includes:
- \usepackage{dcolumn}
- \usepackage{setspace}
- \doublespacing
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, 
                      message=FALSE, 
                      fig.path='Figs/',
                      fig.width=12, fig.height=8) 
setwd("G:/Shared drives/snvdem/snvdem-col/report")
```

# Introduction

There is a need for subnational data on many political phenomena:
violence, democracy, service provision, human rights, gender relations,
discrimination, and a range of other topics. The need is especially
great in large countries, which are likely to contain considerable
geographic variation on these dimensions within their borders. To the
extent that such variation exists, claims about the "typical" levels of
these dimensions for whole countries are less useful, either as
descriptions or as independent or dependent variables.

In this paper, we demonstrate a technique for combining several
underexplored Varieties of Democracy (V-Dem) variables with geocoded
demographic and political data to measure subnational democracy at the
municipal level. We introduce a pipeline that begins with
conceptualization of the individual variables included in V-Dem, and discuss how they can be operationalized in Colombia. We then use factor
analysis to reduce the variables to four underlying dimensions. These dimensions appear to align with arguments associated with modernization theory that suggest democracy levels are higher in populated urban centers where wealth is concentrated and there is demographic diversity across social groups.

This data reduction technique allows us to create weighted averages that can be interacted with V-Dem data to produce quantitative geospatial scores that can be utilized for further comparative subnational research. This exercise is designed to test the viability of mapping cross-sectional country-level data before expanding the dataset into
panel format. Internal data from V-Dem is subset to 2018 for Colombia, a
country with a large supply of geospatial demographic and social data. In the concluding section, we
describe the tasks that are needed to validate our estimates and to make
our measurements more robust.

# Approaches to measuring subnational democracy

While data based on direct observation are generally assumed to be
superior in comparative political analysis, there are some drawbacks to
this data collection strategy below the national-level. In the first
place, hard information about subnational units, such as states or
provinces, counties or municipalities, or even smaller districts, is
often scarce. Second, when the information exists, it can be difficult
or expensive to obtain, and may be collected in non-standardized ways.
Third, such "objective" data -- counts, rates, texts, etc. -- can be
invalid when they capture only a small part of the concept one needs to
measure. It may be feasible to count the number of newspapers and
broadcasters in a locality, but a valid measure of freedom of
information, for example, should also consider the size of their
audiences, their ability to carry out independent investigations, or
their editorial autonomy. Fourth, whether a directly observable measure
matters is often conditional on other factors. The lack of arrests of
journalists in North Korea is not evidence of press freedom; governing
parties sometimes win elections for decades even when elections are very
competitive; party platforms and speeches are often insincere and
inaccurate; and texts of constitutions and laws can be "parchment"
rules, dead letters that are not implemented in practice.

For these reasons, there are sometimes advantages to following an
alternative measurement strategy: relying on the subjective judgments of
multiple country experts. This approach can compensate for some of the
weaknesses of "objective" data [@bernhardMakingEmbeddedKnowledge2017].
Experts absorb information from multiple sources: reading, following
news, conversations, personal experiences, and deep familiarity with the
social and political context. This enables them to reconcile diverse and
often conflicting kinds of information and synthesize it into an
internally consistent account of how things work. Their understanding
extends to a sense of the past and the direction of trends. Because they
possess this depth of understanding, experts are less likely to be
deceived by insincere statements or *de jure* arrangements that are
ignored in practice, and they are less likely to consider one attribute
definitive when they know that other conditions render it irrelevant.

Of course, there is also an obvious disadvantage to subjective
knowledge: different experts tend to synthesize somewhat different
understandings. They arrive at different interpretations of the same
reality. This is why it is crucial to consult multiple experts, which
makes it possible to have greater confidence in the interpretations on
which experts agree, and to exercise prudent caution about matters on
which there is little agreement. Post hoc validation is also essential.
There are no guarantees that subjective ratings will yield useful
information, but it is worth trying as long as we also subject the
outcomes to multiple forms of validation.

## Existing measures of subnational democracy

Subnational democracy is an area where subjective expert ratings have
the potential to improve measurement. There are hundreds of democracy
measures, but almost all of them are at the national level of analysis
[@bernhardEconomicPerformanceInstitutional2001;
@bertelsmannstiftungBertelsmannTransformationIndex2005;
@boixCompleteDataSet2013; @coppedgeVDemDatasetV132023;
@economistintelligenceuniteiuDemocracyIndex20102010;
@freedomhouseMethodology2015; @buhlmannQualityDemocracyDemocracy2008;
@pemsteinDemocraticCompromiseLatent2010; @marshallPolityIVProject2014;
@przeworskiModernizationTheoriesFacts1997;
@skaaningLexicalIndexElectoral2015; @vanhanenNewDatasetMeasuring2000].
The few exceptions [@mickeyPathsOutDixie2015;
@mcmannDemocraticConsolidationTheory2018;
@gervasoniRentierTheorySubnational2010;
@giraudyVarietiesSubnationalUndemocratic2013] cover only one country
(although Giraudy does two).

In a recent methdological effort,
@perezsandovalMeasuringAssessingSubnational2023 designed an Index of
Subnational Electoral Democracy (ISED). The ISED tracks electoral
outcomes at the provincial level (the first jurisdictions below the
national-level) for 11 countries over 40 years, using a framework
attuned to the complexities around territorial conflation around
country-level systems such as judicial branches and political parties.
However, this index relies on electoral data rather than country
expertise. As alerted to earlier, electoral data may be collected in
non-standardized ways, especially in local jurisdictions. More
importantly, the number of parties may not always signify a competitive
political climate, nor does vote count always imply effective democratic
participation. Without considering the relevance of other factors, such
as the strength of civil liberties, electoral information alone will not
always be an efficient or valid measure of subnational democracy.

In addition to providing a novel subnational index, Perez Sandoval
provides a helpful summary of the concepts, operations, and data
collection methods used by other researchers of subnational democracy.
Among the eleven authors cited, only four -- @hernándezvaldez2000,
Ortega and Somuano in @meixueironájera2014, @gervasoni2018, and
@grumbach2023 -- attempt to measure beyond the electoral dimension of
democracy at the subnational level. These authors include measures of
civil liberties components -- such as human rights violations, local
radio and newspapers, or institutional constraints -- however they do so
only for individual countries (Mexico, Argentina, and the United States)
and at the provincial or state level. All four authors use advanced data
reduction techniques, such as Bayesian factor analysis or principal
component analysis, to arrive at subnational democracy scores, however
they stop short of providing a method for cross-national comparison.

Finally, pivoting to subjective expert data,
@mcmannExplainingSubnationalRegime2021 use a selection of V-Dem
variables to measure "unevenness" of elections and civil liberties in
most countries, but theirs is a national-level concept that tells us
nothing about where democracy is better or worse within a state.

## V-Dem subnational data

V-Dem data contains several variables that provide a partial path to
measurement at the subnational level. There are four questions on
elections and civil liberties: Subnational election area less free and
fair characteristics (`v2elsnlfc`), Subnational election area more free
and fair characteristics (`v2elsnmrfc`), Stronger civil liberties
characteristics (`v2clrgstch`), and Weaker civil liberties
characteristics (`v2clrgwkch`). Although elections and civil liberties
are far from capturing every aspect of democracy, they cover the two
most central criteria for polyarchy
[@dahlPolyarchyParticipationOpposition1971].

```{r vdem-q, echo=FALSE, message=FALSE, results='asis'}
library(flextable)
library(dplyr)
library(officer)
library(kableExtra)
library(knitr)

vdem_q <- data.frame(
  Question = c("Question: How would you describe the areas of the country in which elections are significantly less free and fair? [v2elsnlfc]",
               "Question: How would you describe the areas of the country in which elections are significantly more free and fair? [v2elsnmrfc]",
               "Question: How would you describe the areas of the country where government officials’ respect for civil liberties is significantly stronger? [v2clrgstch]", 
               "Question: How would you describe the areas of the country where government officials’ respect for civil liberties is significantly weaker? [v2clrgwkch]",
               "Clarification: Choose all that apply.",
               "Responses:",
               "0: Rural. (0=No, 1=Yes)",
               "1: Urban. (0=No, 1=Yes)",
               "2: Areas that are less economically developed. (0=No, 1=Yes)",
               "3: Areas that are more economically developed. (0=No, 1=Yes)", 
               "4: Inside the capital city. (0=No, 1=Yes)",
               "5: Outside the capital city. (0=No, 1=Yes)",
               "6: North. (0=No, 1=Yes)",
               "7: South. (0=No, 1=Yes)", 
               "8: West. (0=No, 1=Yes)",
               "9: East. (0=No, 1=Yes)",
               "10: Areas of civil unrest (including areas where insurgent groups are active). (0=No, 1=Yes)",
               "11: Areas where illicit activity is widespread. (0=No, 1=Yes)", 
               "12: Areas that are very sparsely populated. (0=No, 1=Yes)",
               "13: Areas that are remote (difficult to reach by available transportation, for example). (0=No, 1=Yes)", 
               "14: Areas where there are indigenous populations. (0=No, 1=Yes)", 
               "15: Areas where the national ruling party or group is strong. (0=No, 1=Yes)", 
               "16: Areas where the national ruling party or group is weak. (0=No, 1=Yes)",
               "17: Areas that were subject to a longer period of foreign rule. (0=No, 1=Yes)",
               "18: Areas that were subject to a shorter period of foreign rule. (0=No, 1=Yes)",
               "19: Areas that were recently subject to foreign rule. (0=No, 1=Yes)",
               "20: Areas that have not recently been subject to foreign rule. (0=No, 1=Yes)",
               "21: None of the above. (0=No, 1=Yes)"
               ))

# Print the table

vdem_q %>%
  kable(format = "latex", escape = FALSE, caption = "V-Dem questions and clarifications", col.names = NULL) %>%
  kable_styling(full_width = FALSE, latex_options = c("hold_position", "scale_down")) %>%
  column_spec(1:ncol(vdem_q), width = "45em") %>%
  row_spec(0:3, italic = TRUE)


```

Table 1 reproduces the four questions, along with the same 21 "choose
all that apply" criteria that apply to all four questions. In the
country-year V-Dem data sets, each criteria becomes a separate variable
that is no longer binary, but rather the proportion of experts who chose
the "yes" response to that option. Thus, in actuality we have 84 (21\*4)
variables to work with.

These unusual options have advantages and disadvantages. On the positive
side, these options ask experts to identify the criteria that
distinguish locations where democracy is better from locations where it
is worse. The options are described in a generic way so that they could
be relevant in any country that V-Dem rates, anywhere in the world, and
over more than two centuries of history. In this paper we deal with 16
of them, leaving aside the options referring to foreign rule (\# 17-20)
which is not very relevant for Colombia, and option #21 which refers to
a negative response ("None of the above"). We only use a cross-section
of 2018 data for exploratory purposes. Still, even these limitations and
response options are varied enough to cast a wide net that captures
diverse criteria for distinguishing where democracy is better or worse.

On the negative side, first, all of these options are conceptually
binary. In effect, they ask "Is this a useful discriminating criterion
or not?" rather than seeking information about degrees of relevance.
Second, the options do not name a specific concept or indicator that
should be used in empirical analysis. This decision is left up to each
country expert and V-Dem does not reveal--and in fact does not
*know*--on what information country experts based their assessments.
Third, the options do not specify a threshold that should be used to
decide whether a location does or does not satisfy the criterion.
Fourth, many of the criteria overlap and are therefore somewhat
interchangeable. This is especially likely for the binaries of
rural-urban, low-high income, sparsely-densely populated and remote
areas, which may also overlap with areas of illicit activity, civil
unrest, and indigenous population. Finally, the criteria are not mapped
onto well-defined locations such as states, counties, or cities that
would have known administrative boundaries.

Given these features, additional work is required to render the V-Dem
data useful for measuring democracy at the subnational level. The
following sections describe a pipeline for utilizing V-Dem data
geographically, from conceptualization, to operationalization, and
finally to measurement. This exercise is designed to test the viability
of mapping cross-sectional country-level data before expanding the
dataset into panel format.

The procedure described below describes this process in three broad
stages: selection of indicators (conceptualization), data reduction and
weighting (operationalization), and mapping subnational democracy scores
(measurement). For each stage, we describe how V-Dem data can be
integrated to show different aspects of subnational democracy. The final
measure of subnational democracy is an attempt to aggregate this data
more systematically.

# Conceptualization: Selection of Indicators

The first stage for measuring democracy begins at the level of
conceptualization. As shown above, V-Dem provides a set of variables
theorized to be relevant for subnational democracy. We choose to measure
subnational democracy grouping these variables into 10 unique criteria, but more
variables certainly exist. We use basic conceptualizations of these
criteria to identify one to three specific indicators -- from official
state data or others' scholarly research -- to measure them. In Colombia, we consider the indicators as shown in Table 2. All sources
in parentheses are official government sources, although alternative
indicators and sources certainly exist.

```{r, op-Col, echo=FALSE, message=FALSE, results='asis'}
library(flextable)
library(dplyr)
library(officer)
library(kableExtra)

#set_flextable_defaults(fontname = "Computer Modern", fontsize = 10)

vdem_op <- data.frame(
  ID = seq(0, 16, 1), stringsAsFactors = FALSE,
  Response = c("Rural", "Urban", 
                 "Less econ devt", "More econ devt", 
                 "Inside capital", "Outside capital", 
                 "North", "South", "West", "East", 
                 "Civil unrest", "Illicit activity", 
                 "Sparse pop.", "Remote", 
                 "Indigenous", 
                 "Ruling party strong", "Ruling party weak"),
  Indicator = c("Rural Index (DANE)", "Rural Index (DANE)",
        "Mun. Value Added / IPM (DNP)", "Mun. Value Added / IPM (DNP)",
        "Geographic distance from Bogotá (DANE)", "Geographic distance from Bogotá (DANE)", 
        "Macro-regions (DANE)", "Macro-regions (DANE)", "Macro-regions (DANE)", "Macro-regions (DANE)",
        "Forced displacement / Confinements (Min. Defense)", "Robberies / Homicides / Crop eradication (Min. Defense/UNODC)",
        "Population density (DANE)", "Distance to market (CEDE)",
        "Indigenous / Ethnic populations (DANE)", 
        "Pres. margin of victory (RNEC)", "Pres. margin of victory (RNEC)"))


# Print dataframe with merged cells
vdem_op %>%
  kable("latex", escape = FALSE, caption = "Conceptualization of V-Dem responses in Colombia") %>%
  kable_styling(full_width = FALSE) %>%
  collapse_rows(columns = 3, valign = "middle") %>%
  row_spec(0, bold = TRUE)

```

We prefer continuous data where possible because this produces
more specific calculations at the subnational-level. Most of the subnational V-Dem indicators can be measured as matters of
degree: income, population density, strength of the ruling party, etc. We can merge these continuous measures with V-Dem scores in at least two ways.
One option is to create a threshold and assign V-Dem
scores accordingly. For example, we could take the mid-point as a
threshold for rural and urban municipalities: if more than half of the
population is registered by the Census as rural inhabitants, then the municipality is
considered rural. This method provides a first cut, and may be more
intuitive to interpret. On the other hand, this approach involves the
setting of an arbitrary threshold that may produce misleading results,
especially given the cases grouped around the threshold. The preferable option, which we follow, is to interact continuous measures with V-Dem scores and
preserve the continuous nature of the data. We call this the Cumulative
Distribution Function (CDF) approach because it involves "ranking"
municipalities on a 0-1 scale for each variable.

```{r, V18_0vdem-plot, echo=FALSE, message=FALSE, results='hide', fig.cap="Example plots showing V-Dem scores applied to levels of rurality."}

# import shapefile
library(sf)
col <- st_read("G:/Shared drives/snvdem/snvdem-col/data/geospatial/2018pmq/BaseLayer/MGN_ANM_MPIOS.shp")
col <- col %>%
  select(1:8)

library(tidyverse)
#import dataset
V18 <- read_csv("G:/Shared drives/snvdem/snvdem-col/data/geospatial/2018pmq/x_Summary/V18.csv")

col <- merge(col, V18, by = "MPIO_CDPMP", all.x = TRUE)


# plot rurality maps
library(ggplot2)
library(gridExtra)
library(stringr)
library(grid)

d0 <- ggplot() +
  geom_sf(data = col, aes(fill = el_d0), color = "transparent") + 
  labs(title = "0. Rural.", caption = "Threshold approach") +
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem score",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "lightgreen",
    high = "darkgreen") + # Change label position
  theme(legend.title = element_text(size = 8)) +
  theme_void()
c0 <- ggplot() +
  geom_sf(data = col, aes(fill = el_c0), color = "transparent") + 
  labs(title = "0. Rural.", caption = "CDF approach") +
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem score",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "lightgreen",
    high = "darkgreen") + # Change label position
  theme(legend.title = element_text(size = 8)) +
  theme_void()

ap_0t1 <- grid.arrange(d0 + theme(plot.title = element_text(size = 16)), 
                               c0 + theme(plot.title = element_text(size = 16)), 
                       top = textGrob("How would you describe the areas of the country in which elections are significantly less free and fair?", gp = gpar(fontsize = 14)),
                               ncol = 2)
suppressMessages(ap_0t1)

```

Figure 1 maps the results of these two approaches using the question:
"How would you describe the areas of the country in which elections are
significantly less free and fair" (`v2elsnlfc`), and only looking at
option #0: Rural. On the left side is the threshold approach and on the
right side is the CDF approach. Clearly, the CDF approach provides much
more detail than the threshold approach, which can only provide a simple
visualization of areas categorized as rural. At the same time, the CDF
approach creates a new continuous measure that is less straightforward
to interpret. The CDF transforms the `0.875` proportion recorded in
V-Dem for this year in Colombia and, when interacted with the Rurality
Index, creates a variable whose municipal-level values range from
`0.438` to `0.736`. The benefit of using this variable, however, is that
it provides a granular and demographically informed estimate of the
association between rurality and less free and fair elections. Because
it is a normalized measure, this new estimate can be compared with the
normalized values for other responses, resulting in a municipal-level
estimate based on V-Dem experts' assessments of the background concept's
relevance.

## Interpreting coder confidence: the case of distance to the capital city

Subnational V-Dem data can be used to examine divergences in the degree
of confidence across elections or civil liberty questions. To
demonstrate this cross-question comparison subnationally, we can
consider variables #4-5 which ask how elections or civil liberties vary
"Inside / Outside the capital city." In the context of the V-Dem survey,
which is cross-national, respondents may reasonably interpret this
question to refer to the national capital (Bogotá). CEDE (University of Los
Andes) also provides a measure for this variable (linear distance in
kilometers) for which we can merge V-Dem data using the CDF approach.
Although the survey asks two questions, the first ("Inside the capital
city") is one extreme of the second ("Outside"), so we create one
continuous variable that reflects the relative proximity of
municipalities from Bogotá.

```{r V18_4-5_vdem-plot, echo=FALSE, message=FALSE, results='hide', fig.cap="Example plots showing V-Dem scores applied to distance from capital city."}
library(ggplot2)
library(gridExtra)
library(stringr)


c5el <- ggplot() +
  geom_sf(data = col, aes(fill = el_c4t5), color = "transparent", alpha = 0.5) + # check results
  labs(title = "Elections are less free and fair", caption = "5. Outside the capital city") + 
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem (ecdf)",
    title.position = "top",  
    label.position = "right"),
    low = "pink",
    high = "darkred",
    limits = c(0, 0.6)) + # Change limits to facilitate comparison
  theme(legend.title = element_text(size = 6)) +
  theme_void()
c5cw <- ggplot() +
  geom_sf(data = col, aes(fill = cw_c4t5), color = "transparent", alpha = 0.5) + # check results
  labs(title = "Civil liberties are weaker", caption = "5. Outside the capital city") +
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem (ecdf)",
    title.position = "top", 
    label.position = "right"),
    low = "pink",
    high = "darkred", 
    limits = c(0, 0.6)) + # Change limits to facilitate comparison
  theme(legend.title = element_text(size = 6)) +
  theme_void()

# Arrange the plots side by side
ap_4t5 <- grid.arrange(c5el + theme(plot.title = element_text(size = 14)), 
                                c5cw + theme(plot.title = element_text(size = 14)), 
                       top = textGrob("Comparison between survey question confidence levels", gp = gpar(fontsize = 14)),    
                                ncol = 2)
suppressMessages(ap_4t5)
```

Figure 2 compares two survey questions: where elections are less free and
fair, and where civil liberties are weaker. The proportions for inside
versus outside the capital city range from `0 - 0.375` for less free
elections, and from `0 - 0.600` for weaker civil liberties,
respectively. When we map the product of the continuous measure of
distance from the capital with both proportions, we see a clear
difference in the salience of this criterion for elections compared to
civil liberties. The contrasting gradients show the degree to which
respondents are more confident civil liberties are weaker outside the
capital city than they are that elections are less free and fair
according to this criterion.

## Layering dimensions of democracy: the case of regional variation

In addition to facilitating comparison for these conceptual questions,
one important advantage to using mapping software is the use of layers.
Layers can be used to produce composite images reflecting values from
more than one variable. For example, the proportional values for less
free and fair elections and weak civil liberties by region can produce a
rough composite image of lower levels of subnational democracy.

Although most criteria in the subnational V-Dem questions can be
conceptualized using continuous measures, criteria #6-9 ask respondents
to assess elections or civil liberties across four cardinal directions:
North, South, West, and East. Rather than using four cardinal regions in
Colombia, DANE classifies the country into six "macro-regions": a)
Caribe, b) Eje Cafetero (Antioquia), c) Centro Oriente, d) Llano, e)
Sur, and f) Pacifico [@dane-i].[^3] We group the six macro-regions in
the following manner: a)-b) as "North"; c)-d) as "East"; e) as "South";
and f) as "West".

[^3]: A public-access table of municipalities classified into these six
    regions can be found
    [here]((<https://www.datos.gov.co/Mapas-Nacionales/Departamentos-y-municipios-de-Colombia/xdk5-pm3f/about_data>).
    There are other divisions, such as the five regions used by CEDE
    studies (Andina, Caribe, Pacifica, Orinoquia, Amazonia), but for now
    we go with the prior classification.

```{r V18_6-9_vdem-plot, echo=FALSE, message=FALSE, results='hide', fig.cap="Example plots showing V-Dem scores applied to importance of region."}


elcw6t9 <- ggplot(col, aes(fill = m1_t6t9)) +
  geom_sf(color = NA) +
  labs(title = "Less F&F Elections and Weak Civil Liberties",
       caption = "6-9. North-South-West-East") +
  scale_fill_gradient(name = "V-Dem (mean)",
                      low = "pink",
                      high = "darkred",
                      guide = guide_colorbar(title.position = "top",
                                             label.position = "right")) +
  scale_alpha_continuous(range = c(0.2, 1), guide = FALSE) + 
  theme_void()

emcs6t9 <- ggplot(col, aes(fill = m2_t6t9)) +
  geom_sf(color = NA) +
  labs(title = "More F&F Elections and Strong Civil Liberties",
       caption = "6-9. North-South-West-East") +
  scale_fill_gradient(name = "V-Dem (mean)",
                      low = "lightgreen",
                      high = "darkgreen", 
                      guide = guide_colorbar(title.position = "top",
                                             label.position = "right")) +
  scale_alpha_continuous(range = c(0.2, 1), guide = FALSE) + 
  theme_void()


# Arrange the plots side by side
ap_6t9 <- grid.arrange(elcw6t9 + theme(plot.title = element_text(size = 14)), 
                                emcs6t9 + theme(plot.title = element_text(size = 14)), 
                       top = textGrob("Combined (layered) elections and civil liberties dimensions", gp = gpar(fontsize = 14)),
                       ncol = 2)
suppressMessages(ap_6t9)
#ggsave("G:/Shared drives/snvdem/snvdem-col/data/geospatial/2018pmq/6-9_Cardinal/6t9_comparison.jpg", arranged_plots4, device = "jpg")
```

Figure 3 depicts two sets of layers across the four regions as
classified above in order to compare V-Dem's democratic dimensions.
There are only two variables involved in each map, so a simple mean is
used to calculate the final scores. Under this method, the South is seen
as less democratic than the North, although both proportional scales
suggest that respondents' confidence in these regional variations are
low: the range for the mean of respondent agreement around less free and
fair elections and weaker civil liberties is `0.175-0.375` while the
range for the mean of respondent agreement around more free and fair
elections and stronger civil liberties is `0.000-0.278`.

These three examples show how data from external sources can be
interacted with 2018 V-Dem values in a straightforward manner. In these
examples, we also included visualizations to demonstrate how each
criteria appears depending on how V-Dem data is applied. However, this
method provides only a piecemeal method of visualizing subnational
democracy, rather than a composite image that integrates all of the
factors at once. Once data are collected for each of the criteria used
by V-Dem, we must find a way to combine these factors in a systematic
and informed way. For these reasons, in the following section we turn to
the operationalization stage, where external data are reduced in order
to identify key dimensions by which to weight the relative importance of
each factor.

# Operationalization: Data Reduction

Each variable can be interacted with V-Dem data individually, but a
measure of democracy that can compile these different features is more
informative. In this section we use a data reduction measurement
approach that allows for the use of multiple indicators without
hindering the final aggregate measure of democracy at the
municipal-level. In particular, we turn to factor analysis for informing
our weighting strategy.

Factor analysis uses the full data set of variables to group co-variates
together into a reduced number of latent dimensions. This allows
indicators to be treated as substitutable and, once grouped into latent
dimensions, these factors can be aggregated with an additive or weighted
average formula [@goertzSocialScienceConcepts2006]. The underlying logic
is that strength on one dimension tends to compensate for weakness in
another. This kind of relationship can be operationalized as a sum, an
average, or a weighted average. Ideally we prefer weighted averages, as
averages preserve the `0-1` range of values and weights employ a more
relaxed assumption than "unweighted" averages, which actually enforce a
strong assumption of equal weighting.

Factor analysis yields factor or component weights based on the
commonalities among our criteria. This technique also allows us to
evaluate whether it makes sense to treat all these criteria as
unidimensional. If it does not, then we would do well to generate
multiple indices that measure somewhat independent dimensions of
democracy.

The next section includes a description of the latent factors revealed
by the factor analysis of the social and demographic data for Colombia
in 2018. The following section describes how V-Dem data is integrated
with the external municipal-level data using a weighting process
informed by the factor analysis results.

## Factor analysis: identifying latent factors in the data

Figure 4 depicts the results of the data reduction technique, revealing
4 latent factors corresponding to demographic and political attributes
at the subnational level in Colombia for 2018. These dimensions can be
characterized as such, with corresponding V-Dem criteria numbers listed in order of relative importance for each dimension in parentheses:

-   Factor 1 = Indigenous areas with less support for ruling party and distant
    from Bogotá (#14, #15-16, #4-5)

-   Factor 2 = Densely populated, wealthy, ethnically diverse urban areas,
    with petty crime (#12, #2-3, #14, #0-1, #11)

-   Factor 3 = Poor areas distant from Bogotá with less petty crime (#2-3,
    #4-5, #11)

-   Factor 4 = Areas with victims of violence and civil unrest (#10, #11)

![Results from the Factor Analysis, mapped at the
municipal-level](analysis/2FAfigs/FA-4_map.png)

The total variance explained by these four factors is `76.2%`. The
results suggest that the first two factors explain nearly `50%` of the
data variation. These factors--indigenous populations less supportive of
the ruling party (at the time, right-wing President Ivan Duque) distant
from Bogotá, and wealthy economically developed areas--suggest a split
along rural and urban, poorer and richer, and perhaps areas dominated by
indigenous and non-indigenous populations.

In terms of democratic theory, these splits may align with modernization
arguments that urban, wealthy, and ethnically heterogeneous areas are
associated with higher levels of democracy.

## Weighting V-Dem responses using factor scores

In order produce a summary measure of a concept closer to "democracy,"
we turn to the data in V-Dem. The factor analysis results can be used to
weight the V-Dem scores at the municipal-level, but the V-Dem scores
must be reduced as well.

As mentioned before, V-Dem subnational data is organized into two
surveys: elections and civil liberties. These capture two of the most
central criteria for polyarchy, according to
@dahlPolyarchyParticipationOpposition1971 . Within each survey, experts
are asked to select from the 21 variables only those that are relevant for
explaining either less free and fair elections or more free and fair
elections, and the same for civil liberties: the variables that explain
stronger or weaker civil liberties.

The relative importance of the subnational survey variables for
understanding elections or civil liberties, for Colombia in 2018, are
summarized in Table 3. The first column includes the names for
individual variables, columns 2 through 5 include the scores recorded in
V-Dem, and columns 6 and 7 show the absolute values for each variable
subtracting the less/weaker from more/stronger elections and civil
liberties, respectively. As the table shows, some survey variables
register high degrees of variation across elections and civil liberties (e.g., Rural or Urban,
Economic development, Capital city, Civil unrest, Illicit activity)
while others register relatively lower variation (e.g.,
North-South-West-East, Ruling party strong or weak). Note that extreme
ends of the range (scores closer to `0` or closer to `1`) indicate more
agreement around the salience of a variable, and conversely, scores
towards the middle (closer to `0.5`) indicate lower amounts of consensus
among coders. The absolute values reveal the estimated relative importance that each
variable has for either dimension of democracy. Variables such as civil unrest, illicit activity, distance from the capital, rural and urban, or economic development appear to be more salient than the regional variables, population density, or support for ruling party. The higher average absolute score for civil liberties variables suggest that coders are more confident in their judgments around this dimension of democracy.


```{r, Col18-table, echo=FALSE, message=FALSE, results='asis'}
library(tidyverse)
#import dataset
col18_weights <- read_csv("G:/Shared drives/snvdem/snvdem-col/report/analysis/3_weighttable.csv")

# Load necessary libraries
library(dplyr)
library(tidyr)
library(knitr)
library(purrr)
library(kableExtra)
library(insight)


new_row_names <- c(
  "Rural", "Urban", "Less econ devt", "More econ devt", "Inside capital", 
  "Outside capital", "North", "South", "West", "East", "Civil unrest", 
  "Illicit activity", "Sparse pop.", "Remote", "Indigenous", 
  "Ruling party strong", "Ruling party weak"
)

# Rename columns and modify row_id variable
summary_table <- col18_weights %>%
  mutate(row_id = new_row_names[as.numeric(row_id) + 1]) %>%
  rename(
    `Variable` = row_id,
    `EL Less free` = v2elsnlfc,
    `EL More free` = v2elsnmrfc,
    `CL Stronger` = v2clrgstch,
    `CL Weaker` = v2clrgwkch,
    `Abs(EL)` = emel,
    `Abs(CL)` = cscw
  ) 

averages <- summary_table %>%
  summarise(
    `Variable` = "Average",
    `EL Less free` = round(mean(`EL Less free`), 3),
    `EL More free` = round(mean(`EL More free`), 3),
    `CL Stronger` = round(mean(`CL Stronger`), 3),
    `CL Weaker` = round(mean(`CL Weaker`), 3),
    `Abs(EL)` = round(mean(`Abs(EL)`), 3),
    `Abs(CL)` = round(mean(`Abs(CL)`), 3)
  )

summary_table2 <- bind_rows(summary_table, averages)

summary_table2 %>%
  kable(format = "html", booktabs = TRUE, caption = "V-Dem response scores for Colombia in 2018") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(nrow(summary_table2), bold = TRUE, color = "white", background = "darkgray")  

k <- summary_table2 %>%
  kable(format = "html", booktabs = TRUE, caption = "V-Dem response scores for Colombia in 2018") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(nrow(summary_table2), bold = TRUE, color = "white", background = "darkgray")  

writeLines(k, "G:/Shared drives/snvdem/snvdem-col/report/analysis/table3.html")

```

Before interacting the factor analysis results with these scores, we also reduce the number of individual variables for a few cases: rural and urban (#0-1), more or less economic development
(#2-3), inside or outside the capital city (#4-5), and more or less
support for the ruling party (#15-16). For these criteria, we take the average of the
absolute values for the pair of individual factors. For example, we
should not be `0.889` confident that the capital is more democratic if we
are only `0.375` confident that it is less democratic outside the
capital. We take the average of these scores for the criteria of
distance from capital for less free and fair elections. When we use
multiple indicators for a single-option criterion, as with indigenous
populations (#14), we have to apply the same weight to all of them.

In order to weight these scores using the factor analysis results, we
first take the absolute value of the proportions across both less/weaker
and more/stronger categories. The less/weaker proportions can be
subtracted from the more/stronger proportions, then converted to
absolute values to represent the weight of each criteria. Whether the
criteria will predict more or less democracy will depend on the
demographic data from the country.

# Measurement: Mapping subnational democracy

The final step in our pipeline is to produce a measurement of
subnational democracy based on the conceptualization and
operationalization results.

In order to use the weighted V-Dem data, we affirm that the factor
scores we calculated earlier have positive signs for more democracy and
negative scores for less democracy, based on the theory. In this case,
we saw that modernization theory predicts that Factor 2 is associated
with stronger democracy, while Factors 1, 3, and 4 are associated with
weaker democracy. After correcting for positive and negative signs, 
we convert the factor scores to a 0-1 scale by
normalizing the data, then we multiply the transformed factor scores by
these above weights. Finally, we add the scores for each municipality,
and divide by the sum of the weights to get a weighted average.

The final result is shown in Figure 4. The higher values in blue indicate areas where subnational democracy levels are estimated to be higher, according to this methodology. The highest levels of democracy appear to be in the Andean region, particularly in Cundinamarca (where Bogotá is located) and in Antioquia (where Medellin is located). Outside of the Andean region, particularly along the Pacific Coast (e.g., Choco) and the Amazon region, democracy levels are relatively lower. 

```{r, FA-summary, echo=FALSE, message=FALSE, results='hide', fig.cap="Measurement of subnational democracy in Colombia (2018) using V-Dem data."}


#import dataset
FAcol18 <- read_csv("G:/Shared drives/snvdem/snvdem-col/report/analysis/3_FAcol18.csv")

# ----- Merge weighted scores w/ geometries to make maps -----
## Maps
library(sf)
library(tidyverse)
library(maps)
library(mapdata)
library(mapproj)

col <- st_read("G:/Shared drives/snvdem/snvdem-col/data/geospatial/2018pmq/BaseLayer/MGN_ANM_MPIOS.shp")
col <- col %>%
  select(1:8)

colmap_data <- merge(x = FAcol18, y = col, by = "MPIO_CDPMP", all.x = TRUE)
colmap_data <- st_as_sf(colmap_data)


ggplot() +
  geom_sf(data = colmap_data, color="transparent", linewidth = 0.01, aes(fill = MRm_dem)) +
  theme_void() + 
  theme(panel.background = element_rect(color = "transparent", fill = "white"),
        plot.caption = element_text(size = 12)) +
  scale_fill_viridis_c(option = "plasma", direction = -1) + 
  labs(title = "Measurement: subnational democracy (Colombia 2018)", 
       fill = "SN Democracy score")


```

# Conclusion

As noted previously, validation is a crucial step in producing data
based on expert ratings. We have not yet been able to do a systematic
validation of the data we have so far, but we have some observations
about face validity based on the maps we have produced. In general, our
data indicate higher levels of democracy (both in free and fair
elections and in protections for civil liberties) in the major urban
areas and lower levels in rural or remote areas such as the Amazon. 

Systematic validation will require comparisons with independent
assessments of subnational democracy in Colombia. An initial source for
such information will be responses to open-ended versions of V-Dem's
four questions provided by a small number of university and master's
students at a Colombian university in February 2024. They were asked to
identify the most salient locations (defined any way they chose) in
Colombia and years where they would expect valid indicators to show
better or worse situations. We can check our data against their
expectations. At some point we could solicit the same kinds of responses
from a more senior panel of Colombian experts. Ultimately, the best kind
of validation would be administering an online survey asking V-Dem's
country experts on Colombia to code these same questions for Colombian
departments or municipalities for selected years. We do not know whether
such an effort is feasible, but it is probably impossible to know until
we try.

Eventually we hope to extend this exercise to other countries where
others have generated measures of subnational democracy, such as
Argentina, Mexico, India, the United States, and Russia. We do not
downplay the difficulty of such an effort. It requires access to very
specific geo-coded data on multiple demographic and political
indicators, which are probably not available for many countries.
Furthermore, producing cross-nationally comparable data involves an
additional layer of complication. The data we have produced for Colombia
is relative: the degree to which local situations are better or worse
than the national norm. Converting such relative data to absolute levels
will require both anchoring the ratings to a national score and
establishing the range of variation within which the relative scores can
vary. V-Dem's Clean Elections Index could serve as the anchor for the
subnational elections variables and its Civil Liberties Index could
serve as the anchor for the subnational civil liberties variables. The
same anchoring indices could be used for other countries. However, we do
not yet have a proposed solution for setting the range of variation
around these anchors. This would require careful conceptual and
methodological effort.

Colombia is among the best cases for this initial proof-of-concept
research due to the well-developed geographic data that DANE provides
(and for which the Lucy Institute, where co-author Matthew Sisk works,
has direct access). Replication in other countries would make it
possible to reach more general conclusions about the validity of the
technique and to decide whether it is promising enough to extend to many
more countries. If results are encouraging, there may be a potential for
extending of this technique to measure other political concepts
subnationally.

# References
