---
title: "Measuring Democracy in Subnational Units Worldwide: A New Method"
subtitle: "Mapping snvdem in Colombia (2000--2023)"
abstract: |
  Scholars have documented that democracy does not exist evenly within 
  countries throughout the world and explained why. Increasingly, studies 
  have demonstrated that this subnational variation in democracy affects 
  other aspects of politics and society, such as state capacity, 
  democratization, violence, and environmental degradation. Further progress 
  in studying subnational democracy, however, is limited by the absence of 
  cross-national, time-series data that incorporate a thick conceptualization 
  and operationalization of democracy and cover all subnational units of many 
  countries. Current datasets of subnational unit democracy are limited to a 
  small number of countries. Moreover, most studies rely on a thin 
  conceptualization of democracy, measured with indicators such as electoral 
  margins of victory, effective numbers of candidates, or voter turnout. We 
  offer a method to generate cross-national time-series data of democracy in 
  subnational units that relies on a thick conceptualization and 
  operationalization of democracy. We combine expert data from elections and 
  civil liberties surveys in the Varieties of Democracy (V-Dem) dataset and 
  geocoded observational data. We use these data to generate annual average 
  levels of how free and fair subnational elections are, how well civil 
  liberties are protected, and overall democracy in each subnational unit of 
  a country, thus capturing not only the electoral, but also the broader 
  liberal, conceptualization of democracy. We pilot the method by generating 
  municipal democracy data for Colombia for years 2000--2023. We conduct 
  three validation tests of these data and our method. We also demonstrate 
  the utility of the data by investigating the impact of Colombia's 2016 
  Comprehensive Peace Agreement with the FARC-EP on democracy in its rural 
  municipalities. Our method can be applied to other countries, so it offers 
  the promise of generating cross-national time-series data of democracy in 
  subnational units in many countries of the world. This will enable us to 
  test the generalizability of conclusions from earlier subnational democracy 
  research and to expand the questions we can answer.
author: "Patrick McQuestion"
date: "April 2, 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: false
    fig_caption: true
    fig_width: 10
    fig_height: 7
    keep_tex: false
    latex_engine: xelatex
geometry: "margin=1in"
fontsize: 11pt
linestretch: 1.15
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
# --- KNIT OPTIONS ---
knitr::opts_chunk$set(
  echo        = FALSE,   # hide code in output
  warning     = FALSE,   # suppress warnings
  message     = FALSE,   # suppress package messages
  fig.align   = "center",
  fig.pos     = "H",     # hold figures in place
  out.width   = "100%",
  dpi         = 300,
  cache       = FALSE
)

# Force markdown pipe tables — prevents kableExtra from injecting
# tabu/threeparttable into the LaTeX preamble
options(knitr.table.format = "pipe")

# --- ROOT DIRECTORY ---
knitr::opts_knit$set(
  root.dir = "G:/Shared drives/snvdem/snvdem-col/data/panel/"
)

# --- LIBRARIES ---
library(tidyverse)
library(sf)
library(patchwork)
library(ggrepel)
library(viridis)
library(rnaturalearth)
library(rnaturalearthdata)
library(knitr)
## kableExtra removed: causes tabu.sty dependency not in base TinyTeX

# --- DATA ---
snvdem <- readRDS(
  "G:/Shared drives/snvdem/snvdem-col/data/panel/09_analysis_scripts/Exploratory/01_Expand/snvdem2.rds"
)

# --- HELPER FUNCTION: pad MPIO codes ---
clean_mpio <- function(x) {
  str_pad(as.character(as.numeric(x)), width = 5, side = "left", pad = "0")
}
```

# Pilot Application: Colombia

## Visualizing Subnational Democracy

From the data, we learn that levels of democracy have varied tremendously
across Colombian municipalities between 2000 and 2023. Interestingly, over
time, democracy appears to have increased the most in those municipalities
that were already more democratic. The range extends consistently from
relatively high levels of democracy in the area surrounding the capital
Bogotá, as indicated by the yellow shading, to relatively low levels
adjacent to the more southern part of the border with Venezuela, as
indicated by the purple. However, in the area surrounding Bogotá the
absolute level of democracy became significantly higher by 2023. This is
evident in country maps from five different years (Figure 4).

```{r facets, echo=FALSE, fig.width=14, fig.height=10, fig.cap="Spatial Evolution of Subnational Democracy in Colombia (2000--2023). Normalized sndem Index (0 = Low, 1 = High). Source: SNDEM 2024."}

# --- 1. DATA PROCESSING ---
master_df <- snvdem[, 1:74]

master_clean <- master_df %>%
  filter(!is.na(snelect) & !is.na(sncivlib) & !is.na(sndem)) %>%
  mutate(
    snelect_norm  = (snelect  - min(snelect))  / (max(snelect)  - min(snelect)),
    sncivlib_norm = (sncivlib - min(sncivlib)) / (max(sncivlib) - min(sncivlib)),
    sndem_norm    = (sndem    - min(sndem))    / (max(sndem)    - min(sndem))
  )

master_clean$MPIO_CDPMP <- as.character(master_clean$MPIO_CDPMP)

# --- 2. LOAD GEOSPATIAL DATA ---
muni_geo <- st_read(
  "G:/Shared drives/snvdem/snvdem-col/data/geospatial/2018pmq/BaseLayer/MGN_ANM_MPIOS.shp",
  quiet = TRUE
) %>%
  mutate(MPIO_CDPMP = clean_mpio(MPIO_CDPMP))

muni_geo$MPIO_CDPMP <- as.character(muni_geo$MPIO_CDPMP)

# --- 3. JOIN SCORES TO SPATIAL DATA ---
years_to_map <- c(2000, 2006, 2012, 2018, 2023)
map_data <- muni_geo %>%
  left_join(
    master_clean %>% filter(year %in% years_to_map),
    by = "MPIO_CDPMP"
  )

# --- 4. CONTEXTUAL DATA ---
world      <- ne_countries(scale = "medium", returnclass = "sf")
neighbors  <- world %>%
  filter(continent == "South America" | name %in% c("Panama"))
col_departments <- ne_states(country = "Colombia", returnclass = "sf")

# --- 5. STYLIZED CONTEXT MAP ---
stylized_context_map <- ggplot() +
  geom_rect(
    aes(xmin = -82, xmax = -66, ymin = -6, ymax = 14),
    fill = "#e0f3ff"
  ) +
  geom_sf(data = neighbors,       fill = "#d5e8d4", color = "grey70") +
  geom_sf(data = col_departments, fill = "white",   color = "grey40",
          linewidth = 0.2) +
  geom_label_repel(
    data = data.frame(
      name = c("Bogotá", "Medellín", "Cali",
               "Barranquilla", "Cartagena", "Bucaramanga"),
      x = c(-74.07, -75.56, -76.52, -74.78, -75.48, -73.12),
      y = c(  4.71,   6.25,   3.45,  10.96,  10.40,   7.12)
    ),
    aes(x = x, y = y, label = name),
    size = 2.8, fontface = "bold",
    box.padding = 0.6, point.padding = 0.3,
    segment.color = "grey30", min.segment.length = 0,
    fill = alpha("white", 0.8)
  ) +
  annotate("text", x = -78,   y = 12.5, label = "Caribbean Sea",
           size = 3, fontface = "italic", color = "#4a90e2") +
  annotate("text", x = -80.5, y = 4,    label = "Pacific Ocean",
           size = 3, fontface = "italic", angle = 90, color = "#4a90e2") +
  annotate("text", x = -70,   y = 10,   label = "VENEZUELA",
           size = 3.2, color = "grey40", fontface = "bold") +
  annotate("text", x = -68.5, y = -3,   label = "BRAZIL",
           size = 3.2, color = "grey40", fontface = "bold") +
  annotate("text", x = -78.5, y = -1.5, label = "ECUADOR",
           size = 3.2, color = "grey40", fontface = "bold") +
  annotate("text", x = -74.5, y = -5,   label = "PERU",
           size = 3.2, color = "grey40", fontface = "bold") +
  annotate("text", x = -79.5, y = 8.8,  label = "PANAMA",
           size = 2.8, color = "grey40", fontface = "bold") +
  coord_sf(xlim = c(-82, -66), ylim = c(-6, 14), expand = FALSE) +
  theme_void() +
  theme(panel.border = element_rect(colour = "grey20", fill = NA,
                                    linewidth = 0.5))

# --- 6. FACETED TIME-SERIES MAPS ---
facet_maps_list <- lapply(years_to_map, function(y) {
  ggplot(data = map_data %>% filter(year == y)) +
    geom_rect(
      aes(xmin = -79.5, xmax = -66.5, ymin = -4.5, ymax = 12.5),
      fill = "#f0f8ff"
    ) +
    geom_sf(data = neighbors, fill = "#f2f9f2",
            color = "grey90", linewidth = 0.05) +
    geom_sf(aes(fill = sndem_norm), color = NA) +
    scale_fill_viridis_c(
      option = "magma", name = "Democracy\nScore",
      na.value = "grey80", limits = c(0, 1),
      guide = guide_colorbar(barheight = unit(3, "cm"), ticks = FALSE)
    ) +
    coord_sf(xlim = c(-79.5, -66.5), ylim = c(-4.5, 12.5),
             expand = FALSE) +
    labs(title = as.character(y)) +
    theme_void() +
    theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5,
                                margin = margin(t = 5, b = 5)),
      legend.position = "none"
    )
})

# --- 7. ASSEMBLE ---
layout_stable <- c(
  area(1, 1), area(1, 2), area(1, 3),
  area(2, 1), area(2, 2), area(2, 3)
)

final_visual <- stylized_context_map +
  facet_maps_list[[1]] + facet_maps_list[[2]] + facet_maps_list[[3]] +
  facet_maps_list[[4]] + facet_maps_list[[5]] +
  plot_layout(design = layout_stable, guides = "collect") +
  plot_annotation(
    title    = "Figure 4: Spatial Evolution of Subnational Democracy in Colombia (2000--2023)",
    subtitle = "Normalized sndem Index (0 = Low, 1 = High)",
    caption  = "Source: SNDEM 2024."
  ) &
  theme(
    plot.title    = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey30", size = 12),
    legend.position = "right"
  )

# Save to disk (wrapped in tryCatch so knitting doesn't fail if path unavailable)
tryCatch(
  ggsave(
    "G:/Shared drives/snvdem/snvdem-col/data/panel/08_final_snvdem_data/MC/imgs/SNDEM_map_legend4.png",
    final_visual, width = 14, height = 10, dpi = 300, bg = "white"
  ),
  error = function(e) message("ggsave skipped: ", e$message)
)

final_visual
```

## Zooming In: What the Colombia Data Tell Us

Zooming in, we can examine what the resulting estimates reveal about
democratic variation across Colombian municipalities from 2000 to 2023.
This section proceeds in three steps: an overview of the data as a whole,
the identification and description of illustrative cases that capture the
range of democratic trajectories in the data, and a brief discussion of two
theoretical questions these data are positioned to address. The goal of
this section is not causal inference but demonstration --- showing what kind
of descriptive and comparative leverage a thick, cross-sectional time-series
measure of subnational democracy makes possible.

```{r data-prep, include=FALSE}
# ---------------------------------------------------------------
# DATA PREPARATION (exploratory; output suppressed)
# ---------------------------------------------------------------

# Normalize scores
master_clean2 <- snvdem %>%
  filter(!is.na(sndem)) %>%
  mutate(
    sndem_norm    = (sndem    - min(sndem,    na.rm = TRUE)) /
                   (max(sndem,    na.rm = TRUE) - min(sndem,    na.rm = TRUE)),
    snelect_norm  = (snelect  - min(snelect,  na.rm = TRUE)) /
                   (max(snelect,  na.rm = TRUE) - min(snelect,  na.rm = TRUE)),
    sncivlib_norm = (sncivlib - min(sncivlib, na.rm = TRUE)) /
                   (max(sncivlib, na.rm = TRUE) - min(sncivlib, na.rm = TRUE))
  )

# Control variables for case matching
ctrl_vars <- c("Acueducto", "MortInfantil1", "NetCover", "AvgS11", "PobTot_12")

# Municipality-level averages
muni_avg <- master_clean2 %>%
  group_by(MPIO_CDPMP, municipio, depto, DPTO_CCDGO) %>%
  summarise(
    across(all_of(ctrl_vars),
           ~ mean(.x, na.rm = TRUE), .names = "{.col}_avg"),
    sndem_avg    = mean(sndem_norm,    na.rm = TRUE),
    snelect_avg  = mean(snelect_norm,  na.rm = TRUE),
    sncivlib_avg = mean(sncivlib_norm, na.rm = TRUE),
    sndem_2000   = mean(sndem_norm[year == 2000], na.rm = TRUE),
    sndem_2023   = mean(sndem_norm[year == 2023], na.rm = TRUE),
    sndem_change = sndem_2023 - sndem_2000,
    .groups = "drop"
  ) %>%
  filter(!is.na(sndem_change))

# National median ribbon
nat_ribbon <- master_clean2 %>%
  group_by(year) %>%
  summarise(
    med = median(sndem_norm, na.rm = TRUE),
    q25 = quantile(sndem_norm, 0.25, na.rm = TRUE),
    q75 = quantile(sndem_norm, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# Sub-index medians
subindex_nat <- master_clean2 %>%
  group_by(year) %>%
  summarise(
    elect_med = median(snelect_norm,  na.rm = TRUE),
    civil_med = median(sncivlib_norm, na.rm = TRUE),
    .groups = "drop"
  )

# Event annotations
events <- tibble(
  year  = c(2005, 2016),
  label = c("AUC\nDemobilization", "FARC-EP\nAgreement")
)

# Focal municipality codes (confirmed)
focal_codes <- c("05197", "19809", "50573",
                 "41349", "41483", "05045", "73461")

focal_meta <- tibble(
  MPIO_CDPMP = focal_codes,
  case_label = c(
    "Extreme Improver (Cocorná)",
    "Extreme Decliner (Timbiquí)",
    "Typical Case (Puerto López)",
    "Most-Similar A (Hobo)",
    "Most-Similar B (Nátaga)",
    "Most-Different A (Apartadó)",
    "Most-Different B (Murillo)"
  ),
  case_group = c(
    "Extreme", "Extreme", "Typical",
    "Most-Similar", "Most-Similar",
    "Most-Different", "Most-Different"
  )
)
```

### Distributional Overview

```{r overview, echo=FALSE, fig.width=9, fig.height=5, fig.cap="National-median trajectory of the composite subnational democracy score (sndem) with interquartile range and sub-index medians. Dotted vertical lines mark the 2005 AUC demobilization and 2016 FARC-EP peace agreement. Grey ribbon = interquartile range across municipalities. Source: SNDEM 2026."}

p_dist <- ggplot() +
  geom_ribbon(
    data = nat_ribbon,
    aes(x = year, ymin = q25, ymax = q75),
    fill = "grey85", alpha = 0.6
  ) +
  geom_line(
    data = nat_ribbon,
    aes(x = year, y = med, linetype = "Composite (sndem)"),
    color = "grey20", linewidth = 0.9
  ) +
  geom_line(
    data = subindex_nat,
    aes(x = year, y = elect_med, linetype = "Electoral (snelect)"),
    color = "#0072B2", linewidth = 0.7
  ) +
  geom_line(
    data = subindex_nat,
    aes(x = year, y = civil_med, linetype = "Civil Liberties (sncivlib)"),
    color = "#D55E00", linewidth = 0.7
  ) +
  geom_vline(
    data = events,
    aes(xintercept = year),
    linetype = "dotted", color = "grey40", linewidth = 0.6
  ) +
  geom_text(
    data = events,
    aes(x = year - 0.3, y = 0.97, label = label),
    size = 2.8, hjust = 1, color = "grey30", lineheight = 0.85
  ) +
  scale_linetype_manual(
    name   = NULL,
    values = c(
      "Composite (sndem)"          = "solid",
      "Electoral (snelect)"        = "dashed",
      "Civil Liberties (sncivlib)" = "dotdash"
    )
  ) +
  scale_x_continuous(breaks = seq(2000, 2023, 4)) +
  scale_y_continuous(limits = c(0.3, 1.0), breaks = seq(0.3, 1.0, 0.1)) +
  labs(x = NULL, y = "Normalized democracy score") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )

p_dist
```

The figure below shows the national-median trajectory of the
composite subnational democracy score (`sndem`) alongside its interquartile
range across municipalities, with the two sub-index medians shown
separately. Three patterns are immediately apparent. First, average
municipal democracy declined modestly from 2000 to approximately 2010 --- a
period corresponding to the intensification of counterinsurgency operations
under the Álvaro Uribe administration --- before recovering strongly after
2016, with the median normalized score rising from 0.522 to 0.730 between
2010 and 2023. Second, this recovery is widespread: the interquartile range
shifts upward across the full distribution, suggesting broad-based rather
than concentrated improvement after the FARC-EP comprehensive peace
agreement. Third, the electoral fairness and civil liberties sub-indices
track one another closely at the national level, though we show below that
they diverge more substantially at the municipal level, underscoring the
value of measuring both dimensions separately.

### Case Selection

```{r focalmpios, echo=FALSE, fig.width=10, fig.height=6, fig.cap="Normalized sndem trajectories for focal municipalities (2000--2023). Grey ribbon = national interquartile range; dashed grey = national median. Dotted vertical lines mark the 2005 AUC demobilization and 2016 FARC-EP peace agreement. Source: SNDEM 2024."}

focal_ts <- master_clean2 %>%
  filter(MPIO_CDPMP %in% focal_codes) %>%
  left_join(focal_meta, by = "MPIO_CDPMP")

# --- ONE COLOR + LINETYPE PER MUNICIPALITY ---
# Colors: colorblind-safe palette, grouped by case type
muni_colors <- c(
  "05197" = "#D55E00",  # Cocorná       - extreme improver  (orange-red)
  "19809" = "#0072B2",  # Timbiquí      - extreme decliner  (blue)
  "50573" = "#757575",  # Puerto López  - typical           (grey)
  "41349" = "#009E73",  # Hobo          - most-similar A    (green)
  "41483" = "#56B4E9",  # Nátaga        - most-similar B    (sky blue)
  "05045" = "#CC79A7",  # Apartadó      - most-different A  (pink)
  "73461" = "#E69F00"   # Murillo       - most-different B  (gold)
)

# Solid = "A" or standalone; dashed = "B" or decliner
muni_linetypes <- c(
  "05197" = "solid",   # Cocorná
  "19809" = "dashed",  # Timbiquí
  "50573" = "solid",   # Puerto López
  "41349" = "solid",   # Hobo
  "41483" = "dashed",  # Nátaga
  "05045" = "solid",   # Apartadó
  "73461" = "dashed"   # Murillo
)

# Short labels for direct annotation
muni_short <- c(
  "05197" = "Cocorná",
  "19809" = "Timbiquí",
  "50573" = "Puerto López",
  "41349" = "Hobo",
  "41483" = "Nátaga",
  "05045" = "Apartadó",
  "73461" = "Murillo"
)

# Case-type annotation (right margin)
case_annot <- tibble(
  x     = 2028,
  y     = c(0.93, 0.64, 0.50, 0.38, 0.22),
  label = c(
    "◆ Extreme cases",
    "◆ Most-different",
    "◆ Typical",
    "◆ Most-similar",
    ""
  ),
  color = c("#D55E00", "#CC79A7", "#757575", "#009E73", "white")
)

# End-of-line label data (last observed year per municipality)
label_data <- focal_ts %>%
  group_by(MPIO_CDPMP) %>%
  slice_max(year, n = 1) %>%
  ungroup() %>%
  mutate(short_label = muni_short[MPIO_CDPMP])

p_cases <- ggplot() +
  # --- BACKGROUND ---
  geom_ribbon(
    data  = nat_ribbon,
    aes(x = year, ymin = q25, ymax = q75),
    fill  = "grey92", alpha = 0.9
  ) +
  geom_line(
    data     = nat_ribbon,
    aes(x = year, y = med),
    color    = "grey55", linetype = "dashed", linewidth = 0.55
  ) +
  # --- EVENT LINES ---
  geom_vline(
    data      = events,
    aes(xintercept = year),
    linetype  = "dotted", color = "grey40", linewidth = 0.5
  ) +
  annotate("text", x = 2005.2, y = 0.115,
           label = "AUC\nDemobilization",
           size = 2.6, hjust = 0, color = "grey35", lineheight = 0.85) +
  annotate("text", x = 2016.2, y = 0.115,
           label = "FARC-EP\nAgreement",
           size = 2.6, hjust = 0, color = "grey35", lineheight = 0.85) +
  # --- FOCAL LINES ---
  geom_line(
    data = focal_ts,
    aes(
      x        = year,
      y        = sndem_norm,
      color    = MPIO_CDPMP,
      linetype = MPIO_CDPMP,
      group    = MPIO_CDPMP
    ),
    linewidth = 1.0
  ) +
  # --- DIRECT END-OF-LINE LABELS ---
  ggrepel::geom_label_repel(
    data         = label_data,
    aes(x = year, y = sndem_norm,
        label = short_label, color = MPIO_CDPMP),
    nudge_x      = 1.5,
    direction    = "y",
    hjust        = 0,
    size         = 3.0,
    fontface     = "bold",
    segment.size = 0.35,
    segment.color = "grey50",
    label.size   = 0.15,
    fill         = alpha("white", 0.85),
    show.legend  = FALSE,
    min.segment.length = 0
  ) +
  # --- SCALES ---
  scale_color_manual(values = muni_colors,    guide = "none") +
  scale_linetype_manual(values = muni_linetypes, guide = "none") +
  scale_x_continuous(
    breaks = seq(2000, 2023, 4),
    limits = c(1999, 2029),   # right margin for labels
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    limits = c(0.10, 1.02),
    breaks = seq(0.2, 1.0, 0.2)
  ) +
  # --- CASE-TYPE LEGEND AS ANNOTATION ---
  annotate("text", x = 2027.5, y = 0.96,
           label = "Case type", size = 3, fontface = "bold",
           hjust = 1, color = "grey20") +
  annotate("segment",
           x = 2024.5, xend = 2026.5,
           y = c(0.91, 0.82, 0.73, 0.64),
           yend = c(0.91, 0.82, 0.73, 0.64),
           color = c("#D55E00", "#CC79A7", "#009E73", "#757575"),
           linewidth = 1.2,
           linetype = c("solid", "solid", "solid", "solid")) +
  annotate("text",
           x = 2026.8,
           y = c(0.91, 0.82, 0.73, 0.64),
           label = c("Extreme", "Most-different",
                     "Most-similar", "Typical"),
           size = 2.7, hjust = 0, color = "grey20") +
  # --- LABELS ---
  labs(
    x = NULL,
    y = "Normalized democracy score",
    caption = paste(
      "Grey ribbon = national IQR; dashed grey line = national median.",
      "Solid lines = primary case in each pair; dashed = comparison case.",
      sep = "\n"
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    plot.caption     = element_text(size = 8, color = "grey40",
                                    hjust = 0, margin = margin(t = 8))
  )

p_cases

# Faceted version: one panel per case type
focal_ts_facet <- focal_ts %>%
  mutate(
    case_group = factor(case_group,
      levels = c("Extreme", "Typical", "Most-Similar", "Most-Different")),
    short_label = muni_short[MPIO_CDPMP]
  )

p_facet <- ggplot() +
  # National ribbon in every panel
  geom_ribbon(
    data = nat_ribbon %>%
      crossing(case_group = factor(
        c("Extreme","Typical","Most-Similar","Most-Different"),
        levels = c("Extreme","Typical","Most-Similar","Most-Different")
      )),
    aes(x = year, ymin = q25, ymax = q75),
    fill = "grey92", alpha = 0.9
  ) +
  geom_line(
    data = nat_ribbon %>%
      crossing(case_group = factor(
        c("Extreme","Typical","Most-Similar","Most-Different"),
        levels = c("Extreme","Typical","Most-Similar","Most-Different")
      )),
    aes(x = year, y = med),
    color = "grey55", linetype = "dashed", linewidth = 0.5
  ) +
  # Event lines
  geom_vline(xintercept = c(2005, 2016),
             linetype = "dotted", color = "grey50", linewidth = 0.4) +
  # Focal lines, colored by municipality
  geom_line(
    data = focal_ts_facet,
    aes(x = year, y = sndem_norm,
        color = MPIO_CDPMP, linetype = MPIO_CDPMP,
        group = MPIO_CDPMP),
    linewidth = 1.1
  ) +
  # Direct labels at line ends
  ggrepel::geom_label_repel(
    data = focal_ts_facet %>%
      group_by(MPIO_CDPMP, case_group) %>%
      slice_max(year, n = 1),
    aes(x = year, y = sndem_norm,
        label = short_label, color = MPIO_CDPMP),
    nudge_x = 1.2, direction = "y", hjust = 0,
    size = 2.8, fontface = "bold",
    segment.size = 0.3, label.size = 0.15,
    fill = alpha("white", 0.85),
    show.legend = FALSE
  ) +
  scale_color_manual(values = muni_colors, guide = "none") +
  scale_linetype_manual(values = muni_linetypes, guide = "none") +
  scale_x_continuous(breaks = seq(2000, 2023, 8),
                     limits = c(1999, 2028), expand = c(0,0)) +
  scale_y_continuous(limits = c(0.10, 1.02),
                     breaks = seq(0.2, 1.0, 0.2)) +
  facet_wrap(~ case_group, nrow = 2) +
  labs(
    x       = NULL,
    y       = "Normalized democracy score",
    caption = "Grey ribbon = national IQR. Dotted lines = 2005 AUC demobilization and 2016 FARC-EP agreement."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position   = "none",
    panel.grid.minor  = element_blank(),
    strip.text        = element_text(face = "bold", size = 11),
    strip.background  = element_rect(fill = "grey95", color = NA),
    plot.caption      = element_text(size = 8, color = "grey40", hjust = 0)
  )

p_facet

```


To illustrate the inferential possibilities opened by these data, we
identify five municipalities --- representing four types of democratic
trajectory --- that together capture the range of variation in the data.
The figure below plots their normalized `sndem` trajectories
alongside the national median and interquartile range. The table
following the case descriptions summarizes their covariate profiles.

**Extreme improver: Cocorná (Antioquia).** Cocorná registered the largest
democratic improvement of any municipality in our data, with its normalized
`sndem` score rising from 0.394 in 2000 to 0.910 in 2023 --- a gain of
0.517 points. Cocorná is located in the *Oriente Antioqueño*, a subregion
that experienced some of the worst paramilitary and guerrilla violence in
Colombia during the late 1990s and early 2000s, including large-scale
forced displacement. Notably, four other Oriente Antioqueño municipalities
--- San Carlos, San Francisco, San Luis, and Argelia --- appear among the
top ten improvers in the data, suggesting that the index is capturing a
coherent regional pattern of post-conflict democratic recovery rather than
an idiosyncratic outlier. Importantly, this recovery is not simply a
reflection of socioeconomic gains: Cocorná's average water access coverage
stands at only 42%, and it remains a small rural municipality (average
population \textasciitilde{}17,000). Its improvement in `sndem` is driven
primarily by the electoral fairness component, consistent with accounts of
normalized local electoral competition following the demobilization of
armed actors in the region.

**Extreme decliner: Timbiquí (Cauca).** At the other end of the spectrum,
Timbiquí --- a predominantly Afro-Colombian municipality on Colombia's
Pacific coast --- experienced the largest democratic decline of any
municipality with usable covariate data, falling from 0.608 in 2000 to
0.331 in 2023 ($-$0.278 points). Its trajectory represents a pattern
documented in the post-conflict literature: as FARC structures dismantled
following the 2016 agreement, criminal organizations moved into Pacific
coast territories with more predatory and less institutionally predictable
forms of violence, producing deteriorating conditions for both civil
liberties and electoral integrity. Timbiquí's decline accelerates precisely
after 2016 --- the opposite of the national trend --- illustrating how
aggregate post-conflict improvement can coexist with severe democratic
deterioration in specific peripheral territories. We note that several
frontier municipalities in Amazonia and the Llanos Orientales also show
sharp declines but are excluded from covariate-based comparisons due to
sparse DANE administrative coverage; their trajectories represent a data
limitation that future iterations of the method should address.

**Typical case: Puerto López (Meta).** Puerto López most closely tracks the
national median trajectory across the full 24-year period, with a mean
squared deviation of 0.000424 from the annual national median --- less than
half that of the next-closest candidate. Located in the Llanos Orientales,
neither a conflict epicenter nor a development showcase, it serves as a
useful reference point for what an ``average'' Colombian municipality looks
like on our index. Its score rises from 0.627 in 2000 to 0.748 in 2023,
tracking the national post-2016 recovery closely.

**Most-similar municipalities, different outcomes: Hobo and Nátaga
(Huila).** The most-similar design logic asks: can we identify pairs of
municipalities that are structurally similar on observable dimensions yet
exhibit meaningfully different democratic outcomes? Our algorithmic search
surfaces Hobo and Nátaga, neighboring municipalities in the department of
Huila, as the cleanest such pair in the data. Their covariate profiles are
strikingly similar: water access coverage (71.8\% vs.\ 70.6\%), infant
mortality (17.8 vs.\ 19.0 per 1,000), net school enrollment (92.4\%
vs.\ 92.9\%), average SABER 11 scores (48.1 vs.\ 48.2), and population
(\textasciitilde{}7,000 vs.\ \textasciitilde{}6,000) are nearly identical
across all five dimensions used in the matching procedure. Yet their
average normalized `sndem` scores diverge by 0.240 points (0.644 vs.\
0.404), a gap that cannot be attributed to differences in any of the
standard socioeconomic suspects. Notably, this gap narrows over the study
period: Hobo declines slightly while Nátaga improves, particularly after
2016, suggesting some convergence. The persistence of the gap despite
matched covariates opens a productive research question: what accounts for
Hobo's democratic advantage? Our data identify it but cannot by themselves
resolve it --- which is precisely the kind of leverage a thick subnational
democracy measure provides for subsequent qualitative or quantitative
investigation.

**Most-different municipalities, similar outcomes: Apartadó (Antioquia)
and Murillo (Tolima).** The complementary most-different design asks: can
municipalities arrive at similar democratic outcomes via entirely different
structural pathways? Apartadó and Murillo have nearly identical average
normalized `sndem` scores across the study period (0.595 vs.\ 0.600), yet
their covariate profiles diverge sharply. Apartadó is a mid-sized city in
the Urabá region (\textasciitilde{}109,000 average population) with
relatively high water access (83.5\%) and low infant mortality (12.3 per
1,000), historically defined by banana plantation labor politics and
intense paramilitary violence. Murillo is a small, remote Andean
\textit{páramo} municipality in northern Tolima (\textasciitilde{}4,000
population) with poor water access (30.9\%) and higher infant mortality
(23.9 per 1,000). Their convergence on similar democratic scores despite
these structural differences --- and despite very different temporal
trajectories, with Apartadó remaining relatively stable and Murillo
improving markedly after 2016 --- illustrates that our index is not simply
recovering socioeconomic or geographic gradients already observable in the
data.

```{r covsprof, echo=FALSE, results='asis'}
profile_table <- muni_avg %>%
  filter(MPIO_CDPMP %in% focal_codes) %>%
  left_join(
    focal_meta %>% select(MPIO_CDPMP, case_label),
    by = "MPIO_CDPMP"
  ) %>%
  select(
    case_label, municipio, depto,
    sndem_avg, sndem_2000, sndem_2023, sndem_change,
    Acueducto_avg, MortInfantil1_avg,
    NetCover_avg, AvgS11_avg, PobTot_12_avg
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  mutate(case_label = factor(case_label, levels = focal_meta$case_label)) %>%
  arrange(case_label) %>%
  select(-case_label)

cat("{\\small\n")
kable(
  profile_table,
  format    = "pipe",
  caption   = "Covariate profiles of focal municipalities. Democracy scores normalized (0-1). Control variables are period averages (2000-2023). Water = % households with piped water; Inf. Mort. = deaths per 1,000 live births; SABER = ICFES SABER 11 standardized exam; Pop = Population",
  col.names = c(
    "Municipality", "Department",
    "Avg.", "2000", "2023", "Change",
    "Water (%)", "Inf. Mort.",
    "Enroll. (%)", "SABER", "Pop."
  ),
  align  = "llrrrrrrrrrr",
  digits = 2
)
cat("}\n")
```

```{r zoommaps, echo=FALSE, fig.width=10, fig.height=8, fig.cap="Municipal democracy scores in Colombia (2023). The context map (top left) shows the full national distribution of normalized sndem scores; dashed boxes indicate the three zoomed regions. Labeled dots mark focal case municipalities. Source: SNDEM 2026."}


# Reload geometry if not in environment from facets chunk
if (!exists("muni_geo")) {
  muni_geo <- st_read(
    "G:/Shared drives/snvdem/snvdem-col/data/geospatial/2018pmq/BaseLayer/MGN_ANM_MPIOS.shp",
    quiet = TRUE
  ) %>%
    mutate(MPIO_CDPMP = clean_mpio(MPIO_CDPMP))
  muni_geo$MPIO_CDPMP <- as.character(muni_geo$MPIO_CDPMP)
}

if (!exists("neighbors")) {
  neighbors <- ne_countries(scale = "medium", returnclass = "sf") %>%
    filter(continent == "South America" | name == "Panama")
}

snap_year <- 2023

map_snap <- muni_geo %>%
  left_join(
    master_clean2 %>%
      filter(year == snap_year) %>%
      select(MPIO_CDPMP, sndem_norm),
    by = "MPIO_CDPMP"
  )

# Build focal_geo: join municipio name from snvdem, not from shapefile
focal_geo <- muni_geo %>%
  filter(MPIO_CDPMP %in% focal_codes) %>%
  st_centroid() %>%
  left_join(focal_meta, by = "MPIO_CDPMP") %>%
  left_join(
    snvdem %>%
      distinct(MPIO_CDPMP, municipio) %>%
      mutate(MPIO_CDPMP = clean_mpio(MPIO_CDPMP)),
    by = "MPIO_CDPMP"
  ) %>%
  left_join(
    master_clean2 %>%
      filter(year == snap_year) %>%
      select(MPIO_CDPMP, sndem_norm),
    by = "MPIO_CDPMP"
  ) %>%
  mutate(
    label_text = paste0(
      toupper(municipio), "\n",
      round(sndem_norm, 2)
    )
  )

# --- Helper: zoom panel ---
zoom_map <- function(data, focal_pts, xlim, ylim, title) {
  ggplot(data = data) +
    geom_sf(data = neighbors, fill = "#f5f5f5",
            color = "grey80", linewidth = 0.2) +
    geom_sf(aes(fill = sndem_norm), color = NA) +
    geom_sf(data = focal_pts, shape = 21, size = 4,
            fill = "white", color = "black", stroke = 1.4) +
    geom_label_repel(
      data               = focal_pts,
      aes(label = label_text, geometry = geometry),
      stat               = "sf_coordinates",
      size               = 3.5,        # was 2.8 -- bigger for print
      fontface           = "bold",
      box.padding        = 0.8,
      segment.color      = "grey30",
      segment.size       = 0.4,
      min.segment.length = 0,
      fill               = alpha("white", 0.90),
      label.size         = 0.25,
      label.padding      = unit(0.20, "lines")
    ) +
    scale_fill_viridis_c(
      option   = "magma",
      name     = "sndem\n(2023)",
      limits   = c(0, 1),
      na.value = "grey75",
      guide    = guide_colorbar(
        barheight = unit(3, "cm"),
        barwidth  = unit(0.4, "cm"),
        ticks     = FALSE,
        title.theme = element_text(size = 9, face = "bold")
      )
    ) +
    coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
    labs(title = title) +
    theme_void(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 11,
                                margin = margin(b = 5)),
      legend.position = "right",
      plot.margin = margin(4, 4, 4, 4)
    )
}

# --- ZOOM PANELS (unchanged bounding boxes) ---
zoom_antioquia <- zoom_map(
  data      = map_snap,
  focal_pts = focal_geo %>% filter(MPIO_CDPMP %in% c("05197", "05045")),
  xlim  = c(-77.0, -73.5), ylim = c(5.0, 8.5),
  title = "A: Antioquia"
)

zoom_southwest <- zoom_map(
  data      = map_snap,
  focal_pts = focal_geo %>%
    filter(MPIO_CDPMP %in% c("19809", "41349", "41483")),
  xlim  = c(-78.0, -74.0), ylim = c(0.5, 4.5),
  title = "B: Pacific Coast & Huila"
)

zoom_east <- zoom_map(
  data      = map_snap,
  focal_pts = focal_geo %>% filter(MPIO_CDPMP %in% c("50573", "73461")),
  xlim  = c(-76.5, -71.5), ylim = c(2.0, 6.5),
  title = "C: Meta & Tolima"
)

# --- CONTEXT MAP (larger text, no in-figure title) ---
context_map <- ggplot() +
  geom_rect(aes(xmin = -82, xmax = -66, ymin = -5, ymax = 13),
            fill = "#d6eaf8") +
  geom_sf(data = neighbors, fill = "#f0f0f0",
          color = "grey70", linewidth = 0.2) +
  geom_sf(data = map_snap, aes(fill = sndem_norm), color = NA) +
  scale_fill_viridis_c(option = "magma", limits = c(0, 1),
                       na.value = "grey75", guide = "none") +
  # White outer stroke
  geom_rect(data = zoom_boxes,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = NA, color = "white", linewidth = 1.2) +
  # Dashed dark stroke
  geom_rect(data = zoom_boxes,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = NA, color = "grey15", linewidth = 0.5,
            linetype = "dashed") +
  # Panel letter labels — larger for print
  geom_label(data = zoom_labels,
             aes(x = label_x, y = label_y, label = panel),
             size       = 4.0,    # was 3.2
             fontface   = "bold",
             fill       = alpha("white", 0.90),
             label.size = 0.25,
             hjust = 0, vjust = 1) +
  coord_sf(xlim = c(-82, -66), ylim = c(-5, 13), expand = FALSE) +
  theme_void(base_size = 11) +
  theme(
    panel.border = element_rect(color = "grey30", fill = NA, linewidth = 0.5),
    plot.margin  = margin(4, 4, 4, 4)
  )

# --- 2×2 LAYOUT ---
# Top row:    context map | Zoom A
# Bottom row: Zoom B      | Zoom C
layout_2x2 <- c(
  area(1, 1),  # context map: top-left
  area(1, 2),  # Zoom A:      top-right
  area(2, 1),  # Zoom B:      bottom-left
  area(2, 2)   # Zoom C:      bottom-right
)

final_zoom <- context_map +
  zoom_antioquia +
  zoom_southwest +
  zoom_east +
  plot_layout(design = layout_2x2, guides = "collect") &
  theme(legend.position = "right")

tryCatch(
  ggsave(
    "G:/Shared drives/snvdem/snvdem-col/data/panel/09_analysis_scripts/ZoomIn/SNDEM_zoom_2x2.png",
    final_zoom, width = 10, height = 8, dpi = 300, bg = "white"
  ),
  error = function(e) message("ggsave skipped: ", e$message)
)

final_zoom
```

### Two Theoretical Questions

Our data are not designed to resolve causal questions in this paper; that
is the work the method enables for future research. We nonetheless sketch
two questions for which these data are well-positioned.

**The relationship between national and subnational democracy.** The figure
above shows that the municipal distribution broadly tracks Colombia's
national democratic arc over time, but with important deviations. The
national median of municipal `sndem` dips from 0.579 (2000) to 0.522
(2010) before recovering sharply to 0.730 (2023) --- a trajectory that
parallels the shift from militarized counterinsurgency toward negotiated
peace. The 2005 AUC demobilization produces little visible movement in the
municipal distribution, while the 2016 FARC agreement is associated with a
pronounced and sustained upward shift. Critically, these aggregate trends
obscure divergent municipal trajectories: Cocorná improves dramatically
while Timbiquí deteriorates sharply in the same post-2016 period. Whether
subnational democratization leads, follows, or selectively diverges from
national trends --- and under what conditions --- is an empirical question
that our time-series data allow scholars to address systematically, and
cross-nationally, once the method is extended beyond Colombia.

**Replication versus autonomous subnational environments.** If municipal
democracy were simply a downward projection of national institutions, we
would expect low cross-municipal variance and tight correlation with
national-level indicators. The evidence presented here suggests otherwise.
Substantial variation persists across municipalities even within years of
stable national-level democracy, the most-similar pair (Hobo and Nátaga)
demonstrates that virtually identical socioeconomic conditions are
compatible with meaningfully different democratic outcomes, and the
most-different pair (Apartadó and Murillo) shows that structurally
dissimilar municipalities can converge on similar democratic scores via
entirely different pathways. Taken together, these patterns point toward
subnational environments as partially autonomous democratic arenas --- a
finding with implications for theories of democratic resilience,
autocratization, and the territorial reach of the state.
