---
title: "Measuring Democracy in Subnational Units Worldwide: A New Method"
subtitle: "Mapping snvdem in Colombia (2000--2023)"
abstract: |
  Scholars have documented that democracy does not exist evenly within 
  countries throughout the world and explained why. Increasingly, studies 
  have demonstrated that this subnational variation in democracy affects 
  other aspects of politics and society, such as state capacity, 
  democratization, violence, and environmental degradation. Further progress 
  in studying subnational democracy, however, is limited by the absence of 
  cross-national, time-series data that incorporate a thick conceptualization 
  and operationalization of democracy and cover all subnational units of many 
  countries. Current datasets of subnational unit democracy are limited to a 
  small number of countries. Moreover, most studies rely on a thin 
  conceptualization of democracy, measured with indicators such as electoral 
  margins of victory, effective numbers of candidates, or voter turnout. We 
  offer a method to generate cross-national time-series data of democracy in 
  subnational units that relies on a thick conceptualization and 
  operationalization of democracy. We combine expert data from elections and 
  civil liberties surveys in the Varieties of Democracy (V-Dem) dataset and 
  geocoded observational data. We use these data to generate annual average 
  levels of how free and fair subnational elections are, how well civil 
  liberties are protected, and overall democracy in each subnational unit of 
  a country, thus capturing not only the electoral, but also the broader 
  liberal, conceptualization of democracy. We pilot the method by generating 
  municipal democracy data for Colombia for years 2000--2023. We conduct 
  three validation tests of these data and our method. We also demonstrate 
  the utility of the data by investigating the impact of Colombia's 2016 
  Comprehensive Peace Agreement with the FARC-EP on democracy in its rural 
  municipalities. Our method can be applied to other countries, so it offers 
  the promise of generating cross-national time-series data of democracy in 
  subnational units in many countries of the world. This will enable us to 
  test the generalizability of conclusions from earlier subnational democracy 
  research and to expand the questions we can answer.
author: "Patrick McQuestion"
date: "April 2, 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: false
    fig_caption: true
    fig_width: 10
    fig_height: 7
    keep_tex: false
    latex_engine: xelatex
    extra_dependencies:
      - booktabs      # better table rules
      - longtable     # tables that span pages
      - float         # figure placement control [H]
      - caption       # caption formatting
      - geometry      # page margins (set below)
      - setspace      # line spacing
      - fancyhdr      # headers/footers
      - hyperref      # clickable TOC links
geometry: "margin=1in"
fontsize: 11pt
linestretch: 1.15
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small, labelfont=bf}
  - \captionsetup[table]{font=small, labelfont=bf}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[L]{\small Measuring Subnational Democracy}
  - \fancyhead[R]{\small McQuestion et al.\ (2026)}
  - \fancyfoot[C]{\thepage}
  - \renewcommand{\headrulewidth}{0.4pt}
---

```{r setup, include=FALSE}
# --- KNIT OPTIONS ---
knitr::opts_chunk$set(
  echo        = FALSE,    # hide code in output (set TRUE to show)
  warning     = FALSE,    # suppress warnings
  message     = FALSE,    # suppress package messages
  fig.align   = "center",
  fig.pos     = "H",      # hold figures in place (requires float package)
  out.width   = "100%",
  dpi         = 300,
  cache       = FALSE
)

# --- ROOT DIRECTORY ---
# Use root.dir instead of setwd() for reliable knitting
knitr::opts_knit$set(
  root.dir = "G:/Shared drives/snvdem/snvdem-col/data/panel/"
)

# --- LIBRARIES ---
library(tidyverse)
library(sf)
library(patchwork)
library(ggrepel)
library(viridis)
library(rnaturalearth)
library(rnaturalearthdata)
library(knitr)
library(kableExtra)   # for better PDF table formatting

# --- DATA ---
snvdem <- readRDS(
  "G:/Shared drives/snvdem/snvdem-col/data/panel/09_analysis_scripts/Exploratory/01_Expand/snvdem2.rds"
)
```

# Pilot Application: Colombia

## Calculating V-Dem weights
## Geopredictor indicators selected for Colombia

## Visualizing subnational democracy

From the data, we learn that levels of democracy have varied tremendously across Colombian municipalities between 2000 and 2023. Interestingly, over time, democracy appears to have increased the most in those municipalities that were already more democratic. The range extends consistently from relatively high levels of democracy in the area surrounding the capital Bogotá, as indicated by the yellow shading, to relatively low levels adjacent to the more southern part of the border with Venezuela, as indicated by the purple. However, in the area surrounding Bogotá the absolute level of democracy became significantly higher by 2023. This is evident in country maps from five different years 

```{r facets, include=FALSE}
# Map with legend

# --- 0. LOAD LIBRARIES ---
library(sf)
library(tidyverse)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(viridis)

# --- 1. DATA PROCESSING ---
# Load panel data
master_df <- snvdem[,1:74]

# Data Cleaning and Normalization
master_clean <- master_df %>%
  filter(!is.na(snelect) & !is.na(sncivlib) & !is.na(sndem)) %>%
  mutate(
    snelect_norm = (snelect - min(snelect)) / (max(snelect) - min(snelect)),
    sncivlib_norm = (sncivlib - min(sncivlib)) / (max(sncivlib) - min(sncivlib)),
    sndem_norm    = (sndem - min(sndem)) / (max(sndem) - min(sndem))
  )

# Function to pad MPIO codes
clean_mpio <- function(x) str_pad(as.character(as.numeric(x)), width = 5, side = "left", pad = "0")

# Load Geospatial data
muni_geo <- st_read("G:/Shared drives/snvdem/snvdem-col/data/geospatial/2018pmq/BaseLayer/MGN_ANM_MPIOS.shp") %>%
  mutate(MPIO_CDPMP = clean_mpio(MPIO_CDPMP))

# Ensure ID columns are character for join
master_clean$MPIO_CDPMP <- as.character(master_clean$MPIO_CDPMP)
muni_geo$MPIO_CDPMP <- as.character(muni_geo$MPIO_CDPMP)

# Join democracy scores to spatial data
years_to_map <- c(2000, 2006, 2012, 2018, 2023)
map_data <- muni_geo %>%
  left_join(master_clean %>% filter(year %in% years_to_map), by = "MPIO_CDPMP")

# --- 2. GET CONTEXTUAL DATA ---
world <- ne_countries(scale = "medium", returnclass = "sf")
neighbors <- world %>% filter(continent == "South America" | name %in% c("Panama"))
col_departments <- ne_states(country = "Colombia", returnclass = "sf")

# --- 3. THE STYLIZED CONTEXT MAP (Adjusted Labels) ---
stylized_context_map <- ggplot() +
  # Ocean background
  geom_rect(aes(xmin = -82, xmax = -66, ymin = -6, ymax = 14), fill = "#e0f3ff") +
  # Neighbor countries (Light Green)
  geom_sf(data = neighbors, fill = "#d5e8d4", color = "grey70") + 
  # Colombia Departments (White)
  geom_sf(data = col_departments, fill = "white", color = "grey40", linewidth = 0.2) + 
  
  # City labels with pointer lines (ggrepel)
  geom_label_repel(
    data = data.frame(
      name = c("Bogotá", "Medellín", "Cali", "Barranquilla", "Cartagena", "Bucaramanga"),
      x = c(-74.07, -75.56, -76.52, -74.78, -75.48, -73.12),
      y = c(4.71, 6.25, 3.45, 10.96, 10.40, 7.12)
    ),
    aes(x = x, y = y, label = name),
    size = 2.8,
    fontface = "bold",
    box.padding = 0.6, 
    point.padding = 0.3,
    segment.color = "grey30",
    min.segment.length = 0,
    fill = alpha("white", 0.8)
  ) +
  
  # Geographic Features and Neighbor Labels
  annotate("text", x = -78, y = 12.5, label = "Caribbean Sea", size = 3, fontface = "italic", color = "#4a90e2") +
  annotate("text", x = -80.5, y = 4, label = "Pacific Ocean", size = 3, fontface = "italic", angle = 90, color = "#4a90e2") +
  
  # ADJUSTED COORDINATES BELOW:
  annotate("text", x = -70, y = 10, label = "VENEZUELA", size = 3.2, color = "grey40", fontface = "bold") +
  annotate("text", x = -68.5, y = -3, label = "BRAZIL", size = 3.2, color = "grey40", fontface = "bold") + # Moved East and South
  annotate("text", x = -78.5, y = -1.5, label = "ECUADOR", size = 3.2, color = "grey40", fontface = "bold") + # Moved West and South
  annotate("text", x = -74.5, y = -5, label = "PERU", size = 3.2, color = "grey40", fontface = "bold") + # Nudged for balance
  annotate("text", x = -79.5, y = 8.8, label = "PANAMA", size = 2.8, color = "grey40", fontface = "bold") +
  
  coord_sf(xlim = c(-82, -66), ylim = c(-6, 14), expand = FALSE) +
  theme_void() +
  theme(panel.border = element_rect(colour = "grey20", fill = NA, size = 0.5))

# --- 4. THE FACETED TIME-SERIES MAPS ---
facet_maps_list <- lapply(years_to_map, function(y) {
  ggplot(data = map_data %>% filter(year == y)) +
    # Background rectangle to simulate sea color in facets
    geom_rect(aes(xmin = -79.5, xmax = -66.5, ymin = -4.5, ymax = 12.5), fill = "#f0f8ff") +
    geom_sf(data = neighbors, fill = "#f2f9f2", color = "grey90", linewidth = 0.05) + 
    geom_sf(aes(fill = sndem_norm), color = NA) +
    scale_fill_viridis_c(
      option = "magma",
      name = "Democracy\nScore",
      na.value = "grey80",
      limits = c(0, 1),
      guide = guide_colorbar(barheight = unit(3, "cm"), ticks = FALSE)
    ) +
    coord_sf(xlim = c(-79.5, -66.5), ylim = c(-4.5, 12.5), expand = FALSE) +
    labs(title = as.character(y)) +
    theme_void() +
    theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5, margin = margin(t = 5, b = 5)),
      legend.position = "none" # Temporarily remove to collect later
    )
})

# --- 5. ASSEMBLE WITH PATCHWORK ---
# Use area() to ensure rectangular patches and avoid the layout error
layout_stable <- c(
  area(1, 1), # Context Map
  area(1, 2), # 2000
  area(1, 3), # 2006
  area(2, 1), # 2012
  area(2, 2), # 2018
  area(2, 3)  # 2023
)

# Combine using patchwork
final_visual <- stylized_context_map + 
  facet_maps_list[[1]] + # 2000
  facet_maps_list[[2]] + # 2006
  facet_maps_list[[3]] + # 2012
  facet_maps_list[[4]] + # 2018
  facet_maps_list[[5]] + # 2023
  plot_layout(design = layout_stable, guides = 'collect') + 
  plot_annotation(
    title = "Figure 4: Spatial Evolution of Subnational Democracy in Colombia (2000-2023)",
    subtitle = "Normalized sndem Index (0 = Low, 1 = High)",
    caption = "Source: SNDEM 2024."
  ) & theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey30", size = 12),
    legend.position = "right"
  )

# --- 6. EXPORT ---
ggsave("G:/Shared drives/snvdem/snvdem-col/data/panel/08_final_snvdem_data/MC/imgs/SNDEM_map_legend4.png", final_visual, width = 14, height = 10, dpi = 300, bg = "white")



```


## Zooming In: What the Colombia Data Tell Us

This section has discussed the "ingredients" (dimensions, variation, covariates) and "recipe" (formulas and calibration) for estimating levels of subnational democracy in Colombia. 
-	The "cake is baked," so what do these estimations tell us about SN democracy anyway? 

- Descriptive analysis of the data
Case selection: 
a) extreme cases: some that dramatically improved or got worse over time? 
b) representative/typical cases: the municipality whose score has tracked the median or average tendency
c) comparative case study, using control variables:
ci)	2 municipalities that are "most-similar" with different democratic outcomes? Maybe they share a border…
cii) 2 municipalities that are most-different with similar democratic outcomes?

- Two theoretical questions, for which our data could be helpful:
a) How are these estimates related to what is happening at the national level? 
ai) Estimating effects around pivotal moments (e.g., election years; 2005 AUC demobilization; 2016 FARC demobilization)
b) Are we expecting a replication of national institutions, or expecting to understand subnational environments as distinct?



```{r extremes}
# --- SETUP ---
library(tidyverse)
library(sf)
library(patchwork)
library(ggrepel)
library(viridis)
library(rnaturalearth)
library(rnaturalearthdata)

# Normalize sndem to 0-1 for consistent plotting
master_clean2 <- snvdem %>%
  filter(!is.na(sndem)) %>%
  mutate(
    sndem_norm    = (sndem    - min(sndem,    na.rm = TRUE)) / 
                   (max(sndem,    na.rm = TRUE) - min(sndem,    na.rm = TRUE)),
    snelect_norm  = (snelect  - min(snelect,  na.rm = TRUE)) / 
                   (max(snelect,  na.rm = TRUE) - min(snelect,  na.rm = TRUE)),
    sncivlib_norm = (sncivlib - min(sncivlib, na.rm = TRUE)) / 
                   (max(sncivlib, na.rm = TRUE) - min(sncivlib, na.rm = TRUE))
  )

# Control variable subset for case matching
ctrl_vars <- c("Acueducto", "MortInfantil1", "NetCover", "AvgS11", "PobTot_12")

# Municipality-level averages (for cross-sectional comparisons)
muni_avg <- master_clean2 %>%
  group_by(MPIO_CDPMP, municipio, depto, DPTO_CCDGO) %>%
  summarise(
    across(all_of(ctrl_vars), ~ mean(.x, na.rm = TRUE), .names = "{.col}_avg"),
    sndem_avg      = mean(sndem_norm, na.rm = TRUE),
    snelect_avg    = mean(snelect_norm, na.rm = TRUE),
    sncivlib_avg   = mean(sncivlib_norm, na.rm = TRUE),
    sndem_2000     = mean(sndem_norm[year == 2000], na.rm = TRUE),
    sndem_2023     = mean(sndem_norm[year == 2023], na.rm = TRUE),
    sndem_change   = sndem_2023 - sndem_2000,
    .groups = "drop"
  ) %>%
  filter(!is.na(sndem_change))
```

```{r extremes, include=FALSE}
# Top 10 improvers and decliners (2000-2023)
top_improvers <- muni_avg %>% 
  slice_max(sndem_change, n = 10) %>%
  select(municipio, depto, sndem_2000, sndem_2023, sndem_change, 
         all_of(paste0(ctrl_vars, "_avg")))

top_decliners <- muni_avg %>% 
  slice_min(sndem_change, n = 10) %>%
  select(municipio, depto, sndem_2000, sndem_2023, sndem_change,
         all_of(paste0(ctrl_vars, "_avg")))

print(top_improvers)
print(top_decliners)

# After reviewing output, select your focal improver and decliner
# and store them for labeling in maps and time-series plots:
# focal_improver <- "MUNICIPIO_CODE"
#        focal_decliner <- "MUNICIPIO_CODE"

```

```{r typicals, include=FALSE}
# National median trajectory
nat_median <- master_clean2 %>%
  group_by(year) %>%
  summarise(nat_med = median(sndem_norm, na.rm = TRUE), .groups = "drop")

# For each municipality, compute mean squared deviation from national median
typical_case <- master_clean2 %>%
  left_join(nat_median, by = "year") %>%
  group_by(MPIO_CDPMP, municipio, depto) %>%
  summarise(
    msd = mean((sndem_norm - nat_med)^2, na.rm = TRUE),
    n_years = n(),
    .groups = "drop"
  ) %>%
  filter(n_years >= 20) %>%   # require near-complete time series
  slice_min(msd, n = 5)

print(typical_case)
```


```{r select1, include=FALSE}
snvdem %>%
  filter(municipio %in% c("Cocorná", "San Carlos", "San Luis", 
                           "San Francisco", "Argelia"),
         year %in% c(2000, 2005, 2010, 2016, 2023)) %>%
  select(municipio, depto, year, sndem) %>%
  mutate(sndem = round(sndem, 3)) %>%
  pivot_wider(names_from = year, values_from = sndem) %>%
  print()


# Confirm MPIO codes for focal cases using code, not name
focal_check <- snvdem %>%
  filter(
    (municipio == "Cocorná"     & depto == "Antioquia") |
    (municipio == "Timbiquí"    & depto == "Cauca")     |
    (municipio == "Puerto López"& depto == "Meta")
  ) %>%
  distinct(MPIO_CDPMP, municipio, depto)

print(focal_check)

# Get MPIO codes for the five Antioquia improvers first
oriente_codes <- snvdem %>%
  filter(
    municipio %in% c("Cocorná","San Carlos","San Francisco",
                     "San Luis","Argelia"),
    depto == "Antioquia"
  ) %>%
  distinct(MPIO_CDPMP, municipio)

# Clean pivot by code
snvdem %>%
  filter(MPIO_CDPMP %in% oriente_codes$MPIO_CDPMP,
         year %in% c(2000, 2005, 2010, 2016, 2023)) %>%
  select(MPIO_CDPMP, municipio, year, sndem) %>%
  left_join(oriente_codes, by = c("MPIO_CDPMP","municipio")) %>%
  mutate(sndem = round(sndem, 3)) %>%
  pivot_wider(names_from = year, values_from = sndem) %>%
  print()


```


```{r mostsim_dept, include=FALSE}
library(proxy)

# Control variables to use for similarity
ctrl_vars <- c("Acueducto_avg", "MortInfantil1_avg", 
               "NetCover_avg", "AvgS11_avg", "PobTot_12_avg")

# Use muni_avg from Chunk 1 (already computed)
# Keep only municipalities with complete control data
muni_complete <- muni_avg %>%
  filter(if_all(all_of(ctrl_vars), ~ !is.na(.x)))

# Standardize controls
muni_scaled <- muni_complete %>%
  mutate(across(all_of(ctrl_vars), 
                ~ scale(.x)[,1], 
                .names = "{.col}_z"))

ctrl_z_vars <- paste0(ctrl_vars, "_z")

# Within-department pairwise search
dept_pair_results <- muni_scaled %>%
  group_by(DPTO_CCDGO, depto) %>%
  filter(n() >= 3) %>%
  group_modify(~ {
    df <- .x
    n <- nrow(df)
    
    cov_mat <- as.matrix(df[, ctrl_z_vars])
    cov_mat[is.na(cov_mat)] <- 0
    dist_mat <- as.matrix(dist(cov_mat))
    
    # Build pair tibble
    idx <- which(upper.tri(dist_mat), arr.ind = TRUE)
    tibble(
      MPIO_a     = df$MPIO_CDPMP[idx[,1]],
      name_a     = df$municipio[idx[,1]],
      MPIO_b     = df$MPIO_CDPMP[idx[,2]],
      name_b     = df$municipio[idx[,2]],
      cov_dist   = dist_mat[idx],
      sndem_a    = df$sndem_avg[idx[,1]],
      sndem_b    = df$sndem_avg[idx[,2]],
      sndem_diff = abs(df$sndem_avg[idx[,1]] - df$sndem_avg[idx[,2]])
    )
  }) %>%
  ungroup()

# Rank: most similar on controls, most different on sndem
dept_pair_results <- dept_pair_results %>%
  mutate(
    cov_dist_pct   = percent_rank(cov_dist),     # low = more similar
    sndem_diff_pct = percent_rank(sndem_diff),    # high = more different
    score_msdd     = sndem_diff_pct - cov_dist_pct  # maximize
  )

# Top 15 most-similar / different-outcome pairs
top_msdd <- dept_pair_results %>%
  slice_max(score_msdd, n = 15) %>%
  select(depto, name_a, name_b, 
         cov_dist, sndem_diff, sndem_a, sndem_b, score_msdd) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

print(top_msdd)
```



```{r mostdiff_dept, include=FALSE}
# Cross-department: different controls, similar sndem
# Use one representative per department (the one closest to dept median sndem)
dept_reps <- muni_scaled %>%
  group_by(DPTO_CCDGO, depto) %>%
  mutate(dept_med = median(sndem_avg, na.rm = TRUE)) %>%
  slice_min(abs(sndem_avg - dept_med), n = 1) %>%
  ungroup()

# Pairwise across all rep municipalities
n_reps <- nrow(dept_reps)
cross_idx <- which(upper.tri(matrix(0, n_reps, n_reps)), arr.ind = TRUE)

cov_mat_reps <- as.matrix(
  dept_reps[, paste0(ctrl_vars, "_z")]
)
cov_mat_reps[is.na(cov_mat_reps)] <- 0
dist_mat_reps <- as.matrix(dist(cov_mat_reps))

cross_pairs <- tibble(
  depto_a    = dept_reps$depto[cross_idx[,1]],
  name_a     = dept_reps$municipio[cross_idx[,1]],
  depto_b    = dept_reps$depto[cross_idx[,2]],
  name_b     = dept_reps$municipio[cross_idx[,2]],
  cov_dist   = dist_mat_reps[cross_idx],
  sndem_a    = dept_reps$sndem_avg[cross_idx[,1]],
  sndem_b    = dept_reps$sndem_avg[cross_idx[,2]],
  sndem_diff = abs(dept_reps$sndem_avg[cross_idx[,1]] - 
                   dept_reps$sndem_avg[cross_idx[,2]])
) %>%
  # Exclude same-department pairs just in case
  filter(depto_a != depto_b) %>%
  mutate(
    cov_dist_pct   = percent_rank(cov_dist),
    sndem_diff_pct = percent_rank(sndem_diff),
    # Maximize covariate distance, minimize sndem difference
    score_mdso     = cov_dist_pct - sndem_diff_pct
  )

top_mdso <- cross_pairs %>%
  slice_max(score_mdso, n = 15) %>%
  select(depto_a, name_a, depto_b, name_b,
         cov_dist, sndem_diff, sndem_a, sndem_b, score_mdso) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

print(top_mdso)

```

```{r ts_plots, include=FALSE}
# 1. Confirm all MPIO codes
focal_check <- snvdem %>%
  filter(
    (municipio == "Cocorná"      & depto == "Antioquia") |
    (municipio == "Timbiquí"     & depto == "Cauca")     |
    (municipio == "Puerto López" & depto == "Meta")       |
    (municipio == "Hobo"         & depto == "Huila")      |
    (municipio == "Nátaga"       & depto == "Huila")      |
    (municipio == "Apartadó"     & depto == "Antioquia")  |
    (municipio == "Murillo"      & depto == "Tolima")
  ) %>%
  distinct(MPIO_CDPMP, municipio, depto) %>%
  arrange(depto, municipio)

print(focal_check, width = 200)

# 2. Time series for Hobo vs. Nátaga
snvdem %>%
  filter(
    (municipio == "Hobo"   & depto == "Huila") |
    (municipio == "Nátaga" & depto == "Huila")
  ) %>%
  select(MPIO_CDPMP, municipio, year, sndem, snelect, sncivlib) %>%
  mutate(across(c(sndem, snelect, sncivlib), ~ round(.x, 3))) %>%
  filter(year %in% c(2000, 2005, 2010, 2016, 2023)) %>%
  arrange(municipio, year) %>%
  print(width = 200)

# 3. Time series for Apartadó vs. Murillo
snvdem %>%
  filter(
    (municipio == "Apartadó" & depto == "Antioquia") |
    (municipio == "Murillo"  & depto == "Tolima")
  ) %>%
  select(MPIO_CDPMP, municipio, year, sndem, snelect, sncivlib) %>%
  mutate(across(c(sndem, snelect, sncivlib), ~ round(.x, 3))) %>%
  filter(year %in% c(2000, 2005, 2010, 2016, 2023)) %>%
  arrange(municipio, year) %>%
  print(width = 200)


# 4. Covariate profiles
muni_avg %>%
  filter(MPIO_CDPMP %in% focal_check$MPIO_CDPMP) %>%
  left_join(
    focal_check %>% select(MPIO_CDPMP),  # join on code only, no duplicate cols
    by = "MPIO_CDPMP"
  ) %>%
  select(
    municipio, depto,
    sndem_avg, sndem_2000, sndem_2023, sndem_change,
    Acueducto_avg, MortInfantil1_avg,
    NetCover_avg, AvgS11_avg, PobTot_12_avg
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  arrange(depto, municipio) %>%
  print(width = 200)


# 5. National median trajectory (needed for time-series plot)
nat_ribbon <- master_clean2 %>%
  group_by(year) %>%
  summarise(
    med = median(sndem_norm, na.rm = TRUE),
    q25 = quantile(sndem_norm, 0.25, na.rm = TRUE),
    q75 = quantile(sndem_norm, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# Quick sanity check
nat_ribbon %>%
  filter(year %in% c(2000, 2005, 2010, 2016, 2023)) %>%
  print()

```

## Zooming In: What the Colombia Data Tell Us

Zooming in, we can examine what the resulting estimates reveal about democratic 
variation across Colombian municipalities from 2000 to 2023. This section proceeds 
in three steps: an overview of the data as a whole, the 
identification and description of illustrative cases that capture the range 
of democratic trajectories in the data, and a brief discussion of two 
theoretical questions these data are positioned to address. The goal of 
this section is not causal inference but demonstration--showing what kind 
of descriptive and comparative leverage a thick, cross-sectional time-series 
measure of subnational democracy makes possible.

```{r overview, include=FALSE}
## --- DISTRIBUTIONAL OVERVIEW FIGURE ---

# National-level summary for ribbon
nat_ribbon <- master_clean2 %>%
  group_by(year) %>%
  summarise(
    med = median(sndem_norm, na.rm = TRUE),
    q25 = quantile(sndem_norm, 0.25, na.rm = TRUE),
    q75 = quantile(sndem_norm, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# Sub-index trends for annotation
subindex_nat <- master_clean2 %>%
  group_by(year) %>%
  summarise(
    elect_med = median(snelect_norm, na.rm = TRUE),
    civil_med = median(sncivlib_norm, na.rm = TRUE),
    .groups = "drop"
  )

# Pivot events
events <- tibble(
  year  = c(2005, 2016),
  label = c("AUC\nDemobilization", "FARC-EP\nAgreement")
)

p_dist <- ggplot() +
  geom_ribbon(
    data = nat_ribbon,
    aes(x = year, ymin = q25, ymax = q75),
    fill = "grey85", alpha = 0.6
  ) +
  geom_line(
    data = nat_ribbon,
    aes(x = year, y = med, linetype = "Composite (sndem)"),
    color = "grey20", linewidth = 0.9
  ) +
  geom_line(
    data = subindex_nat,
    aes(x = year, y = elect_med, linetype = "Electoral (snelect)"),
    color = "#0072B2", linewidth = 0.7
  ) +
  geom_line(
    data = subindex_nat,
    aes(x = year, y = civil_med, linetype = "Civil Liberties (sncivlib)"),
    color = "#D55E00", linewidth = 0.7
  ) +
  geom_vline(
    data = events,
    aes(xintercept = year),
    linetype = "dotted", color = "grey40", linewidth = 0.6
  ) +
  geom_text(
    data = events,
    aes(x = year - 0.3, y = 0.97, label = label),
    size = 2.8, hjust = 1, color = "grey30", lineheight = 0.85
  ) +
  scale_linetype_manual(
    name = NULL,
    values = c(
      "Composite (sndem)"          = "solid",
      "Electoral (snelect)"        = "dashed",
      "Civil Liberties (sncivlib)" = "dotdash"
    )
  ) +
  scale_x_continuous(breaks = seq(2000, 2023, 4)) +
  scale_y_continuous(limits = c(0.3, 1.0), breaks = seq(0.3, 1.0, 0.1)) +
  labs(
    x       = NULL,
    y       = "Normalized democracy score",
    caption = paste(
      "Solid line = national municipal median (composite index).",
      "Grey ribbon = interquartile range.",
      "Colored lines = sub-index medians.",
      sep = "\n"
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )

p_dist

```
### Distributional Overview

Figure X shows the national-median trajectory of the composite subnational 
democracy score (`sndem`) alongside its interquartile range across 
municipalities, with the two sub-index medians shown separately. Three 
patterns are immediately apparent. First, average municipal democracy 
declined modestly from 2000 to approximately 2010 — a period corresponding 
to the intensification of counterinsurgency operations under the Alvaro Uribe 
administration — before recovering strongly after 2016, with the median 
normalized score rising from 0.522 to 0.730 between 2010 and 2023. Second, 
this recovery is widespread: the interquartile range shifts upward across 
the full distribution, suggesting broad-based rather than concentrated 
improvement after the FARC-EP comprehensive peace agreement. Third, the electoral fairness 
and civil liberties sub-indices track one another closely at the national 
level, though we show below that they diverge more substantially at the 
municipal level, underscoring the value of measuring both dimensions 
separately.

```{r focalmpios, include=FALSE}
## --- CASE SELECTION: FOCAL MUNICIPALITIES ---

# Confirmed MPIO codes from focal_check:
# 05197 Cocorná     (Antioquia) - extreme improver
# 19809 Timbiquí    (Cauca)     - extreme decliner
# 50573 Puerto López (Meta)     - typical case
# 41349 Hobo        (Huila)     - most-similar pair A
# 41483 Nátaga      (Huila)     - most-similar pair B
# 05045 Apartadó    (Antioquia) - most-different pair A
# 73461 Murillo     (Tolima)    - most-different pair B

focal_codes <- c("05197","19809","50573",
                 "41349","41483","05045","73461")

focal_meta <- tibble(
  MPIO_CDPMP = focal_codes,
  case_label = c(
    "Extreme Improver\n(Cocorná)",
    "Extreme Decliner\n(Timbiquí)",
    "Typical Case\n(Puerto López)",
    "Most-Similar A\n(Hobo)",
    "Most-Similar B\n(Nátaga)",
    "Most-Different A\n(Apartadó)",
    "Most-Different B\n(Murillo)"
  ),
  case_group = c(
    "Extreme", "Extreme", "Typical",
    "Most-Similar", "Most-Similar",
    "Most-Different", "Most-Different"
  )
)

# Time-series data for focal municipalities (normalized)
focal_ts <- master_clean2 %>%
  filter(MPIO_CDPMP %in% focal_codes) %>%
  left_join(focal_meta, by = "MPIO_CDPMP")

## --- TIME-SERIES PLOT: FOCAL CASES ---

group_colors <- c(
  "Extreme"        = "#E69F00",
  "Typical"        = "#009E73",
  "Most-Similar"   = "#0072B2",
  "Most-Different" = "#CC79A7"
)

group_lines <- c(
  "05197" = "solid",   # Cocorná
  "19809" = "dashed",  # Timbiquí
  "50573" = "solid",   # Puerto López
  "41349" = "solid",   # Hobo
  "41483" = "dashed",  # Nátaga
  "05045" = "solid",   # Apartadó
  "73461" = "dashed"   # Murillo
)

p_cases <- ggplot() +
  # National IQR ribbon
  geom_ribbon(
    data = nat_ribbon,
    aes(x = year, ymin = q25, ymax = q75),
    fill = "grey90", alpha = 0.7
  ) +
  geom_line(
    data = nat_ribbon,
    aes(x = year, y = med),
    color = "grey50", linetype = "dashed", linewidth = 0.6
  ) +
  # Focal cases
  geom_line(
    data = focal_ts,
    aes(
      x        = year,
      y        = sndem_norm,
      color    = case_group,
      linetype = MPIO_CDPMP,
      group    = MPIO_CDPMP
    ),
    linewidth = 0.9
  ) +
  # Event markers
  geom_vline(
    data = events,
    aes(xintercept = year),
    linetype = "dotted", color = "grey30", linewidth = 0.5
  ) +
  geom_text(
    data = events,
    aes(x = year - 0.3, y = 0.13, label = label),
    size = 2.6, hjust = 1, color = "grey30", lineheight = 0.85
  ) +
  scale_color_manual(values = group_colors, name = "Case type") +
  scale_linetype_manual(
    values = group_lines,
    labels = c(
      "05197" = "Cocorná", "19809" = "Timbiquí",
      "50573" = "Puerto López",
      "41349" = "Hobo", "41483" = "Nátaga",
      "05045" = "Apartadó", "73461" = "Murillo"
    ),
    name = "Municipality"
  ) +
  scale_x_continuous(breaks = seq(2000, 2023, 4)) +
  labs(
    x       = NULL,
    y       = "Normalized democracy score",
    caption = paste(
      "Grey ribbon = national IQR; dashed grey = national median.",
      "Vertical dotted lines mark the 2005 AUC demobilization and 2016 FARC peace agreement.",
      sep = "\n"
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )

p_cases

```
### Case Selection

To illustrate the inferential possibilities opened by these data, we 
identify five municipalities--representing four types of democratic 
trajectory--that together capture the range of variation in the data. 
Figure Y plots their normalized `sndem` trajectories alongside the national 
median and interquartile range. Table X summarizes their covariate profiles.

**Extreme improver: Cocorná (Antioquia).** Cocorná registered the largest 
democratic improvement of any municipality in our data, with its normalized 
`sndem` score rising from 0.394 in 2000 to 0.910 in 2023 — a gain of 0.517 
points. Cocorná is located in the *Oriente Antioqueño*, a subregion that 
experienced some of the worst paramilitary and guerrilla violence in 
Colombia during the late 1990s and early 2000s, including large-scale 
forced displacement. Notably, four other Oriente Antioqueño municipalities — 
San Carlos, San Francisco, San Luis, and Argelia — appear among the top ten 
``improvers" in the data, suggesting that the index is capturing a coherent 
regional pattern of post-conflict democratic recovery rather than an 
idiosyncratic outlier. Importantly, this recovery is not simply a reflection 
of socioeconomic gains: Cocorná's average water access coverage stands at 
only 42%, and it remains a small rural municipality (average population 
~17,000). Its improvement in `sndem` is driven primarily by the electoral 
fairness component, consistent with accounts of normalized local electoral 
competition following the demobilization of armed actors in the region.

**Extreme decliner: Timbiquí (Cauca).** At the other end of the spectrum, 
Timbiquí--a predominantly Afro-Colombian municipality on Colombia's Pacific 
coast--experienced the largest democratic decline of any municipality with 
usable covariate data, falling from 0.608 in 2000 to 0.331 in 2023 (−0.278 
points). Its trajectory represents a pattern documented in the post-conflict 
literature: as FARC structures dismantled following the 2016 agreement, 
criminal organizations moved into Pacific coast territories with more 
predatory and less institutionally predictable forms of violence, producing 
deteriorating conditions for both civil liberties and electoral integrity. 
Timbiquí's decline accelerates precisely after 2016 — the opposite of the 
national trend — illustrating how aggregate post-conflict improvement can 
coexist with severe democratic deterioration in specific peripheral 
territories. We note that several frontier municipalities in Amazonia and 
the Llanos Orientales also show sharp declines but are excluded from 
covariate-based comparisons due to sparse DANE administrative coverage; 
their trajectories represent a data limitation that future iterations of the 
method should address.

**Typical case: Puerto López (Meta).** Puerto López most closely tracks the 
national median trajectory across the full 24-year period, with a mean 
squared deviation of 0.000424 from the annual national median — less than 
half that of the next-closest candidate. Located in the Llanos Orientales, 
neither a conflict epicenter nor a development showcase, it serves as a 
useful reference point for what an "average" Colombian municipality looks 
like on our index. Its score rises from 0.627 in 2000 to 0.748 in 2023, 
tracking the national post-2016 recovery closely.

**Most-similar municipalities, different outcomes: Hobo and Nátaga (Huila).** 
The most-similar design logic asks: can we identify pairs of municipalities 
that are structurally similar on observable dimensions yet exhibit 
meaningfully different democratic outcomes? Our algorithmic search surfaces 
Hobo and Nátaga, neighboring municipalities in the department of Huila, as 
the cleanest such pair in the data. Their covariate profiles are 
strikingly similar: water access coverage (71.8% vs. 70.6%), infant 
mortality (17.8 vs. 19.0 per 1,000), net school enrollment (92.4% vs. 
92.9%), average SABER 11 scores (48.1 vs. 48.2), and population (~7,000 
vs. ~6,000) are nearly identical across all five dimensions used in the 
matching procedure. Yet their average normalized `sndem` scores diverge 
by 0.240 points (0.644 vs. 0.404), a gap that cannot be attributed to 
differences in any of the standard socioeconomic suspects. Notably, this 
gap narrows over the study period: Hobo declines slightly while Nátaga 
improves, particularly after 2016, suggesting some convergence. The 
persistence of the gap despite matched covariates opens a productive 
research question: what accounts for Hobo's democratic advantage? Our data 
identify it but cannot by themselves resolve it--which is precisely the 
kind of leverage a thick subnational democracy measure provides for 
subsequent qualitative or quantitative investigation.

**Most-different municipalities, similar outcomes: Apartadó (Antioquia) 
and Murillo (Tolima).** The complementary most-different design asks: can municipalities 
arrive at similar democratic outcomes via entirely different structural 
pathways? Apartadó and Murillo have nearly identical average normalized 
`sndem` scores across the study period (0.595 vs. 0.600), yet their 
covariate profiles diverge sharply. Apartadó is a mid-sized city in the 
Urabá region (~109,000 average population) with relatively high water 
access (83.5%) and low infant mortality (12.3 per 1,000), historically 
defined by banana plantation labor politics and intense paramilitary 
violence. Murillo is a small, remote Andean páramo municipality in 
northern Tolima (~4,000 population) with poor water access (30.9%) and 
higher infant mortality (23.9 per 1,000). Their convergence on similar 
democratic scores despite these structural differences--and despite very 
different temporal trajectories, with Apartadó remaining relatively stable 
and Murillo improving markedly after 2016--illustrates that our index is 
not simply recovering socioeconomic or geographic gradients already 
observable in the data.

```{r covsprof, include=FALSE}
## --- COVARIATE PROFILE TABLE ---

library(knitr)

profile_table <- muni_avg %>%
  filter(MPIO_CDPMP %in% focal_codes) %>%
  left_join(
    focal_meta %>% select(MPIO_CDPMP, case_label),
    by = "MPIO_CDPMP"
  ) %>%
  select(
    case_label, municipio, depto,
    sndem_avg, sndem_2000, sndem_2023, sndem_change,
    Acueducto_avg, MortInfantil1_avg,
    NetCover_avg, AvgS11_avg, PobTot_12_avg
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  # Order by narrative appearance
  mutate(case_label = factor(case_label, levels = focal_meta$case_label)) %>%
  arrange(case_label)

kable(
  profile_table %>% select(-case_label),
  col.names = c(
    "Municipality", "Department",
    "Avg. sndem", "sndem (2000)", "sndem (2023)", "Change",
    "Water Access (%)", "Infant Mortality",
    "Net Enrollment (%)", "SABER Score", "Population"
  ),
  caption = paste(
    "Table X. Covariate profiles of focal municipalities.",
    "Democracy scores are normalized (0–1).",
    "Control variables are period averages.",
    "Water access = % households with piped water.",
    "Infant mortality = deaths per 1,000 live births.",
    "SABER Score = average ICFES SABER 11 composite.",
    sep = " "
  ),
  align = "llrrrrrrrrr"
)
```

```{r zoommaps, include=FALSE}
## --- ZOOM MAPS ---
# Load geometry (reuse from facet chunk)
clean_mpio <- function(x) str_pad(as.character(as.numeric(x)), 5, "left", "0")

muni_geo <- st_read("G:/Shared drives/snvdem/snvdem-col/data/geospatial/2018pmq/BaseLayer/MGN_ANM_MPIOS.shp") %>%
  mutate(MPIO_CDPMP = clean_mpio(MPIO_CDPMP))

snap_year <- 2023

map_snap <- muni_geo %>%
  left_join(
    master_clean2 %>%
      filter(year == snap_year) %>%
      select(MPIO_CDPMP, sndem_norm),
    by = "MPIO_CDPMP"
  )

# Centroids for focal labels
focal_geo <- muni_geo %>%
  filter(MPIO_CDPMP %in% focal_codes) %>%
  st_centroid() %>%
  left_join(focal_meta, by = "MPIO_CDPMP") %>%
  left_join(
    master_clean2 %>%
      filter(year == snap_year) %>%
      select(MPIO_CDPMP, sndem_norm),
    by = "MPIO_CDPMP"
  ) %>%
  mutate(
    label_text = paste0(
      MPIO_CNMBR, "\n",
      round(sndem_norm, 2)
    )
  )

# Neighbor countries for context
neighbors <- ne_countries(
  scale = "medium", returnclass = "sf"
) %>%
  filter(continent == "South America" | name == "Panama")

# --- Helper function (reusable) ---
zoom_map <- function(data, focal_pts, xlim, ylim, title) {
  ggplot(data = data) +
    geom_sf(data = neighbors, fill = "#f5f5f5",
            color = "grey80", linewidth = 0.2) +
    geom_sf(aes(fill = sndem_norm), color = NA) +
    geom_sf(
      data  = focal_pts,
      shape = 21, size = 3.5,
      fill  = "white", color = "black", stroke = 1.2
    ) +
    geom_label_repel(
      data = focal_pts,
      aes(label = label_text, geometry = geometry),
      stat          = "sf_coordinates",
      size          = 2.8,
      box.padding   = 0.7,
      segment.color = "grey30",
      min.segment.length = 0,
      fill          = alpha("white", 0.85),
      label.size    = 0.2
    ) +
    scale_fill_viridis_c(
      option  = "magma",
      name    = "sndem\n(2023)",
      limits  = c(0, 1),
      na.value = "grey75",
      guide   = guide_colorbar(
        barheight = unit(2.5, "cm"), ticks = FALSE
      )
    ) +
    coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
    labs(title = title) +
    theme_void(base_size = 10) +
    theme(
      plot.title = element_text(
        face = "bold", hjust = 0.5, size = 10,
        margin = margin(b = 4)
      ),
      legend.position = "right"
    )
}

# Zoom A: Oriente Antioqueño (Cocorná) + Urabá (Apartadó)
zoom_antioquia <- zoom_map(
  data      = map_snap,
  focal_pts = focal_geo %>%
    filter(MPIO_CDPMP %in% c("05197", "05045")),
  xlim  = c(-77.0, -73.5),
  ylim  = c(5.0, 8.5),
  title = "A: Antioquia"
)

# Zoom B: Pacific coast (Timbiquí) + Huila (Hobo, Nátaga)
zoom_southwest <- zoom_map(
  data      = map_snap,
  focal_pts = focal_geo %>%
    filter(MPIO_CDPMP %in% c("19809", "41349", "41483")),
  xlim  = c(-78.0, -74.0),
  ylim  = c(0.5, 4.5),
  title = "B: Pacific Coast & Huila"
)

# Zoom C: Llanos + Tolima (Puerto López, Murillo)
zoom_east <- zoom_map(
  data      = map_snap,
  focal_pts = focal_geo %>%
    filter(MPIO_CDPMP %in% c("50573", "73461")),
  xlim  = c(-76.5, -71.5),
  ylim  = c(2.0, 6.5),
  title = "C: Meta & Tolima"
)

# --- CONTEXT MAP WITH BOUNDING BOXES ---

# Define the three zoom windows as rectangles
# (must match the xlim/ylim values in your zoom_map() calls)
zoom_boxes <- tibble(
  panel  = c("A", "B", "C"),
  xmin   = c(-77.0, -78.0, -76.5),
  xmax   = c(-73.5, -74.0, -71.5),
  ymin   = c(  5.0,   0.5,   2.0),
  ymax   = c(  8.5,   4.5,   6.5)
)

# Label positions (top-left corner of each box, nudged inward)
zoom_labels <- zoom_boxes %>%
  mutate(
    label_x = xmin + 0.2,
    label_y = ymax - 0.3
  )

context_map <- ggplot() +
  # Ocean background
  geom_rect(
    aes(xmin = -82, xmax = -66, ymin = -5, ymax = 13),
    fill = "#d6eaf8"
  ) +
  # Neighbor countries
  geom_sf(
    data  = neighbors,
    fill  = "#f0f0f0",
    color = "grey70",
    linewidth = 0.2
  ) +
  # Colombia municipalities (full country, sndem fill)
  geom_sf(
    data  = map_snap,
    aes(fill = sndem_norm),
    color = NA
  ) +
  scale_fill_viridis_c(
    option   = "magma",
    limits   = c(0, 1),
    na.value = "grey75",
    guide    = "none"   # legend handled by zoom panels
  ) +
  # Zoom bounding boxes
  geom_rect(
    data = zoom_boxes,
    aes(xmin = xmin, xmax = xmax,
        ymin = ymin, ymax = ymax),
    fill  = NA,
    color = "white",
    linewidth = 1.0
  ) +
  # Outer stroke for contrast on dark backgrounds
  geom_rect(
    data = zoom_boxes,
    aes(xmin = xmin, xmax = xmax,
        ymin = ymin, ymax = ymax),
    fill  = NA,
    color = "grey20",
    linewidth = 0.4,
    linetype = "dashed"
  ) +
  # Panel labels inside boxes
  geom_label(
    data = zoom_labels,
    aes(x = label_x, y = label_y, label = panel),
    size      = 3.2,
    fontface  = "bold",
    fill      = alpha("white", 0.85),
    label.size = 0.2,
    hjust     = 0,
    vjust     = 1
  ) +
  coord_sf(
    xlim   = c(-82, -66),
    ylim   = c(-5, 13),
    expand = FALSE
  ) +
  labs(title = "Colombia") +
  theme_void(base_size = 9) +
  theme(
    plot.title = element_text(
      face = "bold", hjust = 0.5,
      size = 9, margin = margin(b = 3)
    ),
    panel.border = element_rect(
      color = "grey30", fill = NA, linewidth = 0.4
    )
  )

# --- ASSEMBLE FINAL FIGURE ---
# Layout: context map (left column) + three zoom panels (right column, stacked)
# Use patchwork area() for precise control

layout <- c(
  area(1, 1, 3, 2),  # context map spans all 3 rows, columns 1-2
  area(1, 3, 1, 4),  # Zoom A: row 1, columns 3-4
  area(2, 3, 2, 4),  # Zoom B: row 2, columns 3-4
  area(3, 3, 3, 4)   # Zoom C: row 3, columns 3-4
)

final_zoom <- context_map +
  zoom_antioquia +
  zoom_southwest +
  zoom_east +
  plot_layout(design = layout, guides = "collect") +
  plot_annotation(
    title    = "Figure Z: Zoomed Views of Subnational Democracy in Colombia (2023)",
    subtitle = "Labels show municipality name and normalized sndem score. Boxes indicate zoomed regions.",
    caption  = "Source: SNDEM 2024.",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 13, hjust = 0.5),
      plot.subtitle = element_text(size = 10, hjust = 0.5, color = "grey30"),
      plot.caption  = element_text(hjust = 1, color = "grey40", size = 8)
    )
  ) &
  theme(legend.position = "right")

# Export
ggsave(
  "G:/Shared drives/snvdem/snvdem-col/data/panel/09_analysis_scripts/ZoomIn/SNDEM_zoom_with_context.png",
  final_zoom,
  width  = 14,
  height = 9,
  dpi    = 300,
  bg     = "white"
)

final_zoom

```


### Two Theoretical Questions

Our data are not designed to resolve causal questions in this paper; that 
is the work the method enables for future research. We nonetheless sketch 
two questions for which these data are well-positioned.

**The relationship between national and subnational democracy.** Figure X 
shows that the municipal distribution broadly tracks the national V-Dem 
Electoral Democracy Index over time, but with important deviations. The 
national median of municipal `sndem` dips from 0.579 (2000) to 0.522 (2010) 
before recovering sharply to 0.730 (2023) — a trajectory that parallels 
Colombia's national democratic arc but is not identical to it. The 2005 AUC 
demobilization produces little visible movement in the municipal distribution, 
while the 2016 FARC agreement is associated with a pronounced and sustained 
upward shift. Critically, these aggregate trends obscure divergent municipal 
trajectories: Cocorná improves dramatically while Timbiquí deteriorates 
sharply in the same post-2016 period. Whether subnational democratization 
leads, follows, or selectively diverges from national trends — and under 
what conditions — is an empirical question that our time-series data allow 
scholars to address systematically, and cross-nationally, once the method 
is extended beyond Colombia.

**Replication versus autonomous subnational environments.** If municipal 
democracy were simply a downward projection of national institutions, we 
would expect low cross-municipal variance and tight correlation with 
national-level indicators. The evidence presented here suggests otherwise. 
Substantial variation persists across municipalities even within years of 
stable national-level democracy, the most-similar pair (Hobo and Nátaga) 
demonstrates that virtually identical socioeconomic conditions are 
compatible with meaningfully different democratic outcomes, and the 
most-different pair (Apartadó and Murillo) shows that structurally 
dissimilar municipalities can converge on similar democratic scores via 
entirely different pathways. Taken together, these patterns point toward 
subnational environments as partially autonomous democratic arenas — a 
finding with implications for theories of democratic resilience, 
autocratization, and the territorial reach of the state.



