---
title: "LOO_Validation"
format: html
editor: visual
---

# Validation of individual characters

```{r}
library(tidyverse)


rm(list = ls())

Indices_both <- readRDS("G:/Shared drives/snvdem/snvdem-col/data/panel/08_final_snvdem_data/MC/SNDEM_tentative.rds")


chars <- c("rural_urban", "econ_dev", "capital", "north", "south", 
           "west", "east", "violence", "sparse", "remote", 
           "indigenous", "ruling_party")

char_labels <- c(
  "rural_urban"  = "Rural/Urban",
  "econ_dev"     = "Economic Development",
  "capital"      = "Distance from Capital",
  "north"        = "North",
  "south"        = "South",
  "west"         = "West",
  "east"         = "East",
  "violence"     = "Violence Index",
  "sparse"       = "Sparse Population",
  "remote"       = "Remoteness",
  "indigenous"   = "Indigenous Population",
  "ruling_party" = "Ruling Party Strength"
)



```

### Function to do the calculations programatically

```{r}
calc_loco <- function(df, excl) {
  # excl is the character to exclude
  df %>% mutate(
    
    # Each characteristic maps to specific num/den terms.
    
    # ELECTORAL NUMERATOR -- full 16 terms, then subtract excluded
    el_num_x = 
      (if (excl != "rural_urban")  (avg0t1hi * el_Urban) + (avg0t1lo * el_Rural)                else 0) +
      (if (excl != "econ_dev")     (avg2t3hi * el_More_development) + (avg2t3lo * el_Less_development) else 0) +
      (if (excl != "capital")      (avg4t5hi * (1 - el_Inside_capital)) + (avg4t5lo * (1 - el_Outside_capital)) else 0) +
      (if (excl != "north")        (avg6 * el_North)         else 0) +
      (if (excl != "south")        (avg7 * el_South)         else 0) +
      (if (excl != "west")         (avg8 * el_West)          else 0) +
      (if (excl != "east")         (avg9 * el_East)          else 0) +
      (if (excl != "violence")     (avg10t11 * wt_el_1011)   else 0) +
      (if (excl != "sparse")       (avg12 * el_Sparse_population) else 0) +
      (if (excl != "remote")       (avg13 * el_Remote)       else 0) +
      (if (excl != "indigenous")   (avg14 * (1 - el_Indigenous)) else 0) +
      (if (excl != "ruling_party") (avg15t16hi * (1 - el_Ruling_party_strong)) + (avg15t16lo * (1 - el_Ruling_party_weak)) else 0),
    
    # ELECTORAL DENOMINATOR
    el_den_x =
      (if (excl != "rural_urban")  el_Urban + el_Rural                               else 0) +
      (if (excl != "econ_dev")     el_More_development + el_Less_development           else 0) +
      (if (excl != "capital")      (1 - el_Inside_capital) + (1 - el_Outside_capital) else 0) +
      (if (excl != "north")        el_North         else 0) +
      (if (excl != "south")        el_South         else 0) +
      (if (excl != "west")         el_West          else 0) +
      (if (excl != "east")         el_East          else 0) +
      (if (excl != "violence")     wt_el_1011       else 0) +
      (if (excl != "sparse")       el_Sparse_population else 0) +
      (if (excl != "remote")       el_Remote        else 0) +
      (if (excl != "indigenous")   (1 - el_Indigenous)  else 0) +
      (if (excl != "ruling_party") (1 - el_Ruling_party_strong) + (1 - el_Ruling_party_weak) else 0),
    
    # CIVIL LIBERTIES NUMERATOR
    cl_num_x =
      (if (excl != "rural_urban")  (avg0t1hi * cl_Urban) + (avg0t1lo * cl_Rural)     else 0) +
      (if (excl != "econ_dev")     (avg2t3hi * cl_More_development) + (avg2t3lo * cl_Less_development) else 0) +
      (if (excl != "capital")      (avg4t5hi * (1 - cl_Inside_capital)) + (avg4t5lo * (1 - cl_Outside_capital)) else 0) +
      (if (excl != "north")        (avg6 * cl_North)         else 0) +
      (if (excl != "south")        (avg7 * cl_South)         else 0) +
      (if (excl != "west")         (avg8 * cl_West)          else 0) +
      (if (excl != "east")         (avg9 * cl_East)          else 0) +
      (if (excl != "violence")     (avg10t11 * wt_cl_1011)   else 0) +
      (if (excl != "sparse")       (avg12 * cl_Sparse_population) else 0) +
      (if (excl != "remote")       (avg13 * cl_Remote)       else 0) +
      (if (excl != "indigenous")   (avg14 * (1 - cl_Indigenous)) else 0) +
      (if (excl != "ruling_party") (avg15t16hi * 
                                      (1 - cl_Ruling_party_strong)) + (avg15t16lo *
                                      (1 - cl_Ruling_party_weak)) else 0),
    
    # CIVIL LIBERTIES DENOMINATOR
    cl_den_x =
      (if (excl != "rural_urban")  cl_Urban + cl_Rural    else 0) +
      (if (excl != "econ_dev")     cl_More_development + cl_Less_development           else 0) +
      (if (excl != "capital")      (1 - cl_Inside_capital) + (1 - cl_Outside_capital) else 0) +
      (if (excl != "north")        cl_North         else 0) +
      (if (excl != "south")        cl_South         else 0) +
      (if (excl != "west")         cl_West          else 0) +
      (if (excl != "east")         cl_East          else 0) +
      (if (excl != "violence")     wt_cl_1011       else 0) +
      (if (excl != "sparse")       cl_Sparse_population else 0) +
      (if (excl != "remote")       cl_Remote        else 0) +
      (if (excl != "indigenous")   (1 - cl_Indigenous)  else 0) +
      (if (excl != "ruling_party") (1 - cl_Ruling_party_strong) + (1 - cl_Ruling_party_weak) else 0),
    
    snelect_x  = el_num_x / el_den_x,
    sncivlib_x = cl_num_x / cl_den_x,
    sndem_x    = 0.5 * (snelect_x + sncivlib_x)
    
  ) %>%
    select(-el_num_x, -el_den_x, -cl_num_x, -cl_den_x)
}

```

## Leave One Out

```{r}
# Currently tosses the missing years
Indices_both_complete <- Indices_both %>% filter(!is.na(sndem))

loco_results <- map_dfr(chars, function(ch) {
  
  excl_df <- calc_loco(Indices_both_complete, excl = ch)
  
  tibble(
    characteristic   = ch,
    label            = char_labels[ch],
    
    # Correlation of excluded index vs full index
    cor_snelect  = cor(excl_df$snelect_x,  Indices_both_complete$snelect,  use = "complete.obs"),
    cor_sncivlib = cor(excl_df$sncivlib_x, Indices_both_complete$sncivlib, use = "complete.obs"),
    cor_sndem    = cor(excl_df$sndem_x,     Indices_both_complete$sndem,    use = "complete.obs"),
    
    # Mean absolute difference from full index
    mad_sndem    = mean(abs(excl_df$sndem_x - Indices_both_complete$sndem), na.rm = TRUE),
    
    # SD of difference (spread of impact)
    sd_diff_sndem = sd(excl_df$sndem_x - Indices_both_complete$sndem, na.rm = TRUE)
  )
})

print(loco_results %>% arrange(cor_sndem))
```

## Variance decomposition 

Compute each *weighted* predictor contribution and correlate it with the final index. This answers the question about distance/remoteness.

```{r}
vd_df <- Indices_both_complete %>%
  mutate(
    # Weighted electoral contributions (numerator terms)
    wc_el_rural_urban  = (avg0t1hi * el_Urban)  + (avg0t1lo * el_Rural),
    wc_el_econ_dev     = (avg2t3hi * el_More_development) + (avg2t3lo * el_Less_development),
    wc_el_capital      = (avg4t5hi * (1 - el_Inside_capital)) + (avg4t5lo * (1 - el_Outside_capital)),
    wc_el_north        = avg6 * el_North,
    wc_el_south        = avg7 * el_South,
    wc_el_west         = avg8 * el_West,
    wc_el_east         = avg9 * el_East,
    wc_el_violence     = avg10t11 * wt_el_1011,
    wc_el_sparse       = avg12 * el_Sparse_population,
    wc_el_remote       = avg13 * el_Remote,
    wc_el_indigenous   = avg14 * (1 - el_Indigenous),
    wc_el_ruling_party = (avg15t16hi * (1 - el_Ruling_party_strong)) + (avg15t16lo * (1 - el_Ruling_party_weak)),
    
    # Weighted civil liberties contributions
    wc_cl_rural_urban  = (avg0t1hi * cl_Urban)  + (avg0t1lo * cl_Rural),
    wc_cl_econ_dev     = (avg2t3hi * cl_More_development) + (avg2t3lo * cl_Less_development),
    wc_cl_capital      = (avg4t5hi * (1 - cl_Inside_capital)) + (avg4t5lo * (1 - cl_Outside_capital)),
    wc_cl_north        = avg6 * cl_North,
    wc_cl_south        = avg7 * cl_South,
    wc_cl_west         = avg8 * cl_West,
    wc_cl_east         = avg9 * cl_East,
    wc_cl_violence     = avg10t11 * wt_cl_1011,
    wc_cl_sparse       = avg12 * cl_Sparse_population,
    wc_cl_remote       = avg13 * cl_Remote,
    wc_cl_indigenous   = avg14 * (1 - cl_Indigenous),
    wc_cl_ruling_party       = (avg15t16hi * (1 - cl_Ruling_party_strong)) + (avg15t16lo * (1 - cl_Ruling_party_weak))
  )

# Correlate each weighted contribution with snelect and sncivlib
vd_results <- tibble(
  characteristic = chars,
  label          = char_labels[chars],
  cor_with_snelect  = map_dbl(chars, ~cor(vd_df[[paste0("wc_el_", .x)]], vd_df$snelect,  use = "complete.obs")),
  cor_with_sncivlib = map_dbl(chars, ~cor(vd_df[[paste0("wc_cl_", .x)]], vd_df$sncivlib, use = "complete.obs")),
  cor_with_sndem    = map_dbl(chars, ~{
    avg_wc <- (vd_df[[paste0("wc_el_", .x)]] + vd_df[[paste0("wc_cl_", .x)]]) / 2
    cor(avg_wc, vd_df$sndem, use = "complete.obs")
  })
) %>%
  arrange(desc(abs(cor_with_sndem)))

print(vd_results)

```

## Visuals

### LOO Sensitivity

```{r}

plot_loco <- loco_results %>%
  arrange(cor_sndem) %>%
  mutate(label = factor(label, levels = label)) %>%
  ggplot(aes(x = cor_sndem, y = label)) +
  geom_col(fill = "#2c7bb6", width = 0.6) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray40") +
  geom_text(aes(label = round(cor_sndem, 3)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(limits = c(0, 1.05), expand = c(0, 0)) +
  labs(
    title    = "Leave-One-Out Sensitivity",
    # subtitle = "Values near 1.0 indicate the characteristic does not drive the overall index",
    x        = "Pearson correlation with Full SNDEM Index",
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())

print(plot_loco)
# ggsave("loco_sensitivity.png", plot_loco, width = 8, height = 5, bg = "white")


```

### Variance decomposition 

which characteristics explain the most?

```{r}


vd_long <- vd_results %>%
  select(label, cor_with_snelect, cor_with_sncivlib) %>%
  pivot_longer(cols = c(cor_with_snelect, cor_with_sncivlib),
               names_to  = "index",
               values_to = "correlation") %>%
  mutate(
    index = recode(index,
                   "cor_with_snelect"  = "Electoral Fairness",
                   "cor_with_sncivlib" = "Civil Liberties"),
    label = factor(label, levels = vd_results$label)  # sorted by sndem cor
  )

plot_vd <- ggplot(vd_long, aes(x = correlation, y = label, fill = index)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_vline(xintercept = 0, color = "gray30") +
  scale_fill_manual(values = c("Electoral Fairness" = "darkblue", 
                               "Civil Liberties"    = "orange")) +
  labs(
    title    = "Weighted Characteristic Contributions to Index Components",
    # subtitle = "Correlation of each weighted predictor term with its index",
    x        = "Pearson correlation",
    y        = NULL,
    fill     = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        legend.position = "bottom")

print(plot_vd)
# ggsave("variance_decomposition.png", plot_vd, width = 8, height = 5.5, bg = "white")




```

## Summary Table

```{r}
final_table <- loco_results %>%
  select(label, cor_snelect, cor_sncivlib, cor_sndem, mad_sndem) %>%
  left_join(vd_results %>% select(label, cor_with_sndem), by = "label") %>%
  arrange(cor_sndem) %>%
  rename(
    "Characteristic"                          = label,
    "LOO: Electoral Fairness"                 = cor_snelect,
    "LOO: Civil Liberties"                    = cor_sncivlib,
    "LOO: Overall SNVDEM"                     = cor_sndem,
    "Mean Abs. Deviation from Full Index"     = mad_sndem,
    "Weighted Contribution to Overall Index"  = cor_with_sndem
  )

final_table %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  print(n = Inf)
write_csv(final_table, "LOO_Table.csv")
```

```{r}
# Combine LOO and VD results
plot_df <- loco_results %>%
  select(label, cor_snelect, cor_sncivlib) %>%
  left_join(
    vd_results %>% select(label, cor_with_snelect, cor_with_sncivlib),
    by = "label"
  )

# Shared plotting function to avoid repeating code
make_comparison_plot <- function(df, loo_col, vd_col, title, xlab) {
  
  char_order <- df %>% arrange(.data[[vd_col]]) %>% pull(label)
  
  df %>%
    select(label, all_of(c(loo_col, vd_col))) %>%
    pivot_longer(-label, names_to = "type", values_to = "correlation") %>%
    mutate(
      type  = recode(type,
                     !!loo_col := "LOO: excluded from model",
                     !!vd_col  := "Weighted contribution"),
      label = factor(label, levels = char_order)
    ) %>%
    ggplot(aes(x = correlation, y = label, fill = type)) +
    geom_vline(xintercept = 0, color = "gray70", linewidth = 0.4) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6) +
    scale_fill_manual(
      values = c("LOO: excluded from model" = "#d7191c",
                 "Weighted contribution"     = "#2c7bb6")
    ) +
    scale_x_continuous(limits = c(-0.1, 1.05), breaks = seq(0, 1, 0.25)) +
    labs(
      title    = title,
      # subtitle = paste0(
      #   "Blue: how much each characteristic's weighted contribution tracks the full index\n",
      #   "       (higher = explains more variation in final scores)\n",
      #   "Red:  correlation with full index when characteristic is removed\n",
      #   "       (higher = index is robust to its exclusion)"
      # ),
      x    = xlab,
      y    = NULL,
      fill = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      legend.position    = "bottom"
    )
}

# Electoral Fairness
plot_el <- make_comparison_plot(
  plot_df,
  loo_col = "cor_snelect",
  vd_col  = "cor_with_snelect",
  title   = "Electoral Fairness",
  xlab    = "Pearson correlation with full electoral fairness index"
)

# Civil Liberties
plot_cl <- make_comparison_plot(
  plot_df,
  loo_col = "cor_sncivlib",
  vd_col  = "cor_with_sncivlib",
  title   = "Civil Liberties",
  xlab    = "Pearson correlation with full civil liberties index"
)

print(plot_el)
print(plot_cl)

# ggsave("fig_electoral_loo_vd.png", plot_el, width = 8, height = 5.5, bg = "white")
# ggsave("fig_civlib_loo_vd.png",    plot_cl, width = 8, height = 5.5, bg = "white")
```
