---
title: "Visualizing subnational democracy using Varieties of Democracy (V-Dem)"
subtitle: "Merging V-Dem responses with municipal-level data"
author: "Patrick McQuestion^[Department of Political Science & Kroc Institute, University of Notre Dame. Email: <pmcquest@nd.edu>], Michael Coppedge^[Department of Political Science, University of Notre Dame. Email: <mcoppedg@nd.edu>], Matthew Sisk^[Department of Applied and Computational Mathematics and Statistics, University of Notre Dame. Email: <msisk1@nd.edu>]"
date: "First version: 04 February 2024. This version: `r format(Sys.Date(), '%d %B %Y')`."
editor_options: 
  markdown: 
    wrap: 72
bibliography: References/BiblioRA24.bib
csl: "G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/References/ajps.csl"
link-citations: yes
linkcolor: black
output: 
  pdf_document:
    latex_engine: xelatex
header-includes:
   - \usepackage{dcolumn}
   - \usepackage{setspace}
   - \doublespacing
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, 
                      message=FALSE, 
                      fig.path='Figs/',
                      fig.width=12, fig.height=8) 
setwd("G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq")
```

# Abstract

The objective of this methodological paper is to introduce geo-spatial
mapping of democracy at the subnational level using Varieties of
Democracy (V-Dem) data. This effort in democracy studies is data-driven
and descriptive [@coppedgeDemocratizationResearchMethods2012, 33-36]. We
describe the measurement process applied to one country (Colombia) in
order to stimulate discussion on the methodological potential for V-Dem
data to be used to measure democracy within countries. This paper
introduces a process for compiling and visualizing subnational data
based on V-Dem, starting with the systematic conceptualization of the
background concepts contained in the survey, assigning measures to those
concepts on a geographic scale, and then identifying subnational scores.
We demonstrate this process for a portion of the 21 response items in
the survey.

# Introduction

There is a need for subnational data on many political phenomena --
violence, democracy, service provision, human rights, gender relations,
discrimination, and a range of other topics. The need is especially
great in large countries, which are likely to contain considerable
geographic variation on these dimensions within their borders. To the
extent that such variation exists, claims about the "typical" levels of
these dimensions for whole countries are less useful, either as
descriptions or as independent or dependent variables. In this paper, we
demonstrate a technique for combining several less-often-used V-Dem
variables with geocoded demographic and political data to measure levels
of democracy in Colombia at the municipal level.

It is commonly assumed that it would be ideal to have data based on
direct observation in well-defined subnational units such as states or
provinces, counties or municipalities, or even smaller districts.
However, there are some drawbacks to that data collection strategy. In
the first place, hard information about subnational units, especially
very local ones, is often scarce. Second, even if the information
exists, it can be difficult or expensive to obtain, and may be collected
in unstandardized ways. Third, such "objective" data -- counts, texts,
etc. -- can be invalid when it captures only a small part of the concept
one needs to measure. It may be feasible to count the number of
newspapers and broadcasters in a locality, but a valid measure of
freedom of information, for example, should also consider the size of
their audiences, their ability to carry out independent investigations,
or their editorial autonomy. Fourth, whether a directly observable
measure matters is often conditional on other factors. The lack of
arrests of journalists in North Korea is not evidence of press freedom;
governing parties sometimes win elections for decades even when
elections are very competitive; texts of party platforms and speeches
are often insincere and inaccurate; and texts of constitutions and laws
can be "parchment" rules, dead letters that are not implemented in
practice.

For these reasons, there are sometimes advantages to following an
alternative measurement strategy: relying on the subjective judgments of
multiple country experts. This approach can compensate for some of the
weaknesses of "objective" data [@bernhardMakingEmbeddedKnowledge2017].
Experts absorb information from multiple sources -- reading, following
news, conversations, personal experiences, and deep familiarity with the
context. This enables them to reconcile diverse and often conflicting
kinds of information and synthesize it into an internally consistent
account of how things work. Their understanding extends to a sense of
the past and the direction of trends. Because they possess this depth of
understanding, experts are less likely to be deceived by insincere
statements or *de jure* arrangements that are ignored in practice and
less likely to consider one attribute definitive when they know that
other conditions render it irrelevant.

Of course, there is also an obvious disadvantage to subjective
knowledge: specifically, different experts tend to synthesize somewhat
different understandings. They arrive at different interpretations of
the same reality. This is why it is crucial to consult multiple experts,
which makes it possible to have greater confidence in the
interpretations on which experts agree, and to exercise prudent caution
about matters on which there is little agreement. Post hoc validation is
also essential. There are no guarantees that subjective ratings will
yield useful information, but it is worth trying as long as we also
subject the outcomes to multiple forms of validation.

## Measuring subnational democracy with V-Dem

Subnational democracy is an area where subjective expert ratings have
the potential to improve measurement. There are hundreds of democracy
measures, but almost all of them are at the national level of analysis
[@bernhardEconomicPerformanceInstitutional2001;
@bertelsmannstiftungBertelsmannTransformationIndex2005;
@boixCompleteDataSet2013; @coppedgeVDemDatasetV132023;
@economistintelligenceuniteiuDemocracyIndex20102010;
@freedomhouseMethodology2015; @buhlmannQualityDemocracyDemocracy2008;
@pemsteinDemocraticCompromiseLatent2010; @marshallPolityIVProject2014;
@przeworskiModernizationTheoriesFacts1997;
@skaaningLexicalIndexElectoral2015; @vanhanenNewDatasetMeasuring2000].
The few exceptions [@mickeyPathsOutDixie2015;
@mcmannDemocraticConsolidationTheory2018;
@gervasoniRentierTheorySubnational2010;
@giraudyVarietiesSubnationalUndemocratic2013] cover only one country
(although Giraudy does two). @mcmannExplainingSubnationalRegime2021 use
some V-Dem data to measure "unevenness" of elections and civil liberties
in most countries, but theirs is a national-level concept that tells us
nothing about where democracy is better or worse within a state.

However, V-Dem data contains several variables that provide a partial
path to measurement at the subnational level. There are four questions
on elections and civil liberties: Subnational election area less free
and fair characteristics (`v2elsnlfc`), Subnational election area more
free and fair characteristics (`v2elsnmrfc`), Stronger civil liberties
characteristics (`v2clrgstch`), and Weaker civil liberties
characteristics (`v2clrgwkch`). Although elections and civil liberties
are far from capturing every aspect of democracy, they cover the two
most central criteria for polyarchy
[@dahlPolyarchyParticipationOpposition1971].

```{r vdem-q, echo=FALSE, message=FALSE, results='asis'}
library(flextable)
library(dplyr)
library(officer)
library(kableExtra)
library(knitr)

vdem_q <- data.frame(
  Question = c("Question: How would you describe the areas of the country in which elections are significantly less free and fair? [v2elsnlfc]",
               "Question: How would you describe the areas of the country in which elections are significantly more free and fair? [v2elsnmrfc]",
               "Question: How would you describe the areas of the country where government officials’ respect for civil liberties is significantly stronger? [v2clrgstch]", 
               "Question: How would you describe the areas of the country where government officials’ respect for civil liberties is significantly weaker? [v2clrgwkch]",
               "Clarification: Choose all that apply.",
               "Responses:",
               "0: Rural. (0=No, 1=Yes)",
               "1: Urban. (0=No, 1=Yes)",
               "2: Areas that are less economically developed. (0=No, 1=Yes)",
               "3: Areas that are more economically developed. (0=No, 1=Yes)", 
               "4: Inside the capital city. (0=No, 1=Yes)",
               "5: Outside the capital city. (0=No, 1=Yes)",
               "6: North. (0=No, 1=Yes)",
               "7: South. (0=No, 1=Yes)", 
               "8: West. (0=No, 1=Yes)",
               "9: East. (0=No, 1=Yes)",
               "10: Areas of civil unrest (including areas where insurgent groups are active). (0=No, 1=Yes)",
               "11: Areas where illicit activity is widespread. (0=No, 1=Yes)", 
               "12: Areas that are very sparsely populated. (0=No, 1=Yes)",
               "13: Areas that are remote (difficult to reach by available transportation, for example). (0=No, 1=Yes)", 
               "14: Areas where there are indigenous populations. (0=No, 1=Yes)", 
               "15: Areas where the national ruling party or group is strong. (0=No, 1=Yes)", 
               "16: Areas where the national ruling party or group is weak. (0=No, 1=Yes)",
               "17: Areas that were subject to a longer period of foreign rule. (0=No, 1=Yes)",
               "18: Areas that were subject to a shorter period of foreign rule. (0=No, 1=Yes)",
               "19: Areas that were recently subject to foreign rule. (0=No, 1=Yes)",
               "20: Areas that have not recently been subject to foreign rule. (0=No, 1=Yes)",
               "21: None of the above. (0=No, 1=Yes)"
               ))

# Print the table with lighter lines
vdem_q %>%
  kable(format = "latex", escape = FALSE, caption = "V-Dem questions and clarifications", col.names = NULL) %>%
  kable_styling(full_width = FALSE, latex_options = c("hold_position", "scale_down")) %>%
  column_spec(1:ncol(vdem_q), width = "45em")
```

Table 1 reproduces the four questions, along with the same 21 "choose
all that apply" responses that apply to all questions. In the
country-year V-Dem data sets, each response becomes a separate variable
that is no longer binary, but the proportion of experts who chose the
"yes" response to that option. Thus, in actuality we have 84 (21\*4)
variables to work with.

These unusual options have advantages and disadvantages. On the positive
side, these options ask experts to identify the criteria that
distinguish locations where democracy is better from locations where it
is worse. The options are described in a generic way so that they could
be relevant in any country that V-Dem rates, anywhere in the world, and
over more than two centuries of history. In this paper we deal with just
12 of them as an initial test of the plausibility of the technique, but
we intend to incorporate the others, except for the options referring to
foreign rule (\# 17-20), which is not very relevant for Colombia, and
option #21 which refers to a negative response ("None of the above").
Still, even these few response options are varied enough to cast a wide
net that captures diverse criteria for distinguishing where democracy is
better or worse.

On the negative side, first, all of these options are conceptually
binary. In effect, they ask "Is this a useful discriminating criterion
or not?" rather than seeking information about degrees of relevance.
Second, the options do not name a specific concept or indicator that
should be used in empirical analysis. This decision is left up to each
country expert and V-Dem does not reveal, and in fact does not know, on
what information country experts based their assessments. Third, the
options do not specify a threshold that should be used to decide whether
a location does or does not satisfy the criterion. Fourth, many of
criteria overlap and are therefore somewhat interchangeable. This is
especially likely for the binaries of rural-urban, low-high income,
sparsely-densely populated and remote areas, which may also overlap with
areas of illicit activity, civil unrest, and indigenous population.
Finally, the criteria are not mapped onto well-defined locations such as
states, counties, or cities that would have known administrative
boundaries.

## Procedure for rendering subnational V-Dem data

Given these features, additional work is required to render the V-Dem
data useful for measuring democracy at the subnational level. We follow
this procedure: First, we consider the degree of agreement among experts
as a graded measure of the degree to which each criterion is likely to
be useful for discriminating between better and worse locations. We
believe that it is also reasonable to interpret these proportions as, in
part, reflections of degrees of fair elections or civil liberties
because the more democratic a subnational unit is, the more likely
experts are to notice it. In effect, we assume that experts perceive
degrees of democracy, but also have their own subjective thresholds on
that continuum that they use to distinguish more democratic situations
from less democratic ones when they are asked to make a binary
classification. Experts' thresholds differ, but the more democratic a
unit is on the underlying continuous dimension, the larger the
proportion of experts that will judge it to be in the "more democratic"
category. This psychological model is the basis for the Bayesian Item
Response Thoery (IRT) measurement model that V-Dem uses to aggregate
multiple coders' ratings into point measures for country-years
[@marquardtIRTModelsExpertCoded2018;
@pemsteinMeasurementModelLatent2023]. Although it would be ideal to
utilize a comparable model for this project, we lack sufficient data to
attain the statistical power equivalent to that available in V-Dem.

Next, for each criterion, we choose a specific indicator -- from
official state data or others' scholarly research -- to measure it. For
each indicator, we either choose a threshold to represent the binary
concept of whether the criterion applies to a location or not; or we
redefine the concept to be continuous so that it can measure the degree
to which each location satisfies the criterion. Some indicators work
well as matters of degree: income, population density, strength of the
ruling party, etc. We prefer the continuous criteria where possible.
Other criteria are probably best left binary, such as the categories of
North, South, East, and West -- although even these could be measured as
fuzzy sets.

We then multiply the two kinds of data (the degree to which experts
agree that a criterion is relevant, and the degree to which the
criterion applies to each location) to construct summary measures of
more, or less, free and fair elections and stronger or weaker
protections for civil liberties for each subnational unit. Getting to
this stage makes it possible to map these summary measures as
choropleths.

Finally, we combine these summary measures into one or more indices of
democracy (or aspects of it). At this stage, overlapping criteria could
be an advantage if they are treated as substitutable, and therefore
aggregated with an additive or weighted average formula
[@goertzSocialScienceConcepts2006]. (See below for more on this.)

We now justify our choices of indicators for the discriminating criteria
and explain why we treat each one as binary or continuous, and how we
transform them. We measure each criterion (whether binary or continuous)
on a 0-1 scale so that they get equal weight with expert agreement, and
so that products for each criterion will be on a 0-1 scale for later
aggregation with other criteria.

# Mapping exercise: 2018 cross-section of Colombia

This section describes in more detail a pipeline for mapping subnational
democracy, and shows how conceptualization, measurement, and scoring
occurs in practice. This exercise is designed to test the viability of
mapping cross-sectional country-level data before expanding the dataset
into panel format. Therefore, internal data from V-Dem is subset to
Colombia for 2018, and the last five subnational responses related to
foreign rule and "None of the Above" are removed because they are not
relevant for these purposes. Table 2 shows the proportional scores for
each variable, subset to Colombia in 2018, across the four survey
questions in V-Dem.

In order to map these V-Dem scores, the procedure outlined in this paper
uses simple definitions and systematizes them by selecting one or two
principal indicators for each response. Rather than discount alternative
systematizations, our objective is to visualize these indicators in
order to characterize subnational democracy in a given context, and
therefore we prioritize reliable over valid measurements
[@adcockMeasurementValidityShared2001]. This mapping procedure involves
multiple stages, from the systematization of concepts with explicit
definitions, to the assignment of measures based on coder responses, and
finally the identification of subnational "scores" based on these
measures.

```{r, Vdem0, include=FALSE, echo=FALSE}
# make a clean slate
rm(list = ls())
# load needed packages
library(tidyverse)
library(vdemdata)
library(ggplot2)
library(gridExtra)

# Create a data frame of version 13 dataset:
v13 <- vdem
# this df appears to be the same as MC's "full + others" data set: read_dta("C:/Users/mcoppedg/Dropbox/MC/VDemFiles/Archive/V13/V-Dem-CY-Full+Others/V-Dem-CY-Full+Others-v13.dta")

# Filter to just Colombia since 1899:
v13_col <- filter(v13, country_name=="Colombia" & year > 1899)

# Select just the ID and subnational variables
# (Note: v2elsnless and v2elsnmore are not in this df. We will make sure to 
# include them in our request to the data manager.)
v13_col_sn <- v13_col %>%
  select(country_id, country_name, country_text_id, historical_date, year, 
         v2elsnlfc_0:v2elsnlfc_21, v2elsnmrfc_0:v2elsnmrfc_21,
         v2clrgstch_0:v2clrgstch_21, v2clrgwkch_0:v2clrgwkch_21)

# Listing the variable names to check our work:
#names(v13_col_sn)
# Export the result to a CSV file:
#write.csv(v13_col_sn, file = "v13_col_sn.csv")
```

```{r, v13_col_2018, include=FALSE, echo=FALSE}
# create subset of V-Dem data for 2018 
v13_col_sn_2018 <- subset(v13_col_sn, year == 2018) 
v13_col_sn_2018 <- v13_col_sn_2018 %>%
  select(-matches("_17$|_18$|_19$|_20$|_21$")) # remove responses not relevant for Colombia
df2rm = c("v13", "v13_col", "v13_col_sn", "df2rm")
rm(list = df2rm)
```

```{r, Col18-table, echo=FALSE, message=FALSE, results='asis'}

# Load necessary libraries
library(dplyr)
library(tidyr)
library(knitr)
library(purrr)
library(kableExtra)
library(insight)

selected_vars <- grep("v2elsnlfc|v2elsnmrfc|v2clrgstch|v2clrgwkch", colnames(v13_col_sn_2018), value = TRUE)
selected_df <- v13_col_sn_2018 %>%
  select(all_of(selected_vars))

summary_df <- selected_df %>%
  pivot_longer(cols = everything()) %>%
  mutate(row_id = gsub(".*_(\\d+)$", "\\1", name)) %>%
  mutate(column_id = gsub(".*(v2elsnlfc|v2elsnmrfc|v2clrgstch|v2clrgwkch).*", "\\1", name)) %>%
  pivot_wider(names_from = column_id, values_from = value) %>%
  select(-name) %>%
  mutate(row_id = as.numeric(row_id)) %>% 
  group_by(row_id) %>%
  summarise(across(everything(), ~ifelse(all(is.na(.)), NA, first(na.omit(.))))) %>%
  ungroup()

new_col_names <- c("Elections less free", "Elections more free", "Civil liberties stronger", "Civil liberties weaker")
new_row_names <- c(
  "Rural", "Urban", "Less econ devt", "More econ devt", "Inside capital", 
  "Outside capital", "North", "South", "West", "East", "Civil unrest", 
  "Illicit activity", "Sparse pop.", "Remote", "Indigenous", 
  "Ruling party strong", "Ruling party weak"
)

# Rename columns and modify row_id variable
summary_table <- summary_df %>%
  mutate(row_id = new_row_names[as.numeric(row_id) + 1]) %>%
  rename(
    `Survey variable` = row_id,
    `Elec. less free` = v2elsnlfc,
    `Elec. more free` = v2elsnmrfc,
    `Civil lib. stronger` = v2clrgstch,
    `Civil lib. weaker` = v2clrgwkch
  ) 

summary_table %>%
  kable(format = "latex", booktabs = TRUE, caption = "V-Dem response scores for Colombia in 2018") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(0, bold = TRUE)

```

## Systematizing and operationalizing concepts

The country context is important for defining concepts and selecting
appropriate indicators. This occurs for both fuzzy and less fuzzy
concepts. For example, civil unrest and illicit activity are generally
contested or fuzzy concepts. Moreover, they are difficult to observe and
isolate as factors that affect democracy in the same way across
contexts, even within countries. Taking illicit activity, for example,
one area of the country with gang controlled neighborhoods may have
limited voter turnout and less free elections, while another may
tolerate or even encourage electoral competition
[@cordovaLivingGangControlledNeighborhoods2022]. Even basic
categorizations may be contested. For one expert, the North may include
10 provinces, yet for another it may include 8. Some knowledge about
these context-specific contested concepts should inform the
conceptualization process from the start, as these decisions will
ultimately shape how subnational democracy levels are mapped.

After V-Dem responses have been conceptualized for the country in
question, the operationalization of these indicators requires the use of
external (within-country) data. This may come from a wide array of
sources and formats. In Colombia, we consider the indicators as shown in
Table 3. All sources in parentheses are official government sources,
although alternative indicators and sources certainly exist.

```{r, op-Col, echo=FALSE, message=FALSE, results='asis'}
library(flextable)
library(dplyr)
library(officer)
library(kableExtra)

#set_flextable_defaults(fontname = "Computer Modern", fontsize = 10)

vdem_op <- data.frame(
  ID = seq(0, 16, 1), stringsAsFactors = FALSE,
  Response = c("Rural", "Urban", 
                 "Less econ devt", "More econ devt", 
                 "Inside capital", "Outside capital", 
                 "North", "South", "West", "East", 
                 "Civil unrest", "Illicit activity", 
                 "Sparse pop.", "Remote", 
                 "Indigenous", 
                 "Ruling party strong", "Ruling party weak"),
  Indicator = c("Rural Index (DANE)", "Rural Index (DANE)",
        "Mun. Value Added / Dept. GDP (DNP)", "Mun. Value Added / Dept. GDP (DNP)",
        "Geographic distance from center (DANE)", "Geographic distance from center (DANE)", 
        "Six macro-regions (DANE)", "Six macro-regions (DANE)", "Six macro-regions (DANE)", "Six macro-regions (DANE)",
        "Illegally armed groups (Min. Defense)", "Coca cultivation (Min. Defense/UNODC)",
        "Population density (DANE)", "Access (Min. Transportation/Health)",
        "Ethnic population (DANE)", 
        "Pres. margin of victory (RNEC)", "Pres. margin of victory (RNEC)"))


# Print dataframe with merged cells
vdem_op %>%
  kable("latex", escape = FALSE, caption = "Operationalization of V-Dem responses in Colombia") %>%
  kable_styling(full_width = FALSE) %>%
  collapse_rows(columns = 3, valign = "middle") %>%
  row_spec(0, bold = TRUE)


```

## Assigning measures to polygons

In order to visualize V-Dem responses in a geo-referenced environment,
the operationalized country-specific indicators must be assigned some
measure based on the proportional values recorded in each country-year.
This requires a degree of subjectivity. For example, how many votes or
proportion of votes are necessary to signify "strong" support for the
ruling party (question 15)? Or, which municipalities are considered part
of the "North" (question 6) versus the "West" (question 8)? We privilege
the use of granular data that will allow the researcher flexible
interpretations of these questions, as well as facilitate adjustments
based on validation exercises with country experts.

There are at least two approaches to assigning measures to V-Dem's
subnational response variables. One approach is to follow the
dichotomous nature of the survey items themselves. In this way,
municipalities would be categorized according to an arbitrary threshold,
then assigned the value recorded in V-Dem. Another approach, which we
largely pursue below, is to preserve the continuous nature of indicator
data when available, and calculate values based on the cumulative
distribution function (CDF) that can be interacted with V-Dem's
proportional responses.

The following sections show which data from external sources is
collected and assigned 2018 V-Dem values. To reiterate, external data
from Colombia is largely drawn from official sources, but we do not
discount the possibility of alternative data sources. In this exercise
we also include visualizations to demonstrate how each factor appears
depending on how V-Dem data is applied.

## Three stages for mapping

For the purposes of replication, we can summarize the subnational
mapping procedure outlined below into three stages.

The first stage is exploratory, identifying the level of analysis and
available data for the systematized concept in question. In this case,
we observe that Colombia has publicly available municipal-level data
dating back to at least the 1950s, at least for a relevant amount of
time-varying variables reflected in the V-Dem responses (e.g.,
population density, economic development, voting behavior, etc.). We
create a data-frame to join all data at the municipal level using
national statistics department (DANE) codes. These codes are also used
in shape-files and `.dbf` tables, such as Colombia's 2018 Census data
from DANE[^1]. For practical purposes of data collection, some
indicators (e.g., responses 6-9 that ask to identify 4 cardinal
directions) are grouped together.

[^1]: To facilitate integration of the datasets across software, we
    export the Census 2018 attribute table from ArcMaps to `.csv`
    format, then load this data into R, preserving only relevant
    variables for merging purposes. While we use DANE codes to merge
    external data using R, we use the 2018 Census variable "FID" to
    merge `.csv` files into ArcMap. This is because ArcMaps
    automatically translates some string variables (such as the DANE
    codes) into a numeric ones. For this reason, we integrate the FID
    variable for the individual `.csv` files we write. The Census data
    shape-file is useful for initial mapping visualization, containing
    not only the administrative boundaries but also contextual
    information. The
    [dictionary](https://github.com/pmcquest/snvdem24/tree/main/data/geospatial/MGN_ANM_MPIOS)
    is in the Github repository.

Second, we format the data for mapping purposes. This is a multi-step
process whereby external geolocated data must be formatted and merged
with V-Dem data. When data is continuous, as in the rurality index or
economic development measures, for example, this merging process can
take one of two forms. On the one hand, thresholds can be set to define
the areas that correspond to each background concept. On the other hand,
the data can be transformed using the cumulative distribution function
(CDF) before being merged with V-Dem values. For categorical data (e.g.,
cardinal directions) only the first option is used. The sections below
will consider both steps--thresholds and CDF--for assigning values and
creating a measure for the relative salience of factors for subnational
democracy, according to V-Dem country experts.

Finally, the responses are joined to shape-file data (`.shp`) and mapped
using Geographic Information System (GIS) software. In R or in ArcMap,
this can be done by joining data to the attribute table for the base
layer (in this case, the shape-file provided by DANE). During this stage
we create and format layers to show with less or more confidence where
free and fair elections occur (and do not occur), and where civil
liberties are stronger (or weaker), according to the respondents.

```{r, sf1, include=FALSE, echo=FALSE}
#clean things up for the exercise
df2rm = c("selected_df", "summary_df", "summary_table", "top_title", "df2rm")
rm(list = df2rm)

library(sf)
col <- st_read("G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/BaseLayer/MGN_ANM_MPIOS.shp")
col <- col %>%
  select(1:8)
```

```{r, Arc1, include=FALSE, echo=FALSE}
library(readxl)
MGN18 <- read_excel("G:/Shared drives/snvdem/snvdem24/data/geospatial/MGN_ANM_MPIOS/MGN18.xls") 
# We select only relevant variables for merging purposes.
# colnames(MGN18)
MGN18 <- MGN18 %>%
  select(1:7)

# we merge the FID variable to our main shapefile
col <- merge(col, MGN18[, c("MPIO_CDPMP", "FID")], by = "MPIO_CDPMP", all.x = TRUE)
```

## 0-1: Rurality

The survey asks about elections and civil liberties in rural or urban
locations. V-Dem scores for these survey variables reflect the
proportion of respondents who selected rural, urban, or both categories
to describe election fairness and civil liberties.

Rurality or urbanization can be measured in many ways. The World Bank's
[Rural Access Index
(RAI)](https://datacatalog.worldbank.org/search/dataset/0038250) or
European Commission's [Global Human Settlement Layer
(GHSL)](https://ghsl.jrc.ec.europa.eu/CFS.php) are exemplary GIS-based
approaches. [Waldorf and Kim
(2015)](https://purr.purdue.edu/publications/2960/1) create an index
that considers population size, density, remoteness, and built-up areas.
In Colombia, [DANE
(2015)](https://colaboracion.dnp.gov.co/CDT/Estudios%20Econmicos/2015ago6%20Documento%20de%20Ruralidad%20-%20DDRS-MTC.pdf)
uses a classification based on OECD criteria that defines rural
territories as those with 150 or fewer persons per kilometer squared
($km^2$), finding that over 75% of Colombian municipalities do not meet
the threshold of 100 persons per $km^2$. From another perspective, DANE
estimates that close to 4/5 of the population lives in urban centers.

In this exercise, we take a demographic approach that considers the
proportion of the municipal rural population over the total municipal
population. This basic "Rurality Index" measure has a few advantages.
First, it was created by the Economics Center for Development and
Economy (CEDE) at Universidad de Los Andes, using official Census data
from DANE. Second, this proportional measure provides a continuous scale
of municipal rurality that can be can be interacted with the V-Dem score
proportions for the first two variables. We demonstrate this process
below.

```{r, V18_0-1_read, include=FALSE, echo=FALSE}
# municipal-level data for rurality index
IR18 <- read_excel("G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/0-1_Rurality/PANEL_CARACTERISTICAS_GENERALES(2021).xlsx")
colnames(IR18)[3] = "MPIO_CDPMP"
colnames(IR18)[7] = "Year"
IR18 <- IR18 %>%
  filter(Year == 2018)
# this data set also contains important descriptive information (department, province) that we will hang onto
IR18 <- IR18[c("MPIO_CDPMP", "depto", "provincia", "indrural")]

# Because the import creates a numeric field for DANE code, we must convert this numeric variable to character and then assure each observation has the corresponding 5 digits
IR18$MPIO_CDPMP <- as.character(IR18$MPIO_CDPMP)
# Add a 0 before values with 4 digits only
IR18$MPIO_CDPMP <- ifelse(nchar(IR18$MPIO_CDPMP) == 4, paste0("0", IR18$MPIO_CDPMP), IR18$MPIO_CDPMP)

```

We can merge the rurality index with V-Dem scores in at least two ways.
One option is to create a threshold for rural places and assign V-Dem
scores accordingly. For example, we could take the mid-point as a
threshold for rural and urban municipalities: if more than half of the
population is considered rural inhabitants, then the municipality is
considered rural. This method provides a first cut, and may be more
intuitive to interpret. On the other hand, this approach involves the
setting of arbitrary thresholds that may produce misleading results,
especially given the cases grouped around the threshold.

Another option is to interact the rurality index with V-Dem scores and
preserve the continuous nature of the data. We call this the Cumulative
Distribution Function (CDF) approach because it creates a normalized
value for the variable that can be interacted with V-Dem scores
systematically.

```{r, V18_0-1_vdem, include=FALSE, echo=FALSE}
hist(IR18$indrural) # mostly normal, but skewed left (more rural)
# Option #1: interact V-Dem with rural index score
IR18$CDF_0t1 <- pnorm(IR18$indrural) #normal distribution, so we use pnorm (CDF)

# 0: proportion who selected Rural
IR18 <- IR18 %>% 
  mutate(el_c0 = CDF_0t1*v13_col_sn_2018$v2elsnlfc_0) %>%
  mutate(em_c0 = CDF_0t1*v13_col_sn_2018$v2elsnmrfc_0) %>%
  mutate(cs_c0 = CDF_0t1*v13_col_sn_2018$v2clrgstch_0) %>%
  mutate(cw_c0 = CDF_0t1*v13_col_sn_2018$v2clrgwkch_0)
# 1: proportion who selected Urban
IR18 <- IR18 %>% 
  mutate(el_c1 = CDF_0t1*v13_col_sn_2018$v2elsnlfc_1) %>%
  mutate(em_c1 = CDF_0t1*v13_col_sn_2018$v2elsnmrfc_1) %>%
  mutate(cs_c1 = CDF_0t1*v13_col_sn_2018$v2clrgstch_1) %>%
  mutate(cw_c1 = CDF_0t1*v13_col_sn_2018$v2clrgwkch_1)

# Option #2: categorize based on threshold, and assign the corresponding V-Dem score
# "less free and fair subnational elections"
IR18 <- IR18 %>% 
  mutate(el_d0 = case_when(
    indrural > 0.50 ~ v13_col_sn_2018$v2elsnlfc_0, # Rural areas
    TRUE ~ NA_real_)) %>%
  mutate(el_d1 = case_when(
    indrural < 0.50 ~ v13_col_sn_2018$v2elsnlfc_1, # Urban areas
    TRUE ~ NA_real_))
# "more free and fair subnational elections"
IR18 <- IR18 %>% 
  mutate(em_d0 = case_when(
    indrural > 0.50 ~ v13_col_sn_2018$v2elsnmrfc_0,
    TRUE ~ NA_real_)) %>%
  mutate(em_d1 = case_when(
    indrural < 0.50 ~ v13_col_sn_2018$v2elsnmrfc_1,
    TRUE ~ NA_real_))
# "stronger civil liberties"
IR18 <- IR18 %>% 
  mutate(cs_d0 = case_when(
    indrural > 0.50 ~ v13_col_sn_2018$v2clrgstch_0,
    TRUE ~ NA_real_)) %>%
  mutate(cs_d1 = case_when(
    indrural < 0.50 ~ v13_col_sn_2018$v2clrgstch_1,
    TRUE ~ NA_real_))
# "weaker civil liberties"
IR18 <- IR18 %>% 
  mutate(cw_d0 = case_when( 
    indrural > 0.50 ~ v13_col_sn_2018$v2clrgwkch_0, 
    TRUE ~ NA_real_)) %>%
  mutate(cw_d1 = case_when(
    indrural < 0.50 ~ v13_col_sn_2018$v2clrgwkch_1,
    TRUE ~ NA_real_))

# For sf (R), merge this data to shapefile
col <- merge(col, IR18, by = "MPIO_CDPMP", all.x = TRUE)


```

```{r, V18_0vdem-plot, echo=FALSE, message=FALSE, results='hide', fig.cap="Example plots showing V-Dem scores applied to levels of rurality."}
library(ggplot2)
library(gridExtra)
library(stringr)
library(grid)

d0 <- ggplot() +
  geom_sf(data = col, aes(fill = el_d0), color = "transparent") + 
  labs(title = "0. Rural.", caption = "Threshold approach") +
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem (prop.)",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "lightgreen",
    high = "darkgreen") + # Change label position
  theme(legend.title = element_text(size = 8)) +
  theme_void()
c0 <- ggplot() +
  geom_sf(data = col, aes(fill = el_c0), color = "transparent") + 
  labs(title = "0. Rural.", caption = "CDF approach") +
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem (pnorm)",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "lightgreen",
    high = "darkgreen") + # Change label position
  theme(legend.title = element_text(size = 8)) +
  theme_void()

ap_0t1 <- grid.arrange(d0 + theme(plot.title = element_text(size = 16)), 
                               c0 + theme(plot.title = element_text(size = 16)), 
                       top = textGrob("How would you describe the areas of the country in which elections are significantly less free and fair?", gp = gpar(fontsize = 14)),
                               ncol = 2)
suppressMessages(ap_0t1)

```

```{r, V18_0-1_export, include=FALSE, echo=FALSE}

# merge base-layer dataset to obtain FID variable
IR18 <- merge(IR18, MGN18[, c("MPIO_CDPMP", "FID")], by = "MPIO_CDPMP", all.x = TRUE)
IR18 <- IR18 %>%
  rename(indrur0t1 = indrural) %>%
  rename(FID_0t1 = FID)

# For ArcMaps, export the final merge to a .csv file
write.csv(IR18, file = "G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/0-1_Rurality/IR18_0-1.csv", row.names = FALSE)

```

Figure 1 depicts the results of these two approaches using the question:
"How would you describe the areas of the country in which elections are
significantly less free and fair" (`v2elsnlfc`), and only looking at
option #0: Rural. On the left side is the threshold approach and on the
right side is the CDF approach. Clearly, the CDF approach provides much
more detail than the threshold approach, which can only provide a simple
visualization of areas categorized as rural. At the same time, the CDF
approach creates a new continuous measure that is less straightforward
to interpret. The CDF transforms the .875 proportion recorded in V-Dem
for this year in Colombia and, when interacted with the rurality index,
creates a variable whose municipal-level values range from
`r min(IR18$el_c0)` to `r max(IR18$el_c0)`. The benefit of using this
variable, however, is that it provides a granular and demographically
informed estimate of the association between rurality and less free and
fair elections. Because it is a normalized measure, this new estimate
can be compared with the normalized values for other responses,
resulting in a municipal-level estimate based on V-Dem experts'
assessments of the background concept's relevance.

Of course, these composite municipal-level scores are heavily skewed by
how the concept is systematized. In this example, we used a demographic
measure--the rurality index--that relies on Census-defined parameters of
what constitutes a rural inhabitant or not. Census measures may vary
across countries, so the use of this measure may limit its use for
cross-national comparative analysis. Nonetheless, as opposed to the
threshold approach, the CDF approach takes advantage of the continuous
nature of available data. Robustness checks using other continuous
measures of rurality or urbanization would provide more confidence in
the CDF approach going forward.

## 2-3: Economic development

Another advantage of the CDF approach is that it provides a
straightforward method for V-Dem data to be scaled at different levels.
This can be helpful when statistical data are missing at one level but
not at another, or when respondents are likely to have multiple
alternative interpretations. By contrast, the setting of arbitrary
thresholds makes it difficult to scale data or to compare across
measures.

In this section, we consider two measures of "economic development," a
variable that appears in the V-Dem survey under responses 2 ("Areas that
are less economically developed") and 3 ("Areas that are more
economically developed"), using multiple levels of analysis. This
demonstrates how the CDF approach applied to economic and V-Dem data can
provide a flexible tool for understanding multidimensional phenomena
with many possible interpretations and comparable measures.

A common measure of economic development is GDP per capita, however this
measure is problematic when scaled down to the subnational level.
Indeed, according to
[DANE](https://www.dane.gov.co/files/investigaciones/pib/departamentales/pres-PIB-regional-buenas-practicas.pdf),
only 41 countries in the world calculate regional or subnational GDP.
The difficulty may stem from the complex task of isolating economic
productivity for subnational units when inhabitants may be undertaking
inter-municipal daily commutes, for example, or when supply-chains
overlap with multiple departmental jurisdictions. DANE provides
subnational data for GDP (or Producto Bruto Interno, PBI, in Spanish) at
regional and departmental (provincial) levels, but not at the municipal
level. They calculate a historical "retropolation" of departmental GDP
dating back to 1980.[^2]

[^2]: Most departments are covered by the retropolation measure, but
    newer departments (many of which were inducted in 1991) are grouped
    into one unit, and therefore cannot be mapped based on DANE's
    Departmental Code system without data intervention. These are:
    Amazonas (91), Arauca (81), Casanare (85), Guainía (94), Guaviare
    (95), Putumayo (86), San Andrés, Providencia y Santa Catalina
    (Archipiélago) (88), Vaupés (97) y Vichada (99).

At the municipal-level, DANE has created a measure of municipal "value
added", which estimates the relative economic capacity of municipalities
within each department by subtracting consumption levels from gross
production (methodological documentation is included in the
[repository](https://github.com/pmcquest/snvdem24/tree/main/data/geospatial/2018pmq/2-3_EconDevt/biblio)).
Nearly all municipalities are measured by the Municipal Value Added
(VAM) measure, however, data only exists from 2011 onward. It may be
possible that V-Dem country experts are attuned to the economic
disparities occurring within departments, and are considering how these
affect electoral dynamics or civil liberties when responding to the
subnational survey questions.

```{r, V18_2-3_read, include=FALSE, echo=FALSE}
#clean things up for next variable
df2rm = c("IR18", "df2rm", "ap_0t1", "c0", "d0")
rm(list = df2rm)
# municipal-level data for "value added"
VAM18 <- read_excel("G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/2-3_EconDevt/anexo-2020-2021-provisional-valor-agregado-municipio-2011-2021.xlsx", sheet = "Cuadro 9", range = "A11:I1133")
# rename columns to facilitate merging shapefile
VAM18 <- VAM18 %>%
  rename(MPIO_CDPMP = `Código Municipio`) %>%
  rename(DPTO_CCDGO = `Código Departamento`) %>%
  rename(VAM18_2t3 = `Valor agregado\r\n`) %>%
  select(1|3|8)

# merge base-layer dataset to obtain FID variable (for later integration with ArcMaps)
ED18 <- merge(VAM18, MGN18[, c("MPIO_CDPMP", "FID")], by = "MPIO_CDPMP", all.x = TRUE)
ED18 <- ED18 %>%
  rename(FID_2t3 = FID)

# department-level data for "PBI per capita"
PBID <- read_excel("G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/2-3_EconDevt/anex-PIBDep-RetropolacionDepartamento-2022pr.xlsx", sheet = "Cuadro 1", range = "A10:AS36")
PBID <- PBID %>%
  rename(DPTO_CCDGO = `Código Departamento (DIVIPOLA)`) %>%
  select(DPTO_CCDGO, `2018`) %>% #keep only 2018 data 
  rename(PBID18_2t3 = `2018`)

ED18 <- merge(ED18, PBID, by = "DPTO_CCDGO", all.x = TRUE)


```

After compiling these data from DANE at municipal and departmental
levels, we interact each of these measures with the V-Dem responses, as
we did with rurality. Because data for both measures are continuous but
not proportional, we must create new variables capturing the
proportional weight for each department and municipality. Departmental
GDP and municipal VAM data include outliers, especially at the municipal
level. Because data is not normally distributed, we use the
non-parametric empirical cumulative distribution (ECDF) function to
scale them.

```{r, V18_2-3_vdem, include=FALSE, echo=FALSE}
ED18 <- subset(ED18, select = -DPTO_CCDGO) #remove repetitive variable

hist(ED18$PBID18_2t3) #right-skewed, major outlier is Bogota (25% of Colombian GDP and VAM), so the data could be logged
hist(ED18$VAM18_2t3) # heavily skewed by few cases with much higher values
# Because the data is not normally distributed, we use the Empirical CDF
eCDF_2t3d <- ecdf(ED18$PBID18_2t3) 
ED18$eCDF_2t3d <- eCDF_2t3d(ED18$PBID18_2t3) # departments
eCDF_2t3m <- ecdf(ED18$VAM18_2t3)
ED18$eCDF_2t3m <- eCDF_2t3m(ED18$VAM18_2t3) # municipalities


# Option #1: interact V-Dem with ECDF variables
# Department-level data
# 2: proportion who selected less economically developed
ED18 <- ED18 %>% 
  mutate(el_c2d = eCDF_2t3d*v13_col_sn_2018$v2elsnlfc_2) %>%
  mutate(em_c2d = eCDF_2t3d*v13_col_sn_2018$v2elsnmrfc_2) %>%
  mutate(cs_c2d = eCDF_2t3d*v13_col_sn_2018$v2clrgstch_2) %>%
  mutate(cw_c2d = eCDF_2t3d*v13_col_sn_2018$v2clrgwkch_2)
# 3: proportion who selected more economically developed
ED18 <- ED18 %>% 
  mutate(el_c3d = eCDF_2t3d*v13_col_sn_2018$v2elsnlfc_3) %>%
  mutate(em_c3d = eCDF_2t3d*v13_col_sn_2018$v2elsnmrfc_3) %>%
  mutate(cs_c3d = eCDF_2t3d*v13_col_sn_2018$v2clrgstch_3) %>%
  mutate(cw_c3d = eCDF_2t3d*v13_col_sn_2018$v2clrgwkch_3)

# Municipal-level data
# 2: proportion who selected less economically developed
ED18 <- ED18 %>% 
  mutate(el_c2m = eCDF_2t3m*v13_col_sn_2018$v2elsnlfc_2) %>%
  mutate(em_c2m = eCDF_2t3m*v13_col_sn_2018$v2elsnmrfc_2) %>%
  mutate(cs_c2m = eCDF_2t3m*v13_col_sn_2018$v2clrgstch_2) %>%
  mutate(cw_c2m = eCDF_2t3m*v13_col_sn_2018$v2clrgwkch_2)
# 3: proportion who selected more economically developed
ED18 <- ED18 %>% 
  mutate(el_c3m = eCDF_2t3m*v13_col_sn_2018$v2elsnlfc_3) %>%
  mutate(em_c3m = eCDF_2t3m*v13_col_sn_2018$v2elsnmrfc_3) %>%
  mutate(cs_c3m = eCDF_2t3m*v13_col_sn_2018$v2clrgstch_3) %>%
  mutate(cw_c3m = eCDF_2t3m*v13_col_sn_2018$v2clrgwkch_3)

# For sf (R), merge this data to shapefile
col <- merge(col, ED18, by = "MPIO_CDPMP", all.x = TRUE)

```

To demonstrate how the CDF approach facilitates scaling and comparison
of V-Dem data at the subnational level, Figure 2 visualizes both
measures for 2018. Clearly, there is missing data at the departmental
level for the GDP measure given the presence of newer departments.
Obviously, departmental GDP shows less variation than the
municipal-level VAM measure. Nonetheless, there are some similarities
around the center and western parts of the country across both measures.

The comparison shows that depending on the scale of the data and the
economic measures used, one could reach different conclusions about
experts' confidence around development levels and electoral fairness.
Indeed, the northern region displays considerable divergence in V-Dem
levels depending on the scale being used. On the other hand, the port
city of Buenaventura on the Pacific coast is consistent across both
scales.

To be sure, these visualizations provide a limited quantitative measure
of V-Dem country experts' assessment of electoral fairness and economic
development using the ECDF method. Additional quantitative and
qualitative data from entities like the Election Observation Mission
(MOE), for example, would provide additional analysis of electoral
dynamics across the country that can further contextualize the role of
economic development subnationally
[@barrioscabreraMapasFactoresRiesgo2018].

```{r, V18_2-3vdem-plot, echo=FALSE, message=FALSE, results= 'hide', fig.cap="Example plots showing V-Dem scores applied to levels of economic development."}
library(ggplot2)
library(gridExtra)
library(stringr)


c3d <- ggplot() +
  geom_sf(data = col, aes(fill = em_c3d), color = "transparent", alpha = 0.5) + # check results
  labs(title = "3. Areas that are more economically developed", caption = "Department GDP (PBI departamental)") + 
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem (ecdf)",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "lightblue",
    high = "darkblue") + # Change label position
  theme(legend.title = element_text(size = 6)) +
  theme_void()
c3m <- ggplot() +
  geom_sf(data = col, aes(fill = em_c3m), color = "transparent", alpha = 0.5) + # check results
  labs(title = "3. Areas that are more economically developed", caption = "Municipal Value Added (VAM)") +
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem (ecdf)",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "lightblue",
    high = "darkblue") + # Change label position
  theme(legend.title = element_text(size = 6)) +
  theme_void()

# Arrange the plots side by side
ap_2t3 <- grid.arrange(c3d + theme(plot.title = element_text(size = 14)), 
                                c3m + theme(plot.title = element_text(size = 14)), 
                       top = textGrob("How would you describe the areas of the country in which elections are significantly more free and fair?", gp = gpar(fontsize = 14)),         
                       ncol = 2)

suppressMessages(ap_2t3)

```

```{r, V18_2-3_export, include=FALSE, echo=FALSE}

# For ArcMap, export V-Dem merged data to a .csv file that will be joined to map
write.csv(ED18, file = "G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/2-3_EconDevt/ED18_2-3.csv", row.names = FALSE)

#clean things up for next variable
df2rm = c("ED18", "ED18_2t3", "eCDF_2t3d", "eCDF_2t3m", "PBID", "VAM18", "ap_2t3", "df2rm")
rm(list = df2rm)
```

## 4-5: Inside or outside of capital city

Subnational V-Dem data can be used to examine divergences in the degree
of confidence across elections or civil liberty questions. To
demonstrate this cross-question comparison subnationally, this section
considers responses #4-5 which ask how elections or civil liberties vary
"Inside / Outside the capital city." In the context of the V-Dem survey,
which is cross-national, respondents may reasonably interpret this
question to refer to the national capital (Bogota). University of Los
Andes Center for Studies on Economic Development (CEDE) provides a
measure for this variable (linear distance in kilometers) for which we
can merge V-Dem data using the CDF approach. Although the survey asks
two questions, the first ("Inside the capital city") is one extreme of
the second ("Outside"), so we create one continuous variable that
reflects the relative proximity of municipalities from Bogota.

```{r, V18_4-5_read, include=FALSE}
# municipal-level data for distance to capital city
CEDE18 <- read_excel("G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/4-5_DisCapital/PANEL_CARACTERISTICAS_GENERALES(2021).xlsx")
colnames(CEDE18)[3] = "MPIO_CDPMP"
colnames(CEDE18)[7] = "Year"
CEDE18 <- CEDE18 %>%
  filter(Year == 2018)
# this data set also contains important descriptive information (department, province, rurality index) that we will hang onto
CEDE18 <- CEDE18[c("MPIO_CDPMP", "discapital", "disbogota")]

# Because the import creates a numeric field for DANE code, we must convert this numeric variable to character and then assure each observation has the corresponding 5 digits
CEDE18$MPIO_CDPMP <- as.character(CEDE18$MPIO_CDPMP)
# Add a 0 before values with 4 digits only
CEDE18$MPIO_CDPMP <- ifelse(nchar(CEDE18$MPIO_CDPMP) == 4, paste0("0", CEDE18$MPIO_CDPMP), CEDE18$MPIO_CDPMP)

# Note: It may be possible to calculate this variable in ArcMaps using the field calculator, but this requires advanced techniques.

```

```{r, V18_4-5_vdem-scale, include=FALSE}

hist(CEDE18$disbogota) # skewed right (more distant)
# Option #1: interact V-Dem with rural index score
eCDF_4t5 <- ecdf(CEDE18$disbogota) #normal distribution, so we use pnorm (CDF)
CEDE18$eCDF_4t5 <- eCDF_4t5(CEDE18$disbogota) # municipalities

# 4-5. proportion who selected inside capital city and outside
CEDE18 <- CEDE18 %>% 
  mutate(el_c4t5 = ifelse(MPIO_CDPMP == 11001, v13_col_sn_2018$v2elsnlfc_4, v13_col_sn_2018$v2elsnlfc_5*eCDF_4t5)) %>%
  mutate(em_c4t5 = ifelse(MPIO_CDPMP == 11001, v13_col_sn_2018$v2elsnmrfc_4, v13_col_sn_2018$v2elsnmrfc_5*eCDF_4t5)) %>%
  mutate(cs_c4t5 = ifelse(MPIO_CDPMP == 11001, v13_col_sn_2018$v2clrgstch_4, v13_col_sn_2018$v2clrgstch_5*eCDF_4t5)) %>%
  mutate(cw_c4t5 = ifelse(MPIO_CDPMP == 11001, v13_col_sn_2018$v2clrgwkch_4, v13_col_sn_2018$v2clrgwkch_5*eCDF_4t5))


# For sf (R), merge this data to shapefile
col <- merge(col, CEDE18, by = "MPIO_CDPMP", all.x = TRUE)
```

Figure 3 below compares two survey questions: where elections or less
free and fair, and where civil liberties are weaker. The proportions for
inside versus outside the capital city are 0:0.375 for less free
elections, and 0:0.6 for weaker civil liberties, respectively. When we
map the product of the continuous measure of distance from the capital
with both proportions, we see a clear difference in the salience of this
criterion for elections compared to civil liberties. The contrasting
gradients show the degree to which respondents are more confident civil
liberties are weaker outside the capital city than they are that
elections are less free and fair according to this criterion.

```{r V18_4-5_vdem-plot, echo=FALSE, message=FALSE, results='hide', fig.cap="Example plots showing V-Dem scores applied to distance from capital city."}
library(ggplot2)
library(gridExtra)
library(stringr)


c5el <- ggplot() +
  geom_sf(data = col, aes(fill = el_c4t5), color = "transparent", alpha = 0.5) + # check results
  labs(title = "Elections are less free and fair", caption = "5. Outside the capital city") + 
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem (ecdf)",
    title.position = "top",  
    label.position = "right"),
    low = "pink",
    high = "darkred",
    limits = c(0, 0.6)) + # Change limits to facilitate comparison
  theme(legend.title = element_text(size = 6)) +
  theme_void()
c5cw <- ggplot() +
  geom_sf(data = col, aes(fill = cw_c4t5), color = "transparent", alpha = 0.5) + # check results
  labs(title = "Civil liberties are weaker", caption = "5. Outside the capital city") +
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem (ecdf)",
    title.position = "top", 
    label.position = "right"),
    low = "pink",
    high = "darkred", 
    limits = c(0, 0.6)) + # Change limits to facilitate comparison
  theme(legend.title = element_text(size = 6)) +
  theme_void()

# Arrange the plots side by side
ap_4t5 <- grid.arrange(c5el + theme(plot.title = element_text(size = 14)), 
                                c5cw + theme(plot.title = element_text(size = 14)), 
                       top = textGrob("Comparison between survey question confidence levels", gp = gpar(fontsize = 14)),    
                                ncol = 2)
suppressMessages(ap_4t5)
```

```{r, V18_4-5_export, include=FALSE, echo=FALSE}

# merge base-layer dataset to obtain FID variable
CEDE18 <- merge(CEDE18, MGN18[, c("MPIO_CDPMP", "FID")], by = "MPIO_CDPMP", all.x = TRUE)
CEDE18 <- CEDE18 %>%
  rename(discap4t5 = discapital) %>%
  rename(disbog4t5 = disbogota) %>% 
  rename(FID_4t5 = FID)

# For ArcMaps, export the final merge to a .csv file
write.csv(CEDE18, file = "G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/4-5_DisCapital/DC18_4-5.csv", row.names = FALSE)

#clean things up for next variable
df2rm = c("CEDE18", "c5el", "c5cw", "df2rm", "ap_4t5")
rm(list = df2rm)

```

## 6-9: Cardinal directions

Although most subnational criterion in the V-Dem survey can be
conceptualized using continuous measures, some criterion are less
amenable to this operation. A clear example are responses #6-9 which ask
respondents to assess elections or civil liberties across four cardinal
directions: North, South, West, East. While these categories could be
measured using fuzzy sets, where middling municipalities would have
multiple memberships to different regions, this section settles on a
binary measure of membership, grouping together pre-existing regional
categories for the purposes of demonstrating how survey questions can be
layered visually.

Rather than using four cardinal regions in Colombia, DANE classifies the
country into six
"[macro-regions](https://www.dane.gov.co/index.php/estadisticas-por-tema/informacion-regional/informacion-estadistica-desagregada-con-enfoque-territorial-y-diferencial/informacion-del-dane-para-la-toma-de-decisiones-en-departamentos-y-ciudades-capitales)":
a) Caribe, b) Eje Cafetero (Antioquia), c) Centro Oriente, d) Llano, e)
Sur, and f) Pacifico.[^3] In order to create a new variable called
"Cardinal", these six macro-regions are grouped in the following manner:
a)-b) as "North", c)-d) as "East, e) as"South", and f) as "West". Once
grouped, V-Dem data for responses #6-9 are simply assigned to these
categories for each subnational survey question, without any
transformation.

[^3]: A public-access table of municipalities classified into these six
    regions can be found
    [here]((<https://www.datos.gov.co/Mapas-Nacionales/Departamentos-y-municipios-de-Colombia/xdk5-pm3f/about_data>).
    There are other divisions, such as the five regions used by CEDE
    studies (Andina, Caribe, Pacifica, Orinoquia, Amazonia), but for now
    we go with the prior classification.

```{r, V18_6-9_read, include=FALSE, echo=FALSE}
#data from DANE regions requires some formatting before merging
reg <- read_csv("G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/6-9_Cardinal/Departamentos_y_municipios_de_Colombia_20240312.csv")
reg <- reg %>%
  rename(mpio = `CÓDIGO DANE DEL MUNICIPIO`)
reg$MPIO_CDPMP <- sprintf("%.3f", reg$mpio) #Codigo municipio should have 5 digits
reg$MPIO_CDPMP <- gsub("\\.", "", reg$MPIO_CDPMP)
reg$MPIO_CDPMP <- ifelse(nchar(reg$MPIO_CDPMP) == 4, sprintf("0%s", reg$MPIO_CDPMP), reg$MPIO_CDPMP)

```

```{r, V18_6-9_vdem, include=FALSE, echo=FALSE}
unique(reg$REGION)
reg <- reg %>%
  mutate(cdir6t9 = case_when(
    REGION %in% c('Región Caribe', 'Región Eje Cafetero - Antioquia') ~ 'North',
    REGION %in% c('Región Centro Oriente', 'Región Llano') ~ 'East',
    REGION == 'Región Centro Sur' ~ 'South',
    REGION == 'Región Pacífico' ~ 'West',
    TRUE ~ NA_character_
  ))

# V-Dem merge:
# "less free and fair subnational elections"
reg <- reg %>% 
  mutate(el_t6t9 = case_when(
    cdir6t9 == "North" ~ v13_col_sn_2018$v2elsnlfc_6,  
    cdir6t9 == "South" ~ v13_col_sn_2018$v2elsnlfc_7,
    cdir6t9 == "West" ~ v13_col_sn_2018$v2elsnlfc_8,  
    cdir6t9 == "East" ~ v13_col_sn_2018$v2elsnlfc_9,
    TRUE ~ NA_real_ 
))
# "more free and fair subnational elections"
reg <- reg %>% 
  mutate(em_t6t9 = case_when(
    cdir6t9 == "North" ~ v13_col_sn_2018$v2elsnmrfc_6,
    cdir6t9 == "South" ~ v13_col_sn_2018$v2elsnmrfc_7,
    cdir6t9 == "West" ~ v13_col_sn_2018$v2elsnmrfc_8,  
    cdir6t9 == "East" ~ v13_col_sn_2018$v2elsnmrfc_9,
    TRUE ~ NA_real_ 
))
# "stronger civil liberties"
reg <- reg %>% 
  mutate(cs_t6t9 = case_when(
    cdir6t9 == "North" ~ v13_col_sn_2018$v2clrgstch_6, 
    cdir6t9 == "South" ~ v13_col_sn_2018$v2clrgstch_7,
    cdir6t9 == "West" ~ v13_col_sn_2018$v2clrgstch_8,  
    cdir6t9 == "East" ~ v13_col_sn_2018$v2clrgstch_9,
    TRUE ~ NA_real_
))
# "weaker civil liberties"
reg <- reg %>% 
  mutate(cw_t6t9 = case_when( 
    cdir6t9 == "North" ~ v13_col_sn_2018$v2clrgwkch_6,  
    cdir6t9 == "South" ~ v13_col_sn_2018$v2clrgwkch_7,
    cdir6t9 == "West" ~ v13_col_sn_2018$v2clrgwkch_8,  
    cdir6t9 == "East" ~ v13_col_sn_2018$v2clrgwkch_9,
TRUE ~ NA_real_ 
))

# For sf (R), merge this data to shapefile
col <- merge(col, reg, by = "MPIO_CDPMP", all.x = TRUE)

#for mapping below...
col <- col %>%
  mutate(m1_t6t9 = (el_t6t9 + cw_t6t9) / 2) %>%
  mutate(m2_t6t9 = (em_t6t9 + cs_t6t9) / 2)

```

```{r V18_6-9_vdem-plot, echo=FALSE, message=FALSE, results='hide', fig.cap="Example plots showing V-Dem scores applied to importance of region (cardinal directions)."}


elcw6t9 <- ggplot(col, aes(fill = m1_t6t9)) +
  geom_sf(color = NA) +
  labs(title = "Less F&F Elections and Weak Civil Liberties",
       caption = "6-9. North-South-West-East") +
  scale_fill_gradient(name = "V-Dem (mean)",
                      low = "pink",
                      high = "darkred",
                      guide = guide_colorbar(title.position = "top",
                                             label.position = "right")) +
  scale_alpha_continuous(range = c(0.2, 1), guide = FALSE) + 
  theme_void()

emcs6t9 <- ggplot(col, aes(fill = m2_t6t9)) +
  geom_sf(color = NA) +
  labs(title = "More F&F Elections and Strong Civil Liberties",
       caption = "6-9. North-South-West-East") +
  scale_fill_gradient(name = "V-Dem (mean)",
                      low = "lightgreen",
                      high = "darkgreen", 
                      guide = guide_colorbar(title.position = "top",
                                             label.position = "right")) +
  scale_alpha_continuous(range = c(0.2, 1), guide = FALSE) + 
  theme_void()


# Arrange the plots side by side
ap_6t9 <- grid.arrange(elcw6t9 + theme(plot.title = element_text(size = 14)), 
                                emcs6t9 + theme(plot.title = element_text(size = 14)), 
                       top = textGrob("Combined (layered) elections and civil liberties dimensions", gp = gpar(fontsize = 14)),
                       ncol = 2)
suppressMessages(ap_6t9)
#ggsave("G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/6-9_Cardinal/6t9_comparison.jpg", arranged_plots4, device = "jpg")
```

```{r, V18_6-9_export, include=FALSE, echo=FALSE}

# For ArcMaps, merge base-layer dataset to obtain FID variable (for later integration with ArcMaps)
reg <- merge(reg, MGN18[, c("MPIO_CDPMP", "FID")], by = "MPIO_CDPMP", all.x = TRUE)
reg <- reg %>% 
  rename(FID_6t9 = FID)
#Belen de Bajira (Choco) does not have an FID number, which causes problems in ArcMap when merging. This municipality was only recently inaugurated.
reg <- reg[complete.cases(reg$FID_6t9), ] 

# Export the final merge to a .csv file that can be read into ArcMaps:
write.csv(reg, file = "G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/6-9_Cardinal/CD18_6-9.csv", row.names = FALSE)

#clean things up for next variable
df2rm = c("reg", "elcw6t9", "emcs6t9", "ap_6t9", "df2rm")
rm(list = df2rm)

```

One important advantage to using mapping software is the use of layers.
Layers can be used to produce composite images reflecting values from
more than one variable. For example, the proportional values for less
free and fair elections and weak civil liberties by region can produce a
rough composite image of subnational democracy. Figure 4 reproduces two
sets of layers across the four regions as classified above in order to
compare confidence across democratic dimensions. There are only two
variables involved in each map, so a simple mean can be used to simulate
how the layered values would look. Clearly, the South is considered less
democratic than the North, although both proportional scales suggest
that respondents' confidence in these regional variations are low: the
range for the mean of respondent agreement around less free and fair
elections and weaker civil liberties is `r range(col$m1_t6t9)` while the
range for the mean of respondent agreement around more free and fair
elections and stronger civil liberties is `r range(col$m2_t6t9)`.

*Note: The following five variables have yet to be operationalized.
Below each are some initial ideas for the variables we will use.*

## 10: Civil unrest

-   protest data (perhaps from ACLED)
-   HRDAG violence estimates from recent Truth Commission report
-   MinDefense insurgency presence
-   NAVCO on social forces for democracy

## 11: Illicit activity

-   homicide rates (perhaps more relevant for urban areas than for rural
    areas)
-   coca growth (perhaps more relevant for rural areas than for urban
    areas)

## 12: Sparse population density

-   Census data: rural disperse population / total population (likely
    highly correlated with rurality index used for responses #0-1)

## 13: Remoteness

-   road density index

## 14: Indigenous

-   Census data

## 15-16: National ruling party

The survey questions also ask whether elections or civil liberties vary
according to "Areas where national ruling party or group is strong /
weak" (responses #15-16). The crudest measure for this may be a
dichotomous variable identifying areas where the president receives a
majority of the vote or not. A more variegated measure would be the
percentage of votes cast for the president-elect vs. runner-up, for
example. In this case, we can use the CDF function to create a
continuous variable to be interacted with V-Dem respondent data for
these items in each survey question.

The measure for ruling party support used here will be the percentage of
vote that supports the ruling party. In 2018, there was a runoff
election (second round) between Gustavo Petro and Ivan Duque; Duque won,
so his is the ruling party. This makes things easier because there are
only 2 candidates, the winner and runner-up. Elections without second
rounds (prior to 1994, as well as 2002 and 2006) would require
additional coding.

```{r, V18_15-16_read, include=FALSE, echo=FALSE}
# Import voting data from Registraduria (RNEC).
library(readr)
rp18 <- read_csv("G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/15-16_RulingParty/2018_presidencia_segunda_vuelta_dta_c27d4515ed.csv") # this long dataset contains only 2 candidates because it was a second-round election. other datasets will include more rows depending on the number of candidates running.

#Codigo municipio should have 5 digits
rp18$MPIO_CDPMP <- as.character(rp18$codmpio)
rp18$MPIO_CDPMP <- ifelse(nchar(rp18$MPIO_CDPMP) == 4, sprintf("0%s", rp18$MPIO_CDPMP), rp18$MPIO_CDPMP)
# merge base-layer dataset to obtain FID variable (for later integration with ArcMaps)
rp18 <- merge(rp18, MGN18[, c("MPIO_CDPMP", "FID")], by = "MPIO_CDPMP", all.x = TRUE)
```

```{r, V18_15-16_edit, include=FALSE, echo=FALSE}

# Step 1: Calculate total votes cast at municipal level (allows for percentages later on).
rp18 <- rp18 %>%
  group_by(FID) %>%
  mutate(votototal = sum(votos)) %>%
  mutate(vtpc = votos/votototal) %>% #optional: useful for quick look at percentages, SD, etc. 
  mutate(vtpc0 = round(vtpc*100, 2)) 

# For mapping at the municipal level, values may take the form of: percent difference in votes between top winner and runner-up candidate(s). This requires a few steps:
#Step 2: Pivot municipal voting behavior into wide format.
p18d0 <- rp18 %>%
  group_by(FID, codigo_lista) %>%
  pivot_wider(names_from = codigo_lista, values_from = votos) %>% # Pivot to wide format
  rename(Petro = `1`,
         Duque = `2`,
         nomark = `997`,
         null = `998`,
         blank = `999`)

## Step 3: Create new dfs for voting behavior at municipal level -- is there a more efficient code?
p18d2 <- p18d0 %>%
  select(FID, Petro) %>%
  filter(!is.na(Petro))
p18d2 <- p18d2[!duplicated(p18d2$FID), ]
p18d3 <- p18d0 %>%
  select(FID, Duque) %>%
  filter(!is.na(Duque))
p18d3 <- p18d3[!duplicated(p18d3$FID), ]
p18d4 <- p18d0 %>%
  select(FID, nomark) %>%
  filter(!is.na(nomark))
p18d4 <- p18d4[!duplicated(p18d4$FID), ]
p18d5 <- p18d0 %>%
  select(FID, null) %>%
  filter(!is.na(null))
p18d5 <- p18d5[!duplicated(p18d5$FID), ]
p18d6 <- p18d0 %>%
  select(FID, blank) %>%
  filter(!is.na(blank))
p18d6 <- p18d6[!duplicated(p18d6$FID), ]
p18d7 <- p18d0 %>%
  group_by(FID) %>%
  summarise(vtotal = first(votototal))
p18d7 <- p18d7[!duplicated(p18d7$FID), ]

RPs <- list(p18d2, p18d3, p18d4, p18d5, p18d6, p18d7)
merged_RP <- Reduce(function(x, y) merge(x, y, by = "FID", all = TRUE), RPs)

df2rm2 = c("rp18", "p18d0", "p18d2", "p18d3", "p18d4", "p18d5", "p18d6", "p18d7", "df2rm2", "RPs")
rm(list = df2rm2)

## Step 3: Create margin of support variable for winner and runner-up candidates. Positive value indicates support for winner; negative indicates opposition. Variable order may vary depending on the election year data from RNEC
RP18 <- merged_RP %>%
  mutate(MOV_top2 = Duque-Petro) %>% #Raw vote-count margin of victory
  mutate(Duque_pct = Duque/vtotal) %>%
  mutate(Petro_pct = Petro/vtotal) %>%
  mutate(MOV_pct = Duque_pct-Petro_pct) #Standardized percent margin of victory

round((sum(RP18$Duque)/sum(RP18$vtotal))*100, 2) # 52.3% for Duque
round((sum(RP18$Petro)/sum(RP18$vtotal))*100, 2) # 40.8% for Petro

round(mean(RP18$MOV_pct)*100, 2) #Duque municipal-level margin is 26.3% on average (quite high).
round(sd(RP18$MOV_pct), 3) #SD for the margin is larger than the mean (39.7%) suggesting abnormal distribution
hist(RP18$MOV_pct, main = "Histogram of MOV", xlab = "MOV", col = "skyblue", border = "black") #left-tail long (margin for Petro)

```

```{r, V18_15-16_vdem, include=FALSE, echo=FALSE}

# Option #1: interact V-Dem with margin-of-victory
eCDF_15t6 <- ecdf(RP18$MOV_pct) #abnormal distribution, so we use eCDF
RP18$eCDF_15t6 <- eCDF_15t6(RP18$MOV_pct) # municipalities

# 15-16. proportion votes for ruling party (Duque) 
# 15: proportion of respondents who selected "Areas where the national ruling party or group is strong":
RP18 <- RP18 %>% 
  mutate(el_c15 = eCDF_15t6*v13_col_sn_2018$v2elsnlfc_15) %>%
  mutate(em_c15 = eCDF_15t6*v13_col_sn_2018$v2elsnmrfc_15) %>%
  mutate(cs_c15 = eCDF_15t6*v13_col_sn_2018$v2clrgstch_15) %>%
  mutate(cw_c15 = eCDF_15t6*v13_col_sn_2018$v2clrgwkch_15)
# 16: proportion of respondents who selected "Areas where the national ruling party or group is weak":
RP18 <- RP18 %>% 
  mutate(el_c16 = eCDF_15t6*v13_col_sn_2018$v2elsnlfc_16) %>%
  mutate(em_c16 = eCDF_15t6*v13_col_sn_2018$v2elsnmrfc_16) %>%
  mutate(cs_c16 = eCDF_15t6*v13_col_sn_2018$v2clrgstch_16) %>%
  mutate(cw_c16 = eCDF_15t6*v13_col_sn_2018$v2clrgwkch_16)


# For sf (R), merge this data to shapefile
col <- merge(col, RP18, by = "FID", all.x = TRUE)

```

```{r, V18_15-16vdem-plot, echo=FALSE, message=FALSE, results='hide', fig.cap="Example plot showing V-Dem scores applied to levels of support for ruling party."}

c15cs <- ggplot() +
  geom_sf(data = col, aes(fill = cs_c15), color = "transparent") + # check results
  labs(title = "15. Areas where the nat. ruling party is strong", caption = "Margin of victory (for Duque)") + 
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem (ecdf)",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "lightblue",
    high = "darkblue",
    limits = c(0, 0.34)) + # Change limits to facilitate comparison
  theme(legend.title = element_text(size = 6)) +
  theme_void()
c16cw <- ggplot() +
  geom_sf(data = col, aes(fill = cw_c16), color = "transparent") + # check results
  labs(title = "16. Areas where the nat. ruling party is weak", caption = "Margin of victory (for Duque)") +
  scale_fill_continuous(guide = guide_colorbar(
    title = "V-Dem (ecdf)",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "pink",
    high = "darkorchid4",
    limits = c(0, 0.34)) + # Change limits to facilitate comparison 
  theme(legend.title = element_text(size = 6)) +
  theme_void()

# Arrange the plots side by side
ap_15t6 <- grid.arrange(c15cs + theme(plot.title = element_text(size = 14)), 
                                c16cw + theme(plot.title = element_text(size = 14)), 
                        top = textGrob("Describe the areas of the country where government officials’ respect for civil liberties is significantly stronger / weaker.", gp = gpar(fontsize = 12)),
                                ncol = 2)

suppressMessages(ap_15t6)
```

Official voting data comes from Registraduria (RNEC), and it includes
all possible vote entries: top 2 candidates, as well as "no mark",
blank, or null ballots. The data show that Duque received
`r round((sum(RP18$Duque)/sum(RP18$vtotal))*100, 2)`% of the national
vote, and Petro won
`r round((sum(RP18$Petro)/sum(RP18$vtotal))*100, 2)`%. At the
municipal-level, Duque's margin of victory
(`r round(mean(RP18$MOV_pct)*100, 2)`) is larger than the national
average, and the standard deviation (`r round(sd(RP18$MOV_pct), 3)`) is
larger than the mean, suggesting an abnormal distribution of data. For
this reason, we use the empirical CDF function to merge V-Dem data
accordingly. Applying this function results in normalized values between
0-1, and higher values on this scale indicate a higher margin of victory
for Duque.

The proportions of respondent agreement are low; none of the survey
questions accrue above 0.333 agreement. Nonetheless, we can visualize
how these scores are distributed across the territory, according to
operationalization for ruling party support. Figure 5 shows
municipal-level data reflecting the product of the empirical CDF for
V-Dem data around civil liberties and presidential margin of victory.

```{r, V18_15-16_export, include=FALSE, echo=FALSE}
# Export the final merge to a .csv file that can be read into ArcMaps
RP18 <- RP18 %>% 
  select(1:3|7|11:19) %>%
  rename(FID_15t6 = FID) %>%
  rename(Pe_15t6 = Petro) %>%
  rename(Du_15t6 = Duque) %>%
  rename(vt_15t6 = vtotal) %>%
  rename(MV18_15t6 = MOV_pct) %>%
  mutate(MV18_15t6 = round(MV18_15t6, 5))
#Belen de Bajira (Choco) does not have an FID number, which causes problems in ArcMap when merging. This municipality was only recently inaugurated.
RP18 <- RP18[complete.cases(RP18$FID_15t6), ]
# Export data to a csv file that will be joined to map
write.csv(RP18, file = "G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/15-16_RulingParty/RP18_15-16.csv", row.names = FALSE)


```

## Summary indices

```{r V18_summary-stats, message = FALSE}
#clean things up for next variable
df2rm = c("RP18", "merged_RP", "c15cs", "c16cw", "ap_15t6", "df2rm")
rm(list = df2rm)

# calculate the arithmetic mean for the continuous variables
col$el_i = (col$el_c0+col$el_c1+col$el_c2m+col$el_c3m+col$el_c4t5+col$el_t6t9+col$el_c15+col$el_c16) / 8
col$em_i = (col$em_c0+col$em_c1+col$em_c2m+col$em_c3m+col$em_c4t5+col$em_t6t9+col$em_c15+col$em_c16) / 8
col$cs_i = (col$cs_c0+col$cs_c1+col$cs_c2m+col$cs_c3m+col$cs_c4t5+col$cs_t6t9+col$cs_c15+col$cs_c16) / 8
col$cw_i = (col$cw_c0+col$cw_c1+col$cw_c2m+col$cw_c3m+col$cw_c4t5+col$cw_t6t9+col$cw_c15+col$cw_c16) / 8

```

```{r, V18_summary-plot, echo=FALSE, message=FALSE, results='hide', fig.cap= "Example plots showing summary averages across V-Dem survey dimensions."}

sum_el <- ggplot() +
  geom_sf(data = col, aes(fill = el_i), color = "transparent") + # check results
  labs(title = "Areas where elections are less free and fair") + 
  scale_fill_continuous(guide = guide_colorbar(
    title = "Average",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "pink",
    high = "darkred") +
  theme(legend.title = element_text(size = 6)) +
  theme_void()

sum_em <- ggplot() +
  geom_sf(data = col, aes(fill = em_i), color = "transparent") + # check results
  labs(title = "Areas where elections are more free and fair") + 
  scale_fill_continuous(guide = guide_colorbar(
    title = "Average",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "lightgreen",
    high = "darkgreen") +
  theme(legend.title = element_text(size = 6)) +
  theme_void()

sum_cs <- ggplot() +
  geom_sf(data = col, aes(fill = cs_i), color = "transparent") + # check results
  labs(title = "Areas where civil liberties are stronger") + 
  scale_fill_continuous(guide = guide_colorbar(
    title = "Average",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "lightblue",
    high = "darkblue") +
  theme(legend.title = element_text(size = 6)) +
  theme_void()

sum_cw <- ggplot() +
  geom_sf(data = col, aes(fill = cw_i), color = "transparent") + # check results
  labs(title = "Areas where civil liberties are weaker") + 
  scale_fill_continuous(guide = guide_colorbar(
    title = "Average",
    title.position = "top",  # Change title position
    label.position = "right"),
    low = "pink",
    high = "darkorchid") +
  theme(legend.title = element_text(size = 6)) +
  theme_void()

# Arrange the plots side by side
ap_i <- grid.arrange(sum_el + theme(plot.title = element_text(size = 10)), 
                     sum_em + theme(plot.title = element_text(size = 10)), 
                     sum_cs + theme(plot.title = element_text(size = 10)), 
                     sum_cw + theme(plot.title = element_text(size = 10)), 
                     top = "Average municipal scores for municipalities across V-Dem dimensions",
                     ncol = 2)

suppressMessages(ap_i)
```

```{r, V18_summary_export, include=FALSE, echo=FALSE}

# check correlations for summary variables
emel_i <- ggplot(col, aes(x = em_i, y = el_i)) +
  geom_point() +
  labs(title = "Scatterplot of X and Y",
       x = "More F&F Elections",
       y = "Less F&F Elections") +
  theme_minimal()
cscw_i <- ggplot(col, aes(x = cs_i, y = cw_i)) +
  geom_point() +
  labs(title = "Scatterplot of X and Y",
       x = "Stronger Civil Liberties",
       y = "Weaker Civil Liberties") +
  theme_minimal()
ec_i <- grid.arrange(emel_i + theme(plot.title = element_text(size = 10)), 
                     cscw_i + theme(plot.title = element_text(size = 10)), 
                     top = "Scatterplots of summary measures",
                     ncol = 2)

# Outlier in EM x CS appears to be Bogota
# Calculate the mean and standard deviation
mean_em <- mean(col$em_i)
sd_em <- sd(col$em_i)
# Calculate Z-scores
z_scores <- (col$em_i - mean_em) / sd_em
# Define the threshold for outliers
threshold <- 3
# Identify outlier observations
outliers <- col[abs(z_scores) > threshold, "MPIO_CNMBR"]
# Print the outlier observations
print(outliers)


# Convert the shapefile to a DataFrame
col_x <- st_drop_geometry(col)
# Export data to a csv file that will be joined to map
write.csv(col_x, file = "G:/Shared drives/snvdem/snvdem24/data/geospatial/2018pmq/x_Summary/V18.csv", row.names = FALSE)

#clean things up for next variable
df2rm = c("ap_i", "sum_cs", "sum_cw", "sum_el", "sum_em", "df2rm")
rm(list = df2rm)
```

We have produced subnational measures of the levels of subnational
democracy that are believed to correspond to demographic and political
attributes. These measures are certainly imperfect, as democracy is not
a simple function of a single variable; it has many covariates. With
several measures of democracy in hand, it becomes possible to combine
them into a more valid measure. Validity should be enhanced because if
many covariates suggest that a location is democratic, then it probably
is more democratic; if many covariates suggest that a location is
undemocratic, then it is probably very much less democratic. If
different covariates point in different directions, then most likely the
location is closer to the middle of the distribution than to either
extreme. The underlying logic is that strength on one criterion tends to
compensate for weakness in another. This kind of relationship can be
operationalized as a sum, an average, or a weighted average. Ideally we
prefer weighted averages, as averages preserve the 0-1 range of values
and weights employ a more relaxed assumption than "unweighted" averages,
which actually enforce a strong assumption of equal weighting. Weights
can be derived empirically via a data reduction technique such as factor
or principal components analysis, which yield factor or component
weights based on the commonalities among our criteria. These techniques
can also evaluate whether it makes sense to treat all these criteria as
unidimensional. If it does not, then we would do well to generate
multiple indices that measure somewhat independent dimensions of
democracy.

In Figure 6, the normalized data is aggregated to produce a summary
index based on the averages for each survey question at the
municipal-level. These aggregated scores provide an initial measure of
operationalized concepts, revealing geographically located "pockets" of
stronger or weaker V-Dem respondent confidence around subnational
factors affecting subnational democracy in Colombia during 2018. The
summary averages range from around 0-0.34: this signifies that the
maximum level of confidence is around 1 in 3 coders who agree on
subnational factors for democracy (at least for the factors considered
above).

# Conclusion

As noted previously, validation is a crucial step in producing data
based on expert ratings. We have not yet been able to do a systematic
validation of the data we have so far, but we have some observations
about face validity based on the maps we have produced. In general, our
data indicate higher levels of democracy (both in free and fair
elections and in protections for civil liberties) in the major urban
areas and lower levels in rural or remote areas such as the Amazon....
[To be fleshed out after we have a summary map.]

Systematic validation will require comparisons with independent
assessments of subnational democracy in Colombia. An initial source for
such information will be responses to open-ended versions of V-Dem's
four questions provided by a small number of university and master's
students at a Colombian university in February 2024. They were asked to
identify the most salient locations (defined any way they chose) in
Colombia and years where they would expect valid indicators to show
better or worse situations. We can check our data against their
expectations. At some point we could solicit the same kinds of responses
from a more senior panel of Colombian experts. Ultimately, the best kind
of validation would be administering an online survey asking V-Dem's
country experts on Colombia to code these same questions for Colombian
departments or municipalities for selected years. We do not know whether
such an effort is feasible, but it is probably impossible to know until
we try.

Eventually we hope to extend this exercise to other countries where
others have generated measures of subnational democracy, such as
Argentina, Mexico, India, the United States, and Russia. We do not
downplay the difficulty of such an effort. It requires access to very
specific geo-coded data on multiple demographic and political
indicators, which are probably not available for many countries.
Furthermore, producing cross-nationally comparable data involves an
additional layer of complication. The data we have produced for Colombia
is relative: the degree to which local situations are better or worse
than the national norm. Converting such relative data to absolute levels
will require both anchoring the ratings to a national score and
establishing the range of variation within which the relative scores can
vary. V-Dem's Clean Elections Index could serve as the anchor for the
subnational elections variables and its Civil Liberties Index could
serve as the anchor for the subnational civil liberties variables. The
same anchoring indices could be used for other countries. However, we do
not yet have a proposed solution for setting the range of variation
around these anchors. This would require careful conceptual and
methodological effort.

Colombia is among the best cases for this initial proof-of-concept
research due to the well-developed geographic data that DANE provides
(and for which the Lucy Institute, where co-author Matthew Sisk works,
has direct access). Replication in other countries would make it
possible to reach more general conclusions about the validity of the
technique and to decide whether it is promising enough to extend to many
more countries. If results are encouraging, there may be a potential for
extending of this technique to measure other political concepts
subnationally.

# References
