---
title: "Visualizing subnational democracy using Varieties of Democracy (V-Dem) survey
  data"
author: 'Patrick McQuestion^[Department of Political Science & Kroc Institute, University of Notre Dame. Email: <pmcquest@nd.edu>], Michael Coppedge^[Department of Political Science, University of Notre Dame. Email: <mcoppedg@nd.edu>], Matthew Sisk^[Department of Applied and Computational Mathematics and Statistics, University of Notre Dame. Email: <msisk1@nd.edu>]'
date: "First version: 04 February 2024. This version: `r format(Sys.Date(), '%d %B %Y')`."
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: yes
  word_document: default
abstract: "The objective of this methodological paper is to introduce a procedure
  for operationalizing Varieties of Democracy (V-Dem) data for use at the subnational
  level. This effort in democracy studies is data-driven and descriptive, and is intended
  to stimulate discussion on the methodological potential for V-Dem to be applicable
  within countries. After describing the structure and data generating process behind
  V-Dem coding, this paper outlines the systematic conceptualization of the background
  concepts contained within elections and civil liberties surveys, and assigns measures
  to those concepts on a geographic scale. Once systematized, the data are reduced
  using a factor analysis technique that calculates latent attributes whose weights
  can be applied to V-Dem proportional scores. We demonstrate this process for a portion
  of the 21 response items in the survey, using 2018 cross-sectional data from Colombia
  at the municipal-level."
editor_options:
  markdown:
    wrap: 72
bibliography: References/BiblioRA24.bib
csl: G:/Shared drives/snvdem/snvdem24/report/References/ajps.csl
link-citations: yes
linkcolor: black
subtitle: An application using the case of Colombia
header-includes:
- \usepackage{dcolumn}
- \usepackage{setspace}
- \doublespacing
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, 
                      message=FALSE, 
                      fig.path='Figs/',
                      fig.width=12, fig.height=8) 
setwd("G:/Shared drives/snvdem/snvdem24/report")
```

# Introduction

There is a need for subnational data on many political phenomena:
violence, democracy, service provision, human rights, gender relations,
discrimination, and a range of other topics. The need is especially
great in large countries, which are likely to contain considerable
geographic variation on these dimensions within their borders. To the
extent that such variation exists, claims about the "typical" levels of
these dimensions for whole countries are less useful, either as
descriptions or as independent or dependent variables.

In this paper, we demonstrate a technique for combining several
underexplored Varieties of Democracy (V-Dem) variables with geocoded
demographic and political data to measure subnational democracy at the
municipal level. We introduce a pipeline that begins with
conceptualization and measurement, before proceeding to
operationalization via factor analysis and weighting, resulting in
quantitative scores that can be utilized for further comparative
subnational research. This exercise is designed to test the viability of
mapping cross-sectional country-level data before expanding the dataset
into panel format. Internal data from V-Dem is subset to 2018 for
Colombia, a country with a large supply of geospatial demographic and
social data.

In order to map V-Dem scores, the procedure outlined in this paper uses
simple definitions and systematizes them by selecting one or two
principal indicators for each response. Rather than discount alternative
systematizations, our objective is to visualize these indicators in
order to characterize subnational democracy in a given context, and
therefore we prioritize reliable over valid measurements
[@adcockMeasurementValidityShared2001]. This mapping procedure involves
multiple stages, from the systematization of concepts with explicit
definitions, to the weighting of factors using coder-level responses,
and finally the calculation of geographic subnational "scores" based on
these measures.

# Measuring subnational democracy

While data based on direct observation are generally assumed to be
superior in comparative political analysis, there are some drawbacks to
this data collection strategy below the national-level. In the first
place, hard information about subnational units, such as states or
provinces, counties or municipalities, or even smaller districts, is
often scarce. Second, when the information exists, it can be difficult
or expensive to obtain, and may be collected in non-standardized ways.
Third, such "objective" data -- counts, rates, texts, etc. -- can be
invalid when they capture only a small part of the concept one needs to
measure. It may be feasible to count the number of newspapers and
broadcasters in a locality, but a valid measure of freedom of
information, for example, should also consider the size of their
audiences, their ability to carry out independent investigations, or
their editorial autonomy. Fourth, whether a directly observable measure
matters is often conditional on other factors. The lack of arrests of
journalists in North Korea is not evidence of press freedom; governing
parties sometimes win elections for decades even when elections are very
competitive; party platforms and speeches are often insincere and
inaccurate; and texts of constitutions and laws can be "parchment"
rules, dead letters that are not implemented in practice.

For these reasons, there are sometimes advantages to following an
alternative measurement strategy: relying on the subjective judgments of
multiple country experts. This approach can compensate for some of the
weaknesses of "objective" data [@bernhardMakingEmbeddedKnowledge2017].
Experts absorb information from multiple sources: reading, following
news, conversations, personal experiences, and deep familiarity with the
social and political context. This enables them to reconcile diverse and
often conflicting kinds of information and synthesize it into an
internally consistent account of how things work. Their understanding
extends to a sense of the past and the direction of trends. Because they
possess this depth of understanding, experts are less likely to be
deceived by insincere statements or *de jure* arrangements that are
ignored in practice, and they are less likely to consider one attribute
definitive when they know that other conditions render it irrelevant.

Of course, there is also an obvious disadvantage to subjective
knowledge: different experts tend to synthesize somewhat different
understandings. They arrive at different interpretations of the same
reality. This is why it is crucial to consult multiple experts, which
makes it possible to have greater confidence in the interpretations on
which experts agree, and to exercise prudent caution about matters on
which there is little agreement. Post hoc validation is also essential.
There are no guarantees that subjective ratings will yield useful
information, but it is worth trying as long as we also subject the
outcomes to multiple forms of validation.

## Existing research

Subnational democracy is an area where subjective expert ratings have
the potential to improve measurement. There are hundreds of democracy
measures, but almost all of them are at the national level of analysis
[@bernhardEconomicPerformanceInstitutional2001;
@bertelsmannstiftungBertelsmannTransformationIndex2005;
@boixCompleteDataSet2013; @coppedgeVDemDatasetV132023;
@economistintelligenceuniteiuDemocracyIndex20102010;
@freedomhouseMethodology2015; @buhlmannQualityDemocracyDemocracy2008;
@pemsteinDemocraticCompromiseLatent2010; @marshallPolityIVProject2014;
@przeworskiModernizationTheoriesFacts1997;
@skaaningLexicalIndexElectoral2015; @vanhanenNewDatasetMeasuring2000].
The few exceptions [@mickeyPathsOutDixie2015;
@mcmannDemocraticConsolidationTheory2018;
@gervasoniRentierTheorySubnational2010;
@giraudyVarietiesSubnationalUndemocratic2013] cover only one country
(although Giraudy does two).

In a recent methdological effort,
@perezsandovalMeasuringAssessingSubnational2023 designed an Index of
Subnational Electoral Democracy (ISED). The ISED tracks electoral
outcomes at the provincial level (the first jurisdictions below the
national-level) for 11 countries over 40 years, using a framework
attuned to the complexities around territorial conflation around
country-level systems such as judicial branches and political parties.
However, this index relies on electoral data rather than country
expertise. As alerted to earlier, electoral data may be collected in
non-standardized ways, especially in local jurisdictions. More
importantly, the number of parties may not always signify a competitive
political climate, nor does vote count always imply effective democratic
participation. Without considering the relevance of other factors, such
as the strength of civil liberties, electoral information alone will not
always be an efficient or valid measure of subnational democracy.

In additon to providing a novel subnational index, Perez Sandoval
provides a helpful summary of the concepts, operations, and data
collection methods used by other researchers of subnational democracy.
Among the eleven authors cited, only four -- @hernándezvaldez2000,
Ortega and Somuano in @meixueironájera2014, @gervasoni2018, and
@grumbach2023 -- attempt to measure beyond the electoral dimesion of
democracy at the subnational level. These authors include measures of
civil liberties components -- such as human rights violations, local
radio and newspapers, or institutional constraints -- however they do so
only for individual countries (Mexico, Argentina, and the United States)
and at the provincial or state level. All four authors use advanced data
reduction techniques, such as Bayesian factor analysis or principal
component analysis, to arrive at subnational democracy scores, however
they stop short of providing a method for cross-national comparison.

Finally, pivoting to subjective expert data,
@mcmannExplainingSubnationalRegime2021 use a selection of V-Dem
variables to measure "unevenness" of elections and civil liberties in
most countries, but theirs is a national-level concept that tells us
nothing about where democracy is better or worse within a state.

## V-Dem subnational data

V-Dem data contains several variables that provide a partial path to
measurement at the subnational level. There are four questions on
elections and civil liberties: Subnational election area less free and
fair characteristics (`v2elsnlfc`), Subnational election area more free
and fair characteristics (`v2elsnmrfc`), Stronger civil liberties
characteristics (`v2clrgstch`), and Weaker civil liberties
characteristics (`v2clrgwkch`). Although elections and civil liberties
are far from capturing every aspect of democracy, they cover the two
most central criteria for polyarchy
[@dahlPolyarchyParticipationOpposition1971].

```{r vdem-q, echo=FALSE, message=FALSE, results='asis'}
library(flextable)
library(dplyr)
library(officer)
library(kableExtra)
library(knitr)

vdem_q <- data.frame(
  Question = c("Question: How would you describe the areas of the country in which elections are significantly less free and fair? [v2elsnlfc]",
               "Question: How would you describe the areas of the country in which elections are significantly more free and fair? [v2elsnmrfc]",
               "Question: How would you describe the areas of the country where government officials’ respect for civil liberties is significantly stronger? [v2clrgstch]", 
               "Question: How would you describe the areas of the country where government officials’ respect for civil liberties is significantly weaker? [v2clrgwkch]",
               "Clarification: Choose all that apply.",
               "Responses:",
               "0: Rural. (0=No, 1=Yes)",
               "1: Urban. (0=No, 1=Yes)",
               "2: Areas that are less economically developed. (0=No, 1=Yes)",
               "3: Areas that are more economically developed. (0=No, 1=Yes)", 
               "4: Inside the capital city. (0=No, 1=Yes)",
               "5: Outside the capital city. (0=No, 1=Yes)",
               "6: North. (0=No, 1=Yes)",
               "7: South. (0=No, 1=Yes)", 
               "8: West. (0=No, 1=Yes)",
               "9: East. (0=No, 1=Yes)",
               "10: Areas of civil unrest (including areas where insurgent groups are active). (0=No, 1=Yes)",
               "11: Areas where illicit activity is widespread. (0=No, 1=Yes)", 
               "12: Areas that are very sparsely populated. (0=No, 1=Yes)",
               "13: Areas that are remote (difficult to reach by available transportation, for example). (0=No, 1=Yes)", 
               "14: Areas where there are indigenous populations. (0=No, 1=Yes)", 
               "15: Areas where the national ruling party or group is strong. (0=No, 1=Yes)", 
               "16: Areas where the national ruling party or group is weak. (0=No, 1=Yes)",
               "17: Areas that were subject to a longer period of foreign rule. (0=No, 1=Yes)",
               "18: Areas that were subject to a shorter period of foreign rule. (0=No, 1=Yes)",
               "19: Areas that were recently subject to foreign rule. (0=No, 1=Yes)",
               "20: Areas that have not recently been subject to foreign rule. (0=No, 1=Yes)",
               "21: None of the above. (0=No, 1=Yes)"
               ))

# Print the table

vdem_q %>%
  kable(format = "latex", escape = FALSE, caption = "V-Dem questions and clarifications", col.names = NULL) %>%
  kable_styling(full_width = FALSE, latex_options = c("hold_position", "scale_down")) %>%
  column_spec(1:ncol(vdem_q), width = "45em") %>%
  row_spec(0:3, italic = TRUE)


```

Table 1 reproduces the four questions, along with the same 21 "choose
all that apply" criteria that apply to all questions. In the
country-year V-Dem data sets, each criteria becomes a separate variable
that is no longer binary, but rather the proportion of experts who chose
the "yes" response to that option. Thus, in actuality we have 84 (21\*4)
variables to work with.

These unusual options have advantages and disadvantages. On the positive
side, these options ask experts to identify the criteria that
distinguish locations where democracy is better from locations where it
is worse. The options are described in a generic way so that they could
be relevant in any country that V-Dem rates, anywhere in the world, and
over more than two centuries of history. In this paper we deal with 16
of them, leaving aside the options referring to foreign rule (\# 17-20)
which is not very relevant for Colombia, and option #21 which refers to
a negative response ("None of the above"). Still, even these few
response options are varied enough to cast a wide net that captures
diverse criteria for distinguishing where democracy is better or worse.

On the negative side, first, all of these options are conceptually
binary. In effect, they ask "Is this a useful discriminating criterion
or not?" rather than seeking information about degrees of relevance.
Second, the options do not name a specific concept or indicator that
should be used in empirical analysis. This decision is left up to each
country expert and V-Dem does not reveal--and in fact does not
*know*--on what information country experts based their assessments.
Third, the options do not specify a threshold that should be used to
decide whether a location does or does not satisfy the criterion.
Fourth, many of the criteria overlap and are therefore somewhat
interchangeable. This is especially likely for the binaries of
rural-urban, low-high income, sparsely-densely populated and remote
areas, which may also overlap with areas of illicit activity, civil
unrest, and indigenous population. Finally, the criteria are not mapped
onto well-defined locations such as states, counties, or cities that
would have known administrative boundaries.

Given these features, additional work is required to render the V-Dem
data useful for measuring democracy at the subnational level. The
procedure described below is structured around the following steps:
coder-level analysis, selection of indicators (operationalization), data
reduction, and finally mapping.

# Selection of indicators and measurement

In order to operationalize V-Dem criteria for a subnational context, we
choose one or two specific indicators -- from official state data or
others' scholarly research -- to measure them. For each indicator, we
either choose a threshold to represent the binary concept of whether the
criterion applies to a location or not; or we redefine the concept to be
continuous so that it can measure the degree to which each location
satisfies the criterion. In this case, most indicators work well as
matters of degree: income, population density, strength of the ruling
party, etc. We prefer the continuous criteria where possible. However,
other criteria are probably best left binary, such as the categories of
North, South, West, and East -- although even these could be measured as
fuzzy sets.

The country context is important for defining concepts and selecting
appropriate indicators. This occurs for both fuzzy and less fuzzy
concepts. For example, civil unrest and illicit activity are generally
contested or fuzzy concepts. Moreover, they are difficult to observe and
isolate as factors that affect democracy in the same way across
contexts, even within countries. Taking illicit activity, for example,
one area of the country with gang controlled neighborhoods may have
limited voter turnout and less free elections, while another may
tolerate or even encourage electoral competition
[@cordovaLivingGangControlledNeighborhoods2022]. Even basic
categorizations may be contested. For one expert, the North may include
10 provinces, yet for another it may include 8. Some knowledge about
these context-specific contested concepts should inform the
conceptualization process from the start, as these decisions will
ultimately shape how subnational democracy levels are mapped.

In the following subsections, we justify our choices of indicators for
the discriminating criteria and explain how we convert the majority to
continuous scales using the cumulative distribution function (CDF). The
CDF standardizes each criterion on a 0-1 scale so that they are more
easily comparable.

Additionally, we combine natural sets of criteria that can be
operationalized by the same indicators. For example, the concepts of
rural and urban can be measured by a rurality index, and less or more
economic development can be measured by a continuous scale of GDP per
capita or multidimensional poverty index. This has implications for our
measurement approach using V-Dem proportional scores, because these too
must be combined correspondingly. In Colombia, we consider the
indicators as shown in Table 3. All sources in parentheses are official
government sources, although alternative indicators and sources
certainly exist.

```{r, op-Col, echo=FALSE, message=FALSE, results='asis'}
library(flextable)
library(dplyr)
library(officer)
library(kableExtra)

#set_flextable_defaults(fontname = "Computer Modern", fontsize = 10)

vdem_op <- data.frame(
  ID = seq(0, 16, 1), stringsAsFactors = FALSE,
  Response = c("Rural", "Urban", 
                 "Less econ devt", "More econ devt", 
                 "Inside capital", "Outside capital", 
                 "North", "South", "West", "East", 
                 "Civil unrest", "Illicit activity", 
                 "Sparse pop.", "Remote", 
                 "Indigenous", 
                 "Ruling party strong", "Ruling party weak"),
  Indicator = c("Rural Index (DANE)", "Rural Index (DANE)",
        "Mun. Value Added / IPM (DNP)", "Mun. Value Added / IPM (DNP)",
        "Geographic distance from center (DANE)", "Geographic distance from center (DANE)", 
        "Six macro-regions (DANE)", "Six macro-regions (DANE)", "Six macro-regions (DANE)", "Six macro-regions (DANE)",
        "Forced displacement / Confinements (Min. Defense)", "Robberies / Homicides / Coca cultivation (Min. Defense/UNODC)",
        "Population density (DANE)", "Distance to market (CEDE))",
        "Ethnic populations (DANE)", 
        "Pres. margin of victory (RNEC)", "Pres. margin of victory (RNEC)"))


# Print dataframe with merged cells
vdem_op %>%
  kable("latex", escape = FALSE, caption = "Operationalization of V-Dem responses in Colombia") %>%
  kable_styling(full_width = FALSE) %>%
  collapse_rows(columns = 3, valign = "middle") %>%
  row_spec(0, bold = TRUE)


```

## Criteria #0-1: Rurality

The survey asks about elections and civil liberties in rural or urban
locations. Rurality or urbanization can be measured in many ways. The
World Bank's Rural Access Index (RAI)[^1] or European Commission's
Global Human Settlement Layer (GHSL)[^2] are exemplary GIS-based
approaches. [@waldorf] create an index that considers population size,
density, remoteness, and built-up areas. In Colombia, the National
Statistics Administrative Department (DANE) uses a classification based
on the Organization for Economic Co-operation and Development (OECD)
criteria that defines rural territories as those with 150 or fewer
persons per kilometer squared ($km^2$). Using this criteria, the
National Planning Department (DNP) finds that over 75% of Colombian
municipalities do not meet the threshold of even 100 persons per $km^2$.
From another perspective, DNP estimates that close to 4/5 of the
population lives in urban centers.
[@departamentonacionaldeplaneación2014]

[^1]: See: <https://datacatalog.worldbank.org/search/dataset/0038250>.

[^2]: See: <https://ghsl.jrc.ec.europa.eu/CFS.php>.

For operationalizing the rurality criterion, we use a demographic
measure that considers the proportion of the municipal rural population
over the total municipal population. This basic "Rurality Index" measure
has a few advantages. First, it was created by the Economics Center for
Development and Economy (CEDE) at *Universidad de Los Andes*, using
official Census data from DANE. Second, this proportional measure
provides a continuous scale of municipal rurality that can facilitate
cross-municipal comparison.

## Criteria #2-3: Economic development

Economic development appears in the subnational V-Dem criteria under
responses 2 ("Areas that are less economically developed") and 3 ("Areas
that are more economically developed"). A common measure of economic
development is GDP per capita, however this measure is problematic when
scaled down to the subnational level. According to DANE, only 41
countries in the world calculate regional or subnational GDP
[@buenasp2020]. The difficulty may stem from the complex task of
isolating economic productivity for subnational units when inhabitants
may be undertaking inter-municipal daily commutes, for example, or when
supply-chains overlap with multiple departmental jurisdictions. DANE
provides subnational data for GDP (or *Producto Bruto Interno*, PBI, in
Spanish) at regional and departmental (provincial) levels, but not at
the municipal level.[^3]

[^3]: Most departments are covered by the GDP measure, but newer
    departments (many of which were inducted in 1991) are grouped into
    one unit, and therefore cannot be mapped based on DANE's
    Departmental Code system without data intervention. These are:
    Amazonas (91), Arauca (81), Casanare (85), Guainía (94), Guaviare
    (95), Putumayo (86), San Andrés, Providencia y Santa Catalina
    (Archipiélago) (88), Vaupés (97) y Vichada (99).

At the municipal-level, DANE has created a measure of municipal "value
added", which estimates the economic capacity of municipalities within
each department by subtracting consumption levels from gross production
[@departamentoadministrativonacionaldeestadística2022]. Nearly all
municipalities are measured by the Municipal Value Added (VAM) measure,
however, data only exists from 2011 onward. Additionally, DANE estimates
a Multidimensional Poverty Index (IPM) consisting of 5 dimensions:
education levels, childhood conditions, employment, health, and access
to public services and housing conditions
[@departamentoadministrativonacionaldeestadística2021]. Using these two
municipal-level indicators produces a more comprehensive
operationalization of more or less economically developed areas.

## Criteria #4-5: Inside or outside of capital city

The surveys ask how elections or civil liberties vary "Inside / Outside
the capital city." In the context of the V-Dem survey, which is
cross-national, respondents may reasonably interpret this question to
refer to the national capital (Bogotá). *University of Los Andes* Center
for Studies on Economic Development (CEDE) provides a measure for this
variable (linear distance in kilometers). Although the survey asks two
questions, the first ("Inside the capital city") is one extreme of the
second ("Outside"), so we create one continuous variable that reflects
the relative proximity of municipalities from Bogotá.

## Criteria #6-9: North, South, West, East

Although most criteria in the subnational V-Dem questions can be
conceptualized using continuous measures, some criteria are less
amenable to this operation. A clear example are criteria #6-9 which ask
respondents to assess elections or civil liberties across four cardinal
directions: North, South, West, and East. While these categories could
be measured using fuzzy sets, where middling municipalities would have
multiple memberships to different regions, we use a binary measure of
membership, grouping together pre-existing regional categories. Rather
than using four cardinal regions in Colombia, DANE classifies the
country into six "macro-regions": a) Caribe, b) Eje Cafetero
(Antioquia), c) Centro Oriente, d) Llano, e) Sur, and f) Pacifico
[@dane-i].[^4] We group the six macro-regions in the following manner:
a)-b) as "North"; c)-d) as "East"; e) as "South"; and f) as "West".

[^4]: A public-access table of municipalities classified into these six
    regions can be found
    [here]((<https://www.datos.gov.co/Mapas-Nacionales/Departamentos-y-municipios-de-Colombia/xdk5-pm3f/about_data>).
    There are other divisions, such as the five regions used by CEDE
    studies (Andina, Caribe, Pacifica, Orinoquia, Amazonia), but for now
    we go with the prior classification.

## Criterion #10: Civil unrest

The survey questions ask to identify whether elections or civil
liberties are stronger or weaker in "Areas of civil unrest (including
areas where insurgent groups are active)." This criterion can be
operationalized using many potential indicators, including the presence
of insurgent forces, events of forced confinement, protests and
mobilizations, or forced displacement, among others. For the purposes of
this paper, we use reported cases of forced displacement and forced
confinement events. Both indicators are measured by Colombia's national
Victims' Registry, in place since 2005.

## Criterion #11: Illicit activity

The survey questions also ask to identify the strength of elections or
civil liberties in "Areas where illicit activity is widespread." This
concept can take many forms, especially given characteristics of the
local environment. We consider three indicators of widespread illicit
activity: robberies, homicides, and manual illicit crop eradication. The
data for the first two indicators are compiled by the Ministry of
Defense, and coca eradication data are compiled by Colombia's Drugs
Observatory.

## Criterion #12: Sparse population density

The surveys ask coders whether stronger or weaker elections or civil
liberties are correlated with "Areas that are very sparsely populated."
As mentioned before, population density is used to measure levels of
rurality. In this case, however, we include a raw population density
measure of total inhabitants per $km^2$ rather than differentiate
between urban and rural populations. These data come from DNP, using
information from DANE and the Geographic Institute Agustin Codazzi
(IGAC).

## Criterion #13: Remoteness

For the criterion "Areas that are remote (difficult to reach by
available transportation, for example)," we use a measure of the linear
distance of a municipality to the nearest supermarket (*mercado
mayorista*), as calculated by CEDE. If a supermarket is located in the
municipality, this takes a value of `0`. Alternative measures may
include distance to the nearest secondary or tertiary road or to a major
city.

## Criterion #14: Indigenous

The surveys ask about elections or civil liberties in "Areas where there
are indigenous populations." This is salient in Colombia, where the 2018
Census identified indigenous to comprise 13.6% of the total population
[@departamentoadministrativonacionaldeestadística2019]. We include this
demographic information, and also include the population of inhabitants
that identify as pertaining to an ethnic group. Both of these measures
are based on the 2018 Census data from DANE.

## Criteria #15-16: Support for national ruling party

The survey questions ask whether elections or civil liberties vary
according to "Areas where national ruling party or group is strong /
weak". We utilize official voting data comes from the National Civil
Registry (RNEC) to calculate the percentage of votes cast for the
president-elect vs. the runner-up, in 2018. In 2018, there was a runoff
election (second round) between Gustavo Petro and Ivan Duque; Duque won,
so his is the ruling party. This makes things easier because there are
only 2 candidates, the winner and runner-up. Elections without second
rounds (prior to 1994, as well as 2002 and 2006) would require
additional coding. To operationalize this criteria, we create a
continuous measure of the percent margin of victory at the municipal
level (at the national level, Duque's margin of victory was close to 13
percentage points in 2018).

# Data reduction

Overlapping criteria could be an advantage if they are treated as
substitutable, and therefore aggregated with an additive or weighted
average formula [@goertzSocialScienceConcepts2006]. The underlying logic
is that strength on one criterion tends to compensate for weakness in
another. This kind of relationship can be operationalized as a sum, an
average, or a weighted average. Ideally we prefer weighted averages, as
averages preserve the 0-1 range of values and weights employ a more
relaxed assumption than "unweighted" averages, which actually enforce a
strong assumption of equal weighting. Weights can be derived empirically
via a data reduction technique such as factor or principal components
analysis, which yield factor or component weights based on the
commonalities among our criteria. These techniques can also evaluate
whether it makes sense to treat all these criteria as unidimensional. If
it does not, then we would do well to generate multiple indices that
measure somewhat independent dimensions of democracy.

In this case, we use a factor analysis data reduction technique to
identify latent factors that emerge from the data. Exploratory factor
analysis using the set of variables associated with V-Dem criteria
provides a set of latent factors with different relative weights for
each variable.

```{r V18_FA1, include = FALSE, echo = FALSE}
## Factor analysis of Colombia variables
library(tidyverse)
library(knitr)
library(psych)
library(GPArotation)

colvars_cdf <- read.csv("G:/Shared drives/snvdem/snvdem24/report/analysis/1_col18e.csv", encoding = "UTF-8")
#Codigo municipio should have 5 digits
colvars_cdf$MPIO_CDPMP <- as.character(colvars_cdf$MPIO_CDPMP)
colvars_cdf$MPIO_CDPMP <- ifelse(nchar(colvars_cdf$MPIO_CDPMP) == 4, sprintf("0%s", colvars_cdf$MPIO_CDPMP), colvars_cdf$MPIO_CDPMP)
#----- standardized variables -----
# omitting the cardinal directions variable for now
fa_cdf <- fa(colvars_cdf[, c(5:6, 8:9, 20:32)], nfactors = 4, rotate = "oblimin", 
             scores = "regression", SMC=FALSE, fm="minres")
# Try different numbers of factors....

fa_cdf[["Structure"]]

preds_cdf <- as.data.frame(predict(fa_cdf, colvars_cdf[, c(5:6, 8:9, 20:32)]))

```

![Results from the factor
analysis](figs/Loadings_FA.png){width="4.28in"}

Figure 2 depicts the results of our data reduction technique, revealing
4 latent factors corresponding to demographic and political attributes
at the subnational level in Colombia for 2018. These dimensions can be
characterized as such:

-   MR3 = Richer, populated, urban environments with everyday violence
    (#2-3, 12, and 11)

-   MR1 = Indigenous concentrations with less official party (Duque)
    supporters (#14 mainly, some of 12, and maybe 15-16)

-   MR2 = Poor, rural disperse, and insecure areas (#12, 2-3, 11)

-   MR4 = Peripheral and poor ethnically populated areas (#4-5, 14, 2-3)

The total variance explained by these four factors is 88.9%. The results
suggest that the first two factors explain 60% of the data variation,
and these are characterized by either more economically developed,
populated areas, or less populated but concentrated indigenous and
ethnic areas with less crime. Crime appears to group together with both
more and less economically developed areas. Poorer areas distant from
Bogotá explain a large portion of variation on its own (over 14%).

[pending further analysis, as are the following sections]

# Weighting factors using V-Dem responses

Our next task is to explore ways to combine these factor scores into a
summary measure of a concept closer to "democracy." For this task, we
turn to the data from the V-Dem survey. We describe the data generated
by the subnational V-Dem surveys and how to apply them to the latent
factor scores using a weighting technique.

The degree of agreement among experts can be interpreted as a graded
measure of the degree to which each criterion is likely to be useful for
discriminating between better and worse locations. Arguably, it is also
reasonable to interpret these proportions as, in part, reflections of
degrees of fair elections or civil liberties because the more democratic
a subnational unit is, the more likely experts are to notice it. In
effect, we assume that experts perceive degrees of democracy, but also
have their own subjective thresholds on that continuum that they use to
distinguish more democratic situations from less democratic ones when
they are asked to make a binary classification. Experts' thresholds
differ, but the more democratic a unit is on the underlying continuous
dimension, the larger the proportion of experts that will judge it to be
in the "more democratic" category. This psychological model is the basis
for the Bayesian Item Response Thoery (IRT) measurement model that V-Dem
uses to aggregate multiple coders' ratings into point measures for
country-years [@marquardtIRTModelsExpertCoded2018;
@pemsteinMeasurementModelLatent2023]. Although it would be ideal to
utilize a comparable model for this project, we lack sufficient data to
attain the statistical power equivalent to that available in V-Dem.
Instead, we rely on coder-level information to assess the relative
importance of the subnational survey criteria for a given country in a
given year.

## Proportion scores

Coder scores for each criteria-year observation are proportions.
However, these proportional scores are "noisy" data, in the sense that
coders may not have access to country-wide municipal data on phenomena
such as illicit activity or population sparsity. Moreover, some
information may be missing especially in areas with weak institutional
capacity.

To visualize how these scores vary across survey items, Table 2 shows
the proportional scores for V-Dem data subset to Colombia for 2018. We
use these scores to demonstrate the mapping procedure below. As the
table illustrates, some survey criteria register high degrees of
variation across surveys (e.g., Rural or Urban, Economic development,
Capital city, Civil unrest, Illicit activity) while others register
relatively lower variation (e.g., North-South-West-East, Ruling party
strong or weak). However, in this case, extreme ends of the range
(scores closer to `0` or closer to `1`) indicate more agreement around
the salience of a criteria, and conversely, scores towards the middle
(closer to `0.5`) indicate lower amounts of consensus among voters.

```{r, Vdem0, include=FALSE, echo=FALSE}
# make a clean slate
rm(list = ls())
# load needed packages
library(tidyverse)
library(vdemdata)
library(ggplot2)
library(gridExtra)

# Create a data frame of version 13 dataset:
v13 <- vdem
# this df appears to be the same as MC's "full + others" data set: read_dta("C:/Users/mcoppedg/Dropbox/MC/VDemFiles/Archive/V13/V-Dem-CY-Full+Others/V-Dem-CY-Full+Others-v13.dta")

# Filter to just Colombia since 1899:
v13_col <- filter(v13, country_name=="Colombia" & year > 1899)

# Select just the ID and subnational variables
# (Note: v2elsnless and v2elsnmore are not in this df. We will make sure to 
# include them in our request to the data manager.)
v13_col_sn <- v13_col %>%
  select(country_id, country_name, country_text_id, historical_date, year, 
         v2elsnlfc_0:v2elsnlfc_21, v2elsnmrfc_0:v2elsnmrfc_21,
         v2clrgstch_0:v2clrgstch_21, v2clrgwkch_0:v2clrgwkch_21)

# Listing the variable names to check our work:
#names(v13_col_sn)
# Export the result to a CSV file:
#write.csv(v13_col_sn, file = "v13_col_sn.csv")
```

```{r, v13_col_2018, include=FALSE, echo=FALSE}
# create subset of V-Dem data for 2018 
v13_col_sn_2018 <- subset(v13_col_sn, year == 2018) 
v13_col_sn_2018 <- v13_col_sn_2018 %>%
  select(-matches("_17$|_18$|_19$|_20$|_21$")) # remove responses not relevant for Colombia
df2rm = c("v13", "v13_col", "v13_col_sn", "df2rm")
rm(list = df2rm)
```

```{r, Col18-table, echo=FALSE, message=FALSE, results='asis'}

# Load necessary libraries
library(dplyr)
library(tidyr)
library(knitr)
library(purrr)
library(kableExtra)
library(insight)

selected_vars <- grep("v2elsnlfc|v2elsnmrfc|v2clrgstch|v2clrgwkch", colnames(v13_col_sn_2018), value = TRUE)
selected_df <- v13_col_sn_2018 %>%
  select(all_of(selected_vars))

summary_df <- selected_df %>%
  pivot_longer(cols = everything()) %>%
  mutate(row_id = gsub(".*_(\\d+)$", "\\1", name)) %>%
  mutate(column_id = gsub(".*(v2elsnlfc|v2elsnmrfc|v2clrgstch|v2clrgwkch).*", "\\1", name)) %>%
  pivot_wider(names_from = column_id, values_from = value) %>%
  select(-name) %>%
  mutate(row_id = as.numeric(row_id)) %>% 
  group_by(row_id) %>%
  summarise(across(everything(), ~ifelse(all(is.na(.)), NA, first(na.omit(.))))) %>%
  ungroup()

new_col_names <- c("Elections less free", "Elections more free", "Civil liberties stronger", "Civil liberties weaker")
new_row_names <- c(
  "Rural", "Urban", "Less econ devt", "More econ devt", "Inside capital", 
  "Outside capital", "North", "South", "West", "East", "Civil unrest", 
  "Illicit activity", "Sparse pop.", "Remote", "Indigenous", 
  "Ruling party strong", "Ruling party weak"
)

# Rename columns and modify row_id variable
summary_table <- summary_df %>%
  mutate(row_id = new_row_names[as.numeric(row_id) + 1]) %>%
  rename(
    `Response Criteria` = row_id,
    `Elec. less free` = v2elsnlfc,
    `Elec. more free` = v2elsnmrfc,
    `Civil lib. stronger` = v2clrgstch,
    `Civil lib. weaker` = v2clrgwkch
  ) 

summary_table %>%
  kable(format = "latex", booktabs = TRUE, caption = "V-Dem response scores for Colombia in 2018") %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(0, bold = TRUE)

```

With regard to scores of `0`, the data do not distinguish between coders
who selected "no" and coders who did not respond. For example, for the
survey question on "Elections that were less free and fair" in 2018, a
total of eight coders responded. However, we are unable to determine
(given certain use restrictions around the data) whether the final score
`0.000` for "Areas where the ruling party is weak" results from a
particular combination of coders, say five that agreed "no" this is not
a relevant criterion, and another three did not respond ("NA") because
they did not have an opinion. For example, in the case of Indigenous,
the final score is `0.500` but all we know is that half of all coders
thought this was relevant and the other half thought it was not or had
no opinion.

## Confidence scores

A high number of coders may shy away from even providing responses to
the four subnational democracy questions in the V-Dem surveys. There are
21 questions in the elections survey, of which only 2 are subnational,
and 24 questions in the civil liberties survey, of which only 2 are
subnational. Figure 1 shows the difference in the average number of
coders (V-Dem variables ending in `*_nr`) who responded to each category
of questions for the topics of elections and civil liberties in Colombia
from 1900-2022. Subtracting the average number of coders that responded
to subnational questions from the average number of coders that
responded to non-subnational questions, the differences have always
remained positive. Fewer coders respond to subnational questions than to
non-subnational questions. Because fewer coders decide to respond to
these questions, this may suggest a lack of conclusive information or
familiarity around the topic of subnational democracy in Colombia, at
least as conceptualized using V-Dem criteria. Comparing across surveys,
the elections questions register a higher average difference compared to
civil liberties questions, escalating notably after the direct elections
of mayors was introduced in 1988. There is a major escalation in
elections-related differences to about 8.5 coders between 2005-2012,
while the difference in civil liberties-related questions has declined
steadily over time. This jump may have been a result of V-Dem bridge
coders incorporated in a recent version of the survey.

![Difference between the average number of coders that responded to the
national-level survey questions and the average that responded to
subnational-level questions, in the case of Colombia.](figs/diff_sn.png)

These differences in the overall number of coders and the number of
coders who responded to subnational questions provide a relevant sense
of the "attrition" rate of coders. This information can be useful for
assessing the relative reliability of the proportional scores for a
given year. For example, in Colombia for 2018, the number of respondents
for each subnational survey question and maximum number for all other
survey questions in the survey were as follows:

-   Elections less free and fair: 8 over 13 maximum coders

-   Elections more free and fair: 9 over 13 maximum coders

-   Civil liberties stronger: 9 over 10 maximum coders

-   Civil liberties weaker: 10 over 10 maximum coders

These attrition rates suggest that civil liberties data have more
inter-coder reliability than election data. When operationalizing the
criteria from each survey, these coder attrition rates can serve to
temper the interpretation of composite scores for subnational democracy.
In this case the relative reliability of proportional scores for
"Elections less free and fair" will be lower than those for "Civil
liberties weaker." This means that resulting estimates for the civil
liberties dimension of subnational democracy in Colombian municipalities
in 2018 will be of higher confidence than those for elections.

Additional analysis of coder-level behavior may inform the weighting
procedure further. For example, bridge coders recruited in 2000s may
explain the escalation in differences between coders in national versus
subnational-level questions. Bridge coders typically code more than one
country, but may be hesitant to respond to subnational-level questions,
particularly if they reside outside of the country under analysis.
Controlling for bridge coders would lead to different weighted estimates
of subnational democracy. Another exercise could include comparing
individual coder behavior by year: some individuals may be more prone to
attrition during a certain period, such as 2005-2012, compared to those
just in that period.

# Conclusion

As noted previously, validation is a crucial step in producing data
based on expert ratings. We have not yet been able to do a systematic
validation of the data we have so far, but we have some observations
about face validity based on the maps we have produced. In general, our
data indicate higher levels of democracy (both in free and fair
elections and in protections for civil liberties) in the major urban
areas and lower levels in rural or remote areas such as the Amazon. [To
be fleshed out after we have a summary map.]

Systematic validation will require comparisons with independent
assessments of subnational democracy in Colombia. An initial source for
such information will be responses to open-ended versions of V-Dem's
four questions provided by a small number of university and master's
students at a Colombian university in February 2024. They were asked to
identify the most salient locations (defined any way they chose) in
Colombia and years where they would expect valid indicators to show
better or worse situations. We can check our data against their
expectations. At some point we could solicit the same kinds of responses
from a more senior panel of Colombian experts. Ultimately, the best kind
of validation would be administering an online survey asking V-Dem's
country experts on Colombia to code these same questions for Colombian
departments or municipalities for selected years. We do not know whether
such an effort is feasible, but it is probably impossible to know until
we try.

Eventually we hope to extend this exercise to other countries where
others have generated measures of subnational democracy, such as
Argentina, Mexico, India, the United States, and Russia. We do not
downplay the difficulty of such an effort. It requires access to very
specific geo-coded data on multiple demographic and political
indicators, which are probably not available for many countries.
Furthermore, producing cross-nationally comparable data involves an
additional layer of complication. The data we have produced for Colombia
is relative: the degree to which local situations are better or worse
than the national norm. Converting such relative data to absolute levels
will require both anchoring the ratings to a national score and
establishing the range of variation within which the relative scores can
vary. V-Dem's Clean Elections Index could serve as the anchor for the
subnational elections variables and its Civil Liberties Index could
serve as the anchor for the subnational civil liberties variables. The
same anchoring indices could be used for other countries. However, we do
not yet have a proposed solution for setting the range of variation
around these anchors. This would require careful conceptual and
methodological effort.

Colombia is among the best cases for this initial proof-of-concept
research due to the well-developed geographic data that DANE provides
(and for which the Lucy Institute, where co-author Matthew Sisk works,
has direct access). Replication in other countries would make it
possible to reach more general conclusions about the validity of the
technique and to decide whether it is promising enough to extend to many
more countries. If results are encouraging, there may be a potential for
extending of this technique to measure other political concepts
subnationally.

# References
