---
title: "Measuring Democracy in Subnational Units Worldwide: A New Method"
subtitle: "Validation Test 2 – Colombian Student Survey (Arboleda)"
author: "Patrick McQuestion"
date: "2026-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load data
snvdem <- readRDS("G:/Shared drives/snvdem/snvdem-col/data/panel/09_analysis_scripts/Exploratory/01_Expand/snvdem2.rds")
Arb    <- readRDS("G:/Shared drives/snvdem/snvdem-col/data/survey/Arb_final.rds")
```

---

## Overview

This section validates the SNDEM index against assessments from nine Colombian undergraduate and master's students from Universidad Sergio Arboleda. Respondents were asked to identify municipality-years where elections were more/less free and fair and where civil liberties were stronger/weaker.

We test whether higher index scores correspond to municipalities positively assessed by respondents, using a fixed-effects Linear Probability Model (LPM) with respondent fixed effects to control for individual bias.

---

## 1. Data Preparation

```{r data-prep}
library(dplyr)
library(zoo)
library(stringr)

# --- SNDEM: Normalize and calculate 3-year rolling averages ---
sndem_rolling <- snvdem %>%
  mutate(
    MPIO_CDPMP = str_pad(as.character(as.numeric(MPIO_CDPMP)), 5, pad = "0"),
    year = as.numeric(year)
  ) %>%
  mutate(
    sndem_norm    = (sndem    - min(sndem,    na.rm = TRUE)) / (max(sndem,    na.rm = TRUE) - min(sndem,    na.rm = TRUE)),
    snelect_norm  = (snelect  - min(snelect,  na.rm = TRUE)) / (max(snelect,  na.rm = TRUE) - min(snelect,  na.rm = TRUE)),
    sncivlib_norm = (sncivlib - min(sncivlib, na.rm = TRUE)) / (max(sncivlib, na.rm = TRUE) - min(sncivlib, na.rm = TRUE))
  ) %>%
  group_by(MPIO_CDPMP) %>%
  arrange(year) %>%
  mutate(
    sndem_roll_norm    = rollapply(sndem_norm,    width = 3, FUN = mean, align = "right", fill = NA, partial = TRUE),
    snelect_roll_norm  = rollapply(snelect_norm,  width = 3, FUN = mean, align = "right", fill = NA, partial = TRUE),
    sncivlib_roll_norm = rollapply(sncivlib_norm, width = 3, FUN = mean, align = "right", fill = NA, partial = TRUE)
  ) %>%
  ungroup()

# --- Arboleda: Clean and binarize quality labels ---
# Q_id 2 & 3 = high quality (more free elections / stronger civil liberties)
# Q_id 1 & 4 = low quality (less free / weaker)
arb_nominations <- Arb %>%
  filter(!is.na(Q_id), !is.na(MPIO_CDPMP), year >= 2000 & year <= 2023) %>%
  mutate(
    MPIO_CDPMP   = str_pad(as.character(as.numeric(MPIO_CDPMP)), 5, pad = "0"),
    year         = as.numeric(year),
    is_high_quality = ifelse(Q_id %in% c(2, 3), 1, 0)
  )

# --- Join: Match nominations to SNDEM scores ---
validation_set <- arb_nominations %>%
  inner_join(sndem_rolling, by = c("MPIO_CDPMP", "year"))

# Diagnostics
cat("Nominations in Arb:", nrow(arb_nominations), "\n")
cat("Matched observations in validation_set:", nrow(validation_set), "\n")
```

---

## 2. Linear Probability Models

```{r lpm-models}
library(fixest)

# NOTE ON STANDARD ERRORS:
# Cluster-robust SEs are theoretically appropriate here (FE by respondent),
# but with only N=9 clusters the asymptotic assumptions break down and SEs
# become unreliable. HC1 heteroskedasticity-robust SEs are used instead.
# See: Cameron & Miller (2015) on few-cluster problems.

# Model 1: Overall Index (3-Year Rolling Average)
# Tests whether sustained democratic performance predicts respondent opinion
mod_overall_roll <- feols(is_high_quality ~ sndem_roll_norm | Arb_id,
                          data = validation_set, vcov = "HC1")

# Model 2: Overall Index (Annual Score)
# Main robustness check – strongest result
mod_overall_annual <- feols(is_high_quality ~ sndem_norm | Arb_id,
                            data = validation_set, vcov = "HC1")

# Model 3: Components (Annual Normalized)
# Tests which pillar (elections vs. liberties) respondents are sensitive to
mod_comp_annual <- feols(is_high_quality ~ snelect_norm + sncivlib_norm | Arb_id,
                         data = validation_set, vcov = "HC1")

# Model 4: Components (Raw, Asymmetric Scale)
# Robustness check without 0-1 normalization
mod_comp_raw <- feols(is_high_quality ~ snelect + sncivlib | Arb_id,
                      data = validation_set, vcov = "HC1")

# Quick console output — signif codes match modelsummary stars below
etable(mod_overall_roll, mod_overall_annual, mod_comp_annual, mod_comp_raw,
       signif.code = c("***" = 0.01, "**" = 0.05, "*" = 0.1))
```

---

## 3. Variance Inflation Factor (VIF) – Multicollinearity Check

VIF is only meaningful for models with **two or more predictors**, so it applies to **Models 3 and 4** only. Models 1 and 2 each have a single predictor and are not affected by multicollinearity.

Because `car::vif()` does not support `fixest` objects directly, we replicate the fixed-effects estimator by **demeaning all variables within each respondent group** (`Arb_id`) before running a standard `lm()`. This is equivalent to what `feols()` does internally with the `| Arb_id` fixed effect.

```{r vif}
library(car)

# --- Within-respondent demeaning (replicates the FE transformation) ---
validation_demeaned <- validation_set %>%
  group_by(Arb_id) %>%
  mutate(
    snelect_norm_dm  = snelect_norm  - mean(snelect_norm,  na.rm = TRUE),
    sncivlib_norm_dm = sncivlib_norm - mean(sncivlib_norm, na.rm = TRUE),
    snelect_dm       = snelect       - mean(snelect,       na.rm = TRUE),
    sncivlib_dm      = sncivlib      - mean(sncivlib,      na.rm = TRUE),
    is_high_quality_dm = is_high_quality - mean(is_high_quality, na.rm = TRUE)
  ) %>%
  ungroup()

# --- Model 3 VIF: Normalized Components ---
lm_comp_annual_dm <- lm(is_high_quality_dm ~ snelect_norm_dm + sncivlib_norm_dm - 1,
                        data = validation_demeaned)
vif_m3 <- vif(lm_comp_annual_dm)

# --- Model 4 VIF: Raw Components ---
lm_comp_raw_dm <- lm(is_high_quality_dm ~ snelect_dm + sncivlib_dm - 1,
                     data = validation_demeaned)
vif_m4 <- vif(lm_comp_raw_dm)

# --- Print results ---
cat("=== VIF: Model 3 (Normalized Components) ===\n")
print(round(vif_m3, 3))

cat("\n=== VIF: Model 4 (Raw Components) ===\n")
print(round(vif_m4, 3))

# --- Correlation between the two predictors (for reference) ---
r_norm <- cor(validation_set$snelect_norm, validation_set$sncivlib_norm, use = "complete.obs")
r_raw  <- cor(validation_set$snelect,      validation_set$sncivlib,      use = "complete.obs")

cat("\nCorrelation between snelect_norm and sncivlib_norm:", round(r_norm, 3), "\n")
cat("Correlation between snelect and sncivlib:", round(r_raw, 3), "\n")
```

> **Interpretation guide:** VIF < 5 = low concern; VIF 5–10 = moderate; VIF > 10 = high multicollinearity. Given that `snelect` and `sncivlib` are the two components of `sndem` (their average), a high VIF here is expected and theoretically meaningful — it explains why the components are not individually significant in Models 3 and 4.

---

## 4. Results Table with VIF

```{r results-table}
library(modelsummary)
library(tibble)

# --- Build VIF row to append to table ---
# Models 1 & 2: N/A (single predictor); Models 3 & 4: report the higher of the two VIF values
vif_row <- tibble(
  term   = "VIF (max)",
  `Overall (Rolling)`   = "—",
  `Overall (Annual)`    = "—",
  `Components (Annual)` = as.character(round(max(vif_m3), 2)),
  `Components (Raw)`    = as.character(round(max(vif_m4), 2))
)

# --- Export table ---
modelsummary(
  list(
    "Overall (Rolling)"   = mod_overall_roll,
    "Overall (Annual)"    = mod_overall_annual,
    "Components (Annual)" = mod_comp_annual,
    "Components (Raw)"    = mod_comp_raw
  ),
  output = "G:/Shared drives/snvdem/snvdem-col/data/panel/09_analysis_scripts/Validation/ArboledaV/SNDEM_Validation_Results_VIF.docx",
  # Match significance thresholds to etable output above:
  # *** p < 0.01, ** p < 0.05, * p < 0.1
  stars  = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
  fmt    = 3,
  title  = "Table 3. Linear Probability Model results regressing index scores on student responses",
  coef_map = c(
    "sndem_roll_norm"  = "Index (3-Year Rolling Average)",
    "sndem_norm"       = "Index (Annual Score)",
    "snelect_norm"     = "Electoral Component (Annual)",
    "sncivlib_norm"    = "Civil Liberties Component (Annual)",
    "snelect"          = "Electoral Component (Raw)",
    "sncivlib"         = "Civil Liberties Component (Raw)"
  ),
  add_rows  = vif_row,
  gof_omit  = "AIC|BIC|Log.Lik",
  notes = c(
    "All models are Linear Probability Models (LPM) with Respondent (Arb_id) Fixed Effects.",
    "Heteroskedasticity-robust standard errors (HC1) used in place of cluster-robust SEs",
    "due to small number of respondent clusters (N = 9); see Cameron & Miller (2015).",
    "VIF calculated on within-respondent demeaned variables (replicating the FE transformation).",
    "VIF is not applicable for single-predictor models (Models 1-2).",
    "+ p < 0.1, * p < 0.05, ** p < 0.01, *** p < 0.001"
  )
)

cat("Table exported successfully.\n")
```

---

## 5. Interpretation Notes

```{r interpretation, echo=FALSE}
cat(
  "Models 1-2 (Overall Index):\n",
  "  - A unit increase in the normalized index is associated with a ~98% increase\n",
  "    in the probability that a respondent identifies a municipality as high quality.\n",
  "  - No multicollinearity concern (single predictor).\n\n",
  "Models 3-4 (Components):\n",
  "  - Neither snelect nor sncivlib is individually significant.\n",
  "  - High VIF (expected, given that sndem = mean(snelect, sncivlib)) confirms\n",
  "    that the two components are too correlated to distinguish individually.\n",
  "  - This aligns with the finding that respondents cannot easily separate\n",
  "    elections from civil liberties in their assessments.\n",
  sep = ""
)
```
