---
title: "Validation with Existing Data: Comparison with Rojas-Betancur's CdDS Index"
author: "McQuestion, Coppedge, Sisk, and McMann"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
    fig_caption: true
    keep_tex: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

# Adjust these paths to match your local environment
DATA_DIR  <- "G:/Shared drives/snvdem/snvdem-col/data/panel/"
GEO_DIR   <- "G:/Shared drives/snvdem/snvdem-col/data/geospatial/2018pmq/BaseLayer/"
OUT_DIR   <- "G:/Shared drives/snvdem/snvdem-col/data/panel/09_analysis_scripts/Validation/UMedellin/"
```

```{r libraries}
library(readr)
library(dplyr)
library(stringi)
library(ggplot2)
library(tidyr)
library(reshape2)
library(knitr)
library(kableExtra)
```

```{r load-data, cache=TRUE}
UMedM  <- read_csv(paste0(OUT_DIR, "Datos_municipal.csv"))
snvdem <- readRDS(paste0(DATA_DIR, "08_final_snvdem_data/MC/SNDEM_tentative.rds"))
MunYrs <- readRDS(paste0(DATA_DIR, "01_raw_data/MunYrs.rds"))
```

```{r clean-and-join, cache=TRUE}
# Cleaning function: removes accents, non-alphanumeric chars, and trims whitespace
clean_col_text <- function(x) {
  x <- as.character(x)
  x <- stri_trans_general(x, "Latin-ASCII")
  x <- gsub("[^[:alnum:][:space:]]", "", x)
  tolower(trimws(x))
}

# Prepare the bridge dataframe
bridge_clean <- MunYrs %>%
  mutate(
    match_mpio = clean_col_text(MPIO_CNMBR),
    match_dpto = clean_col_text(departamento),
    year = as.numeric(year)
  ) %>%
  select(MPIO_CDPMP, match_mpio, match_dpto, year) %>%
  distinct()

# Prepare Rojas-Betancur dataset
UMedM_clean <- UMedM %>%
  rename(CdDS = `Valor CdDS`) %>%
  mutate(
    match_mpio = clean_col_text(Municipio),
    match_dpto = clean_col_text(Departamento),
    Year = as.numeric(Year)
  )

# Match DANE codes to UMedM via bridge
UMedM_with_codes <- UMedM_clean %>%
  left_join(bridge_clean,
            by = c("match_mpio" = "match_mpio",
                   "match_dpto" = "match_dpto",
                   "Year"       = "year"))

# Final join with snvdem
validation_df <- snvdem %>%
  mutate(year = as.numeric(year),
         MPIO_CDPMP = as.character(MPIO_CDPMP)) %>%
  filter(year >= 2016) %>%
  inner_join(UMedM_with_codes,
             by = c("MPIO_CDPMP", "year" = "Year"))

# Standardize both indices for comparisons
validation_df <- validation_df %>%
  mutate(
    sndem_z = (sndem - mean(sndem, na.rm = TRUE)) / sd(sndem, na.rm = TRUE),
    CdDS_z  = (CdDS  - mean(CdDS,  na.rm = TRUE)) / sd(CdDS,  na.rm = TRUE)
  )
```

Our fourth validation test compares our estimates against the most comprehensive 
alternative measure of municipal-level democracy in Colombia currently available. 
We use the *Calidad de la Democracia Subnacional* (CdDS) index, developed by 
Milton Rojas-Betancur at the Universidad de Medellín's Observatory on Subnational 
Democracy. Rojas-Betancur (2021) draws on Leonardo Morlino's (2012) "Theory of Democratization" (or, TODEM) framework to construct 
an index of seven dimensions: rule of law, participation, electoral competition, 
electoral accountability, institutional accountability, civil liberties, and solidarism 
(education and health coverage). Rojas Betancur groups these into two factors--procedural 
and substantive--selects one to three indicators for each one, and standardizes each a 0-100 scale using quantitative data from 
Colombian official sources. His dataset covers all Colombian municipalities from 
2016 to 2023, overlapping temporally with our own index over eight years and 
across roughly 975 municipalities per year.

The two indices share important design features. Both seek to capture more than 
narrow electoral competition, both draw on administrative and observational 
data from Colombian government and third-party sources, and both cover all or nearly all 
municipalities rather than a selected subset. At the same time, they differ 
substantially in conceptualization, measurement strategy, and weighting procedure. 
Our index uses a liberal democracy framework anchored in V-Dem, emphasizing free 
and fair elections and civil liberties protection as assessed by country experts; 
the CdDS adopts a "democratic quality" framework that also encompasses institutional 
accountability and social rights (solidarism), with no expert or subjective components. 
Crucially, our index derives its weights from expert judgments about which local 
characteristics predict democratic variation, while Rojas Betancur assigns equal 
weight to each dimension. These differences lead us to expect a positive but 
moderate correlation: the two indices should share a common underlying signal about 
democratic conditions across Colombian municipalities, but they are not measuring 
the same thing.

The results confirm this expectation. Table 4 reports Pearson and Spearman 
correlations between our overall *sndem* index and the CdDS composite score for 
each year from 2016 to 2023.

```{r table4}
yearly_stats <- validation_df %>%
  group_by(year) %>%
  summarize(
    `Pearson r`        = round(cor(sndem, CdDS, use = "complete.obs"), 3),
    `Spearman r`       = round(cor(sndem, CdDS, method = "spearman", use = "complete.obs"), 3),
    `N municipalities` = n(),
    .groups = "drop"
  ) %>%
  rename(Year = year)

kable(
  yearly_stats,
  booktabs = TRUE,
  align    = c("l", "c", "c", "c"),
  caption  = "Annual correlations between \\textit{sndem} and Rojas-Betancur's CdDS index (Colombian municipalities, 2016-2023)"
) %>%
  kable_styling(latex_options = c("striped", "hold_position"), font_size = 10) %>%
  footnote(
    general = "All correlations are statistically significant at $p < 0.001$.",
    escape  = FALSE
  )
```

Correlations are consistently positive and statistically significant, ranging from 
0.22 to 0.44 (Pearson) and 0.25 to 0.46 (Spearman). The indices agree most 
strongly in the early years of overlap (2016--2018) and show a notable dip in 
2019--2021, recovering thereafter. This pattern is plausible: the 
dip may correspond to the period immediately following the 2016 FARC-EP demobilization, during which our 
index--which incorporates conflict and civil unrest variables weighted by expert 
judgment--may have responded differently to the uneven security 
improvements than did Rojas-Betancur's purely quantitative indicators.

Figure 8 plots the standardized Z-scores of both indices against each other for each 
year, confirming a consistent positive relationship across the full period.

```{r fig8, fig.cap="Yearly validation: \\textit{sndem} vs.\\ Rojas-Betancur's CdDS index (standardized Z-scores, 2016--2023)", fig.width=8, fig.height=5}
ggplot(validation_df, aes(x = CdDS_z, y = sndem_z)) +
  geom_point(alpha = 0.1, color = "steelblue", size = 0.8) +
  geom_smooth(method = "lm", color = "firebrick", se = TRUE) +
  facet_wrap(~year, ncol = 4) +
  labs(
    x = "Rojas-Betancur CdDS (Z-score)",
    y = "sndem (Z-score)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    strip.background = element_rect(fill = "gray95"),
    strip.text       = element_text(face = "bold")
  )
```

Figure 9 presents the cross-correlations between our component measures and each 
of Rojas-Betancur's seven dimensions. The component-level correlations are 
revealing. Both our electoral fairness (*snelect*) and civil liberties (*sncivlib*) 
components correlate most strongly with institutional accountability 
(A\_Inst, $r \approx 0.49$--$0.51$) and solidarism (Solid, $r \approx 0.46$--$0.50$). 
This is not surprising: institutional accountability in the CdDS is measured largely 
by DNP's Integral Performance Index--the same fiscal performance measure we use 
as a geopredictor for economic development--and solidarism captures education 
and health coverage, which are correlated with the socioeconomic conditions our 
index uses as predictors of democratic quality. Both indices therefore pick up a 
shared socioeconomic signal. Correlations with Rojas-Betancur's rule of law, 
participation, and electoral competition dimensions are near zero, reflecting genuine 
measurement differences: his participation measure is voter turnout and his 
competition measure is party alternation, neither of which enters our measurement 
approach directly.

```{r fig9, fig.cap="Cross-correlations between \\textit{sndem} components and Rojas-Betancur's CdDS dimensions (pooled, 2016--2023). Columns abbreviated: Law = rule of law; Part = participation; Comp = electoral competition; A.Elec = electoral accountability; A.Inst = institutional accountability; CivLib = civil liberties; Solid = solidarism.", fig.width=7, fig.height=3}
comp_data <- validation_df %>%
  select(
    sndem, snelect, sncivlib,
    Law    = `Estado de Derecho`,
    Part   = `Participacion`,
    Comp   = `Competencia Electoral`,
    A.Elec = `A-Electoral`,
    A.Inst = `A-Institucional`,
    CivLib = `Libertades civiles`,
    Solid  = `Solidaridad igualdad`,
    CdDS
  )

cor_matrix_full <- cor(comp_data, use = "pairwise.complete.obs")

my_vars     <- intersect(c("sndem", "snelect", "sncivlib"), rownames(cor_matrix_full))
author_vars <- intersect(c("Law","Part","Comp","A.Elec","A.Inst","CivLib","Solid","CdDS"),
                         colnames(cor_matrix_full))

cross_cor <- cor_matrix_full[my_vars, author_vars]

melted_cor <- melt(cross_cor)
colnames(melted_cor) <- c("snvdem_Var", "UMedM_Var", "Correlation")

ggplot(melted_cor, aes(x = UMedM_Var, y = snvdem_Var, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low      = "#0072B2",
    high     = "#D55E00",
    mid      = "white",
    midpoint = 0,
    limit    = c(-1, 1),
    name     = "Pearson r"
  ) +
  geom_text(aes(label = round(Correlation, 2)), size = 3.2) +
  theme_minimal(base_size = 10) +
  labs(x = "CdDS components", y = "sndem components") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The *negative* correlation between our civil liberties 
component and Rojas-Betancur's civil liberties dimension ($r \approx -0.28$ to $-0.30$) likely reflects a conceptual difference.
His civil liberties dimension takes the average of total homicides and the volume of *tutelas* (judicial petitions invoking 
constitutional rights) successfully resolved per municipality per year. More *tutelas* filed in a 
jurisdiction can reflect either more violations of rights *or* greater civic legal 
literacy and institutional engagement; more homicides, however, presumably restrict civil liberties. 
This more "negative" interpretation civil liberties contrasts with our civil liberties measure, anchored in 
V-Dem expert assessments of civil liberties understood more broadly as the degree to which the state protects rights, 
independent of whether citizens know to petition for them. The negative sign 
therefore reflects a real conceptual difference rather than measurement error in 
either index.

Taken together, these results support the validity of our method. The moderate 
positive correlations between the two overall indices --- despite their different 
conceptual frameworks, weighting schemes, and data sources --- indicate that both 
capture meaningful and partially overlapping variation in democratic conditions 
across Colombian municipalities. The component-level patterns clarify where the 
two approaches converge (on institutional performance and socioeconomic conditions) 
and where they diverge (on electoral processes and civil liberties operationalization). 
Rather than contradicting our measure, these divergences illuminate what is 
distinctive about each approach, and they underscore the value of the thick, 
expert-weighted, and multi-dimensional conceptualization our method provides.
